### 2012-03-10 LateX/paper5.tex

\documentclass{IEEEtran}
\usepackage{color}
\usepackage{array}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{paralist}
\usepackage{cite}
 \usepackage{color}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{subfigure}
\usepackage{tikz}
\usepackage{pgf}
\usepackage{pgfplots}
\usepackage{url}
\DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}

\begin{document}

\title{Acceleration of Bilateral filtering algorithm for manycore and multicore architectures}

\author{\IEEEauthorblockN{Dinesh Agarwal, Sami Wilf, Abinashi Dhungel, and Sushil K. Prasad}\\
\IEEEauthorblockA{ Georgia State University\\
 Department of Computer Science\\
 Atlanta, Georgia 30303\\
 } }
\maketitle
\begin{abstract}
This work explores multicore and manycore acceleration for embarrassingly parallel compute-intensive Bilateral filtering kernel. For manycore architectures, we have created a pair-symmetric algorithm to avoid redundant calculations and for multicore architectures we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads. We present empirical data evidencing the performance gains we achieved over variety of hardware architectures including Nvidia GTX280, AMD Barcelona, AMD Shanghai, Intel Harpertown, AMD Phenom, Intel Core i7 quad core, and Intel Nehalem 32 core machines. The best speedup achieved was a 235.5x speedup by our CUDA-based implementation of our pair-symmetric algorithm running on Nvidia's GTX280. Our CPU multicore implementations resulted in a speedup of up to 38x using 16 cores of AMD Barcelona each with 4-stage vector pipeline, up to 23x using 8 cores for Intel Harpertown, and up to 26x using 32 cores of Intel Nehalem multicore machines compared to a compiler-optimized code.
\end{abstract}
% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
\section{Introduction}
\label{sec:intro} The recent advances in parallel computing and architecture design have provided a series of avenues for application performance improvements. However, unlike the days when applications automatically became faster with increased clock frequency, newer chips require programmers and researchers to revisit their algorithm designs to harness the power of underlying architectures. Moreover, there is no well defined list of guidelines that can be followed for each application, which makes it challenging to deploy existing applications, even fundamental ones, to a myriad of available multicore and manycore architectures.

As reported in \cite{landscapeofpc} one of the critical questions for parallel computing community is to identify the common kernels being used in various applications. One such kernel is the Bilateral filter, a non-iterative image processing kernel widely used to perform directionally dependent image smoothing. In this paper we propose a set of optimizations and a non-naive algorithm for Bilateral filtering kernel. Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to cut processing time in half. We propose architecture specific optimizations, such as exploiting the unique capabilities of special registers available in modern multicore architectures and the rearrangement of data access patterns as per the computations to exploit special purpose instructions. We also propose optimizations for Nvidia's CUDA, including utilization of CUDA's implicit synchronization capability and the maximization of single-instruction-multiple-thread efficiency.

Our work explores the Bilateral filtering kernel proposed by Tomasi and Manduchi \cite{Tomasi1998}, which is used to smooth an image without blurring its edges. Bilateral filtering is ubiquitous in various image processing applications in various contexts such as denoising\cite{zhang2008}, texture editing and relighting\cite{eisemann2004}, tone management\cite{Bae2006}, stylization\cite{DeCarlo2002}, optical-flow estimations\cite{Xiao2006}, and demosaicking\cite{Ramanath2003} etc. A faster approximation to the original algorithm \cite{Paris2009} extends the image to 3D space by treating the pixel intensity as the third dimension and using a linear shift-invariant convolution in 3D.

A separable implementation \cite{Pham2005} applies the filter kernel first to one of the image dimensions and the intermediate result is filtered again on other dimensions. This reduces the complexity from $O(N*(m^{d}))$ to $O(N*d)$ where $N$ is the total number of pixels in the image and $d$ is the dimensions of the image. Yang et.al.\cite{yang2009}, have reported a constant time algorithm by dividing the Bilateral filtering kernel into a constant number of time spatial kernels. However, we do not parallelize these methods since our interest is to develop a parallelization method that can be applied, in general, to any spatially invariant and inseparable filtering algorithm.

Our methods are effective over a range of multicore chips with varying number of cores. In order to assess the efficacy of our optimization techniques, our experimental testbed includes a comprehensive set of multicore and manycore architecture chips including the Nvidia GTX280, AMD quad-core 2376 (Shanghai), AMD quad-core 8350 (Barcelona), Intel quad-core Xeon E5410 (Harpertown), Intel eight-core Xeon X7560 (Nehalem-EX), Intel quad-core Core i7, and AMD Phenom II six-core architectures. Our code is available under GNU GPL license from \url{http://code.google.com/p/blfilter/}.

Rest of the paper is organized as following: Section~\ref{sec:stencil} defines the Bilateral filtering kernel and introduces a naive algorithm and our pair-symmetric algorithm for Bilateral filtering. Section~\ref{sec:GPUoptimizations} introduces the optimizations employed to devise the pair-symmetric algorithm for CUDA architectures. Section~\ref{sec:optimizations} introduces the optimizations for CPU architectures and establishes the rationale behind ignoring traditional memory related optimizations for Bilateral filtering kernel. Our hardware and software testbed for experiments is detailed in Section~\ref{sec:expt}, immediately followed by Section~\ref{sec:results} that reports the performance of Bilateral filtering kernel on both multicore and manycore architectures. Section~\ref{sec:conclusion} concludes this paper with remarks on general applicability of our optimizations.

\section{Bilateral Filtering Kernel}
\label{sec:stencil}

\begin{figure}[h!]
\centering 
\includegraphics[width=3.2in]{images/filter} 
\caption{Example of a filter at the starting position of the image}
\label{fig:filter} 
\end{figure}

Filtering, an essential application of image processing, is carried out by a relatively small window, known as the filter (``Filter'' or ``Window''), which iterates over the source image, performing the same execution at each step of the iteration. The elements that the Filters cover in a given iteration makeup the neighborhood (``Neighborhood'') and are neighbors (``Neighbors'' or ``Neighboring Pixels'') of the Filter's center element (``Center'' or ``Center Pixel'').

Figure~\ref{fig:filter} illustrates the starting position of the Filter as commonly seen in standard, image processing algorithms; the source image is on the left and the destination image on the right is used to store the result of the Filter's computation at the respective Center's position.

Bilateral filtering combines two different kernels---the spatial kernel and the photometric kernel. With the photometric kernel, when the Center of the Filter rests on a pixel that boarders or is a part of a pictorial edge, e.g., an outline of an object or person against a contrasting background, the Neighbors that make up the other (contrasting) side of the edge have an insignificant impact on the Center Pixel's corresponding output pixel, which is stored in the destination array storing the processed image. Furthermore, when the Center is not bordering, and is not a part of a pictorial edge, the Neighbors contribute equally and the noise is filtered out. Hence, pictorial edges maintain sharpness while noise in the pictorial regions are filtered, resulting in more appealing images. Mathematically, the Bilateral filter can be defined in terms of the spatial kernel and photometric kernel as follows\cite{Tomasi1998}:\\

Spatial kernel: 
\begin{equation}
h(x)=k_{d}^{-1}(x)\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}f(\xi)c(\xi,x)d\xi\label{spatial}
\end{equation}

\begin{equation}
k_{d}(x)=\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}c(\xi,x)d\xi
\end{equation}
 Photometric kernel: 
\begin{equation}
h(x)=k_{r}^{-1}(x)\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}f(\xi)s(f(\xi),f(x))d\xi\label{photometric}
\end{equation}

\begin{equation}
k_{r}(x)=\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}s(f(\xi),f(x))d\xi
\end{equation}

The combined kernel, which is the Bilateral filter, is the product
of spatial kernel and photometric kernel and is described as: 
\begin{equation}
h(x)=k^{-1}(x)\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}f(\xi)c(\xi,x)s(f(\xi),f(x))d\xi\label{combined}
\end{equation}

\begin{equation}
k^{-1}(x)=\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}c(\xi,x)s(f(\xi),f(x))d\xi
\end{equation}

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation for a Center Pixel is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images. As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration. A formal expression is given below:

\begin{equation}
\label{eq:depen}
filtered\: pixel=\frac{\substack{N\\
\sum\\
i=1
}
x_{i}\cdot c(\xi,x)\cdot s(f(\xi),f(x))d\xi}{\substack{N\\
\sum\\
i=1
}
c(\xi,x)\cdot s(f(\xi),f(x))d\xi}
\end{equation}

In addition to restating the Bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the Bilateral filter. Specifically, Equation~\ref{eq:depen} shows that before the filtered pixel can be calculated by dividing the numerator by the denominator, both the summation in the numerator and summation in the denominator must first be calculated. Consequently, the Bilateral filter has two data dependencies. As explained later in this paper, the pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.

Another observation taken into account is that the values of equation (\ref{spatial}) for the Neighborhood can be computed in advance as the values depend on a simple Gaussian probability distribution function. However, the values of equation (\ref{photometric}) and consequently of equation (\ref{combined}) cannot be known a priori as they depend on the similarity of signals between the Center Pixel and the Neighborhood. Moreover, as the size of the Window increases, which only happens by choice of the user as the size is user defined, the number of computations per pixel grows at a rate of $O(FilterWindowWidth^{2})$ where FilterWindowWidth is the diameter of the Window. By reason of the foregoing, the bulk of the time spent in processing an image using the Bilateral filter is in the arithmetic-intensive computations, because the number of writes to the destination image is constant as the rate of growth of Center-Neighbor computations with respect to the Window size increase.

\subsection{A Naive Approach to Bilateral filetering Processing}

\begin{figure}
\centering\includegraphics[scale=0.35]{images/the1}
\includegraphics[clip,scale=0.35]{images/the202}\includegraphics[scale=0.35]{images/the2}\includegraphics[scale=0.35]{images/the3}
\caption{Example of a naive Bilateral filter iterations}
\end{figure}

The naive stencil algorithm is the standard model that is predominantly used in Bilateral filtering kernel. Previous implementations\cite{Paris2009,Pham2005} of Bilateral filtering kernel improves the hardware dependent aspects of Bilateral filtering kernel and hence expend resources on achieving performance gains that can not match the effort taken. Here, we address a root problem of the standard naive algorithm by creating a new algorithm, and then we also develop hardware dependent improvements for greater speedup. Below is the pseudo-code for the standard naive stencil kernel.

\begin{algorithm}[!h]
\caption{\label{alg:Example-Algorithm-float}Naive Stencil Kernel Algorithm}
\begin{algorithmic}
\FORALL{element  $e_{i,j}$ in source image  s}
\STATE{$C_{i,j}$ = $e_{i,j}$}, /* C is the center pixel */
\FOR{$k=-filter\_hw \to filter\_hw$}
\FOR {$l  = -filter\_hw \to filter\_hw$}
\FORALL{data dependency $n$}
\STATE{$X_n$ += $f_n(s[C_{i,j}],s[N_{i+k,j+l}])$}
\STATE /* N is the neighboring pixel */
\ENDFOR
\ENDFOR
\ENDFOR
\ENDFOR
\STATE $output[C_{i,j}] = f_{output}(X)$
\end{algorithmic}
\end{algorithm}

\subsection{Pair-symmetric Bilateral kernel Processing}
The pair-symmetric algorithm reduces the amount of Center-Neighbor computations by half and thus can theoretically cut processing time in half at the cost of increased memory requirement which increases with the amount of dependencies. Its implementation is non-trivial due to potential race-conditions if not done carefully.

Our pair-symmetric Bilateral filtering algorithm is premised on the fact that each pixel assumes the role of both a Center Pixel and a Neighboring Pixel and that the codes are symmetric with respect to Center-Neighbor paired computations. For example, in the Bilateral filter, $c(\xi,x)$ returns the same value as $c(x,\xi)$ and $s(f(\xi),f(x))$ returns the same value as $s(f(x),f(\xi))$. It then follows that the naive Bilateral filtering algorithm dictated in Algorithm~\ref{alg:Example-Algorithm-float} above, although simple, performs twice as many calculations than necessary.

Pair-symmetric algorithm eliminates this redundancy that is inherent in the naive Bilateral filtering algorithm. Instead of reading from all Neighbors, pair-symmetric algorithm reads from only half of the Neighbors. The result of each Center-Neighbor paired computation is then added to the $X_n$ register, which eventually obtains the result of data dependency $n$. This is trivial as the naive Bilateral filtering algorithm already does this. What pair-symmetric algorithm does unique is that the result of each Center-Neighbor pair computation is also added to and stored in the Neighbor's position in a temporary array (``Map'') that is used for storing the results of a data dependency. So, there are $n$ Maps for $n$ number of dependencies.

After the data dependency conditions are met, the function that returns the output data from the dependencies iterates over the Maps. As applied to the Bilateral filter, the data dependency lies in calculating each pixel's numerator and denominator that is used to derive the pixel's output result. After the numerators and denominators are derived, a function iterates and divides the corresponding elements of the two arrays to return the output. The pseudocode is given below:

\begin{algorithm}[h]
\caption{\label{alg:Example-Algorithm-float-1}Pair-symmetric Bilateral Filtering Algorithm}
\begin{algorithmic}
\FORALL{element $e_{i,j}$ in source image $s$}
\STATE $C_{i,j}$ = $e_{i,j}$, /* C is the center pixel */
\STATE{$X_n$= 0}
\FOR{$k=0 \to filter\_hw$}
\FOR{$l = 0 \to filter\_hw$}
\FORALL{data dependency $n$}
\STATE $temp$ = $f_n$ ($s$[$C_{i,j}$],$s$[$N_{i+k,j+l}$])
\STATE /* N is the neighboring pixel */
\STATE $X_n$ += $temp$
\STATE $D_n$[$N_{k,l}$] += $temp$
\STATE /* D is the dependency array */
\FOR{$k =-1 \to -filter\_hw$}
\FOR{$l = -1 \to -filter\_hw$}
\STATE $temp$ = $f_n(s[C_{i,j}],s[N_{i+k,j+l}]$
\STATE $X_n$+= $temp$
\STATE $D_n$[$N_{k,l}$] += $temp$
\STATE $D_n$[$C_{i,j}$] += $X_n$
\FORALL{element $e_{i,j}$ of output image}
\STATE{output[$e_{i,j}$]=$f_{output}(D)$}
\ENDFOR
\ENDFOR
\ENDFOR
\ENDFOR
\ENDFOR
\ENDFOR
\ENDFOR
\end{algorithmic}
\end{algorithm}

\section{CUDA-based Implementation Details}
\label{sec:GPUoptimizations} 
As Bilateral filtering algorithm belongs to the class of embarrassingly parallel algorithms, CUDA is well suited for its implementation. However,
in order to effectively implement such an algorithm in CUDA, one must understand how CUDA operates, which is significantly different from any CPU.

\subsection{Implicit Intra-Warp Synchronization and Memory Access Patterns}

CUDA's intra-warp implicit synchronization capability is the cornerstone of our CUDA implementation of pair-symmetric algorithm. Without implicit synchronization, pair-symmetric algorithm would suffer significantly over CUDA because it would require expensive synchronization calls per Center-Neighbor Filter computation.

In CUDA, threads are organized into groups called warps. Each warp executes one instruction at a time for its own group of threads. Because all threads in a warp share the same program counter and are therefore driven by the same sequential instruction stream, all threads in a warp are always in sync with each other without requiring explicit synchronization calls. Therefore, synchronization between threads in a warp is implicit.

To take full advantage of CUDA's implicit intra-warp synchronization feature, divergent branching between threads in the same warp must be avoided. This is so because all the threads in the same warp share only one program counter. So, a divergent branch will cause some or most of the threads to be idle as the program counter runs through the branch and drives only those threads that meet the branch condition to work. An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)\{...\} else\{...\})$.

As it relates to stencil codes in general, divergent branching will commonly occur in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts. For example, if eight threads were to read from eight non-contiguous segments of global memory--such as that which is commonly seen in strided access patterns like array{[}threadIdx.x {*} 2{]}--divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined.

In other words, a memory transaction will be serialized unless all the threads in the warp to which the transaction belongs are aligned consecutively and fit within a segment of memory. Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory. What this means in the implementation of Bilateral filtering algorithm is that warps should correspond to horizontal rows of pixels for an i{*}jMax+j representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp, and if, for example, there were two rows such that half of the threads were reading from memory on each row, then there would be two transactions instead of one which is inefficient.

While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable. As a consequence, work must be arranged such that warps can work as independent from each other as possible in order to minimize explicit synchronization calls as shown in Figure~\ref{fig:pspro}. This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the instant row of input data has completed. This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call will result in a warp trespassing into the territory of another warp, thereby causing race conditions as threads from two different warps will consequently be prone to write to the same memory location.

\subsection{Shared Memory and Tile Processing}

\begin{figure}
\includegraphics[width = 1 \columnwidth]{images/project22}
\caption{Illustration of CUDA implementation of pair-symmetric processing}
\label{fig:pspro}
\end{figure}

CUDA features three types of memory with read/write capability suitable for general-purpose computing on GPU: \begin{inparaenum}[(i)] \item global memory, which is accessible by all threads in all thread blocks, \item shared memory, which is provided separately for each block of threads to share among themselves, and \item registers, which are separate and distinct for each thread and of which each is only accessible by the thread to which it belongs\end{inparaenum}. Of these three types of memory, the slowest is global memory, which unfortunately is where the host must first send the input data before the GPU can begin its work. So, in order to see any meaningful performance in CUDA, data stored in global memory must be transferred to faster memory--i.e., shared memory, which is hundreds of times faster. However, the trade-off is that the shared memory size is much smaller than the global memory size, and this fact compels a tile-based approach to the implementation of Bilateral filtering algorithm.

Our tile-based processing approach is self-explanatory. Equal sized partitions of the input data are processed, and then any remaining data on and between the edge and the last partition that is closest to (but not falling over) the edge boundary of the 2-dimensional data representation is processed. With respect to the naive Bilateral filtering algorithm, tile-based processing is trivial and needs no further explanation because simply partitioning the space and processing it will suffice. However, for the pair-symmetric algorithm that we propose, tile-based Bilateral filtering algorithm requires adjustments in the order in which tiles are processed in order to guarantee the exclusion of race conditions.

\begin{figure}
\includegraphics[width=0.48\textwidth]{images/stackedtiles}
\caption{Tiling for Pair-Symmetric Stencil Processing with each tile having a unique color and overlapping tiles have a resultant color that results from mixing colors of overlapping tiles. The pair-symmetric thread access pattern ensures exclusive memory writes.}
\label{fig:Tiling-for-Pair-Symmetric}
\end{figure}

\begin{figure}
\includegraphics[width=0.48\textwidth]{images/lenatiles}
\caption{Input image after one iteration of pair-symmetric algorithm. Tiles do not have perfect boundaries due to overlaps.}
\label{fig:An-iteration-of}
\end{figure}

If processes were assigned to adjacent tiles, race conditions would occur because the edge cells of the tiles, a by-product of the Filter's half-width, would overlap each other, as illustrated in Figure~\ref{fig:Tiling-for-Pair-Symmetric}. To get around this problem, simultaneously processed tiles can be spaced apart in strides such that no two tiles being processed at the same time intersect with one another. An example of such a technique is illustrated in Figure~\ref{fig:An-iteration-of} (the tiles are flipped sideways in this rendering).

\section{CPU-based Implementation Details}
\label{sec:optimizations}

\subsection{Memory to computation ratio of Bilateral filtering kernel}
\begin{figure}[h!]
\begin{centering}
\subfigure{ \includegraphics[width=0.48\textwidth]{images/noexplarge}} 
\par\end{centering}
\caption{Percentage of time taken by sequential kernel when run without $exp$ instruction, $filter$ $radius$=10.}
\label{fig:noexp} 
\end{figure}

Unlike traditional regular embarrassingly parallel stencil kernel codes where memory management plays a pivotal role in realizing performance, Bilateral filtering kernel is extremely computationally intensive with only a meager amount of time spent in memory accesses. Bilateral filtering kernel requires computation of $exponential$ function ($exp$ function) which is a major component of time spent in processing. Figure~\ref{fig:noexp} shows the percentage of time taken by Naive Bilateral filtering algorithm\cite{Tomasi1998} and an improved algorithm employing tiled-memory-access pattern on different platforms for a 512x512 pixels small image and a 3000x3000 pixels large image after the $exp$ function was commented out from the computation. The only purpose of this experiment is to justify the fact that traditional memory-access-pattern related optimizations can not improve the performance of Bilateral filtering kernel.

It can be seen that after taking the $exp$ instruction out of the computation, the total sequential execution time is consistently less than 2\% of the original execution time with $exp$ instruction in place. For the smaller image it is strictly less than 1\% of the original time with $exp$ instruction in place. To be precise it is 1.39\% maximum on Intel Nehalem for the larger image and 0.65\% for the smaller image on Core i7 processor. The lower memory transfer time for smaller image is due to the fact that the smaller image fits into the cache memory entirely and hence there are no repeated data transfers between the processor and the main memory.

Since the $exp$ instruction does not involve transfer of data from main memory, it follows from this experiment that memory reuse and memory access pattern related optimizations are not of much help to speed up Bilateral filtering kernel. What Amdahl's law claims, for inherently sequential portion of a parallel algorithm, can be applied here. \\ \textit{The total speedup achieved by optimizations that deal with memory management is limited by the amount of time spent in memory transfers}. Therefore, in this case, traditional optimizations such as tiling/blocking, cache efficient initializations, prefetching, and other NUMA optimizations can only result in a speed up of at most 2\% and hence we have concentrated on optimizations that deal with the complexities of computational aspects of Bilateral filtering kernel.

\subsection{SIMD optimizations}
\label{sub:SIMD1} 
$Intel$'s Streaming SIMD Extension (SSE) is a technique to exploit micro-level parallelism in x86 based processors using SIMD capabilities. SSE can be ideal for tasks that involve repetitive computation such as the 3D heat equation or any other stencil kernels such as Jacobi iterations\cite{kdatta2008}. The matrix multiplication kernel was implemented in \cite{matrix-multi,MatrixMultiFloating} for matrices with integers and floating point numbers. SSE instructions have also been used for image or template matching algorithms such as normalized cross correlation (``NCC'') \cite{NCCmatching}.

Modern day CPUs use SSE that is an extension of MMX technology and is fundamentally similar to MMX. It has 8, 128-bit registers named XMM0 through XMM7. SSE instructions can operate in parallel on data packed in these registers thus providing a theoretically maximum speedup of 16 when working with 1 Byte data structures or 8 when working with 2 Bytes long data structures and so on. There are other extensions of SSE such as SSE2, SSE3, and SSE4.2. Moreover, recent extension of SSE, named Advanced Vector Extensions (``AVX''), uses 256-bit data registers and thus a theoretical peak speedup of 32 is possible.

Our CPU based implementation of Bilateral filtering algorithm explores the use of special purpose registers using both assembly level programming and SSE intrinsics. The advantage of using assembly language is that a programmer can have explicit control of underlying hardware. However, programmer productivity is affected and an in-depth knowledge of underlying hardware is essential.

Although SSE intrinsics are wrappers for assembly language, it is preferred to use SSE intrinsics due to ease of coding and also due to efficient utilization of available assembly level instructions without knowing the low level details of underlying architecture. A programmer, working with intrinsics, can specify the intrinsic and the appropriate instruction will automatically replace this intrinsic during compilation. Moreover, there are some operations for which there are no assembly instructions and hence in case of assembly level programming the code size can grow to such an extent that readability gets compromised.

Although the modern compilers usually try to improve the performance by automatically vectorizing, unrolling loops, prefetching, and SIMDization, we were still able to get significant performance improvements alongside compiler optimizations.

\subsection{Reduction Methods}

\label{sub:reduction} Sum by reduction is a technique to compute sum of multiple elements usually stored in contiguous memory locations. There are assembly instructions that can add horizontally packed, single precision, floating point numbers. We used $haddps$ assembly instruction that takes two (source and destination) 128 bit operands, and adds the single precision floating point numbers stored in first and second $dword$ of destination and source operand and stores it in first and third $dword$ of the destination operand respectively. Similarly, the third and fourth $dword$ of destination and source operand are added and stored in second and fourth $dword$ of the destination operand respectively. Both source and destination values must be stored in MMX registers. The SSE intrinsic to perform horizontal add is $\_mm\_hadd\_ps$. It takes two 128 bit operands, each of which is treated as four 32-bit floating point elements. The result of this operation on operand X(X0, X1, X2, X3) and Y(Y0, Y1, Y2, Y3) is (Y3 + Y2, Y1 + Y0, X3 + X2, X1 + X0).

\section{Experimental Testbed}
\label{sec:expt}

\begin{table*}[t]
\begin{tabular}{|l|p{1.2cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{3.5cm}|l|p{2cm}|l|p{0.9cm}|}
\hline 
\textbf{Model}  & \textbf{Clock frequency}  & \textbf{cores per chip}  & \textbf{chips per node}  & \textbf{cores per node}  & \textbf{SIMD support}  & \textbf{Cache/Memory}  & Compiler\\
\hline 
Intel Xeon E5410  & 2.33 $GH_z$  & 4  & 2  & 8  & SSE, SSE2  & 12 MB  & icpc 11.1\\
\hline 
Intel Xeon X7560  & 2.27 $GH_z$  & 32  & 1  & 32  & SSE, SSE2  & 24 MB & icpc 11.1\\
\hline 
Intel Core i7-870  & 2.93 $GH_z$  & 4  & 1  & 4  & SSE, SSE2, SSE3, SSE4.2  & 8MB & icpc 12.0.2\\
\hline 
AMD Opteron 2376  & 2.3 $GH_z$  & 4  & 2  & 8  & SSE, SSE2  & 6MB  & icpc 11.1\\
\hline 
AMD Opteron 8350  & 2.0 $GH_z$  & 4  & 4  & 16  & SSE, SSE2  & 2MB  & icpc 11.1 \\
\hline 
AMD Phenom II X6 1045T  & 2.7 $GH_z$  & 3  & 1  & 6  & SSE, SSE2, SSE4a  & 6MB  & icpc 12.0.2 \\
\hline 
NVidia GeForce GTX 280 & 1296 MHz & 240 & 1 & 240 & - & 1 GB GDDR3 & CUDA 2.0\\
\hline
\end{tabular}\caption{Architectural details of multicore chips employed for experiments}
\label{tab:archs} 
\end{table*}

In order to evaluate our implementation we have tried it on a range of multicore and manycore chips. Table \ref{tab:archs} summarizes architectural details of systems we used during our experiments.

\subsection{Hardware}
\label{subsec:hw} 
\subsubsection{Intel Xeon}
\label{ss:xeon} We have used two different Xeon chips, E5410 (Harpertown) and X7560 (Nehalem EX). Harpertown is a dual-socket, dual-processor relatively old system with a clock frequency of 2.33 $GH_z$.  Harpertown does not have Intel's hyperthreading (HT) technology which means the maximum number of threads is equal to the number of cores available. Intel Nehalem EX is considerably newer chip having eight-cores processing at a clock frequency of 2.27 $GH_z$. We had access to a Linux cluster from Intel's Manycore Testing Lab (MTL) containing four sockets of Nehalem EX chips thereby making 32 cores available to experiment with. The processors are connected to each other via Intel QPI (Quickpath interconnect) links that can transfer data at 6.4 GT/s. Both of these machines support 64 bit instructions.

\subsubsection{AMD Opteron}
\label{ss:opteron} AMD Opteron 2376 (Shanghai) chip has private L1 and L2 caches of size 128KB (64 D + 64 I) and 512KB respectively per core along with a 6MB cache shared among all of the cores. It has a clock frequency of 2.30 $GH_z$ and it is a quad-core architecture. Our compute node has two of Shanghai chips making it a total of eight cores to be used for experiments. AMD Opteron 8350 (Barcelona) is similar to Shanghai in many aspects but it has a clock frequency of 2.0 $GH_z$ and the shared cache is only 2MB. Our compute node contains 4 of Barcelona chips and thus we are able to run 16 threads on this node.

\subsubsection{Intel Core i7}

\label{ss:corei7}Intel Core i7 - 870 supports SSE 4.2 and has AVX technology. The clock frequency is an impressive 2.93 $GH_z$ and it can operate at a maximum of 3.6 $GH_z$ supported by $Intel$'s power boost technology. Its architecture supports hyperthreading which lets an application to execute 8 threads in total; however, we only used one thread per core. It contains 2 memory channels that provide a theoretical maximum memory bandwidth of 21 GB/s.

\subsubsection{AMD Phenom II}
\label{ss:phenom} AMD Phenom is a six core processor with each core capable of operating at 2.7 $GH_z$. If only three cores are in operation it can run at a maximum of 3.2 $GH_z$. It boasts 128 KB (64 D + 64 I) L1 cache, and 512 KB L2 cache per core in addition to a 6MB L3 cache shared among all six cores. The cache latencies are 3, 13, and 47 clock cycles for L1, L2, and L3 caches respectively. The cache line size is 64 bytes, L1 caches are 2 way set associative, L2 caches are 16-way set associative and L3 cache is 48-way set associative.

\subsubsection{NVidia GeForce GTX 280}

GTX 280 by NVidia is a GPU chip with 240 CUDA cores with 602 $MH_z$ graphics clock and 1296 $MH_z$ processor clock. GTX 280 is equipped with a 1 GB standard memory with 1107 $MH_z$ memory clock and 141.7 GB/s memory bandwidth. GTX 280 boasts a memory interface width of 512-bit with texture fill rate of 48.2 billion/second.

\subsection{Software Framework}

\label{subsec:sw} Our implementation for multicores is written in $C++$ and uses OpenMP for parallelization; the GPU implementation uses CUDA. To enrich supported image formats, we use $CImg$ library. 
\subsubsection{Algorithm variations on multicore architectures}
We have implemented a number of variations of Bilateral filtering kernel for multicore chips, results of which are reported in section \ref{sec:results}. Our GPU implementation uses naive algorithm and the pair symmetric algorithm. For multicore chips we have created multiple variations of naive algorithm. These algorithms are labeled as follows: 
\begin{itemize}
\item $BL\_Naive$ is the sequential version of the code, as described by Tomasi\cite{Tomasi1998}. 
\item $BL\_Tiles$ is an improved version of Bilateral filtering kernel that employs data blocking to leverage the availability of multi-level cache hierarchy in the modern multicore CPUs.   
\item $BL\_Assembly$ is the version of Bilateral filtering kernel implemented using assembly level code. 
\item $BL\_A\_Reduction$ is the version of Bilateral filtering that uses assembly language as well as reduction optimizations. 
\item $BL\_Intrinsic$ version uses SSE intrinsics instead of assembly language.
\item $BL\_I\_Reduction$ is the extended version of $BL\_Intrinsic$ that uses the intrinsics that can compute sum of two 128 bit operands in parallel by exploiting the vector array processing. 
\end{itemize}
$BL\_Naive$ has a redundant calculation inside the innermost loop which has been moved out of loop for all other versions. The original instruction involving this calculation looks like
 $$gaussian\_ph = exp\left( \frac{-pd} {2 * sigma\_ph*sigma\_ph}\right);$$
After moving the redundant calculation out of the loop the same instruction becomes
 $$gaussian\_ph = exp( pd * pre\_calc\_sigma\_sq );$$

Our Bilateral filtering code is open-source licensed under GPL and available from $Google~Code$ repository \url{http://code.google.com/p/blfilter} through SVN version management system. 
\subsubsection{Input Image}
In order to analyze the behavior of our kernel with varying image sizes, we have executed our kernel with two different images. One that fits into cache easily and the other that does not fit into cache but fits into main memory. Both of the images were converted to single spectrum images represented by single precision numbers for each pixel. The smaller image was 512 pixels wide and 512 pixels high, thus requiring 1MB memory to store it. As shown in Table~\ref{tab:archs} all of the multicore systems have cache memories bigger than 1 MB. The second image was 3000 pixels wide and 3000 pixels high and hence requires little more than 34 MB of memory to store it. None of the multicore cache memories were big enough to store this image entirely.

\section{Performance Results}
\label{sec:results}
\subsection{Performance of Bilateral filtering kernel on CPU}
For all of our experiments the filter radius was kept at 10, subsequently making the filter Window of size 21 x 21. The spatial spread and the photometric spread were also set to 10. The optimization flag $O3$ was used for all experiments in addition to SSE2 for SIMDization. For Core i7 chip compiler flag SSE4.2 was used instead of SSE2. We do not use hyperthreading, i.e., there are no more than one thread per core.

\begin{figure*}[t!] 
\begin{centering} 
\subfigure[Nehalem-EX Xeon X7650]{ \includegraphics[width=0.3\textwidth]{images/nehalem} \label{fig:Nehalem} } 
\subfigure[Harpertown Xeon E5410]{ \includegraphics[width=0.3\textwidth]{images/harpertown} \label{fig:harpertown} } 
\subfigure[Core i7 - 870]{ \includegraphics[width=0.3\textwidth]{images/corei7} \label{fig:corei7} } 
\end{centering}
\caption{Bilateral filtering kernel on Intel chips. The baseline version is BL\_Naive version, auto optimized by compiler with O3 and SSE flags.} 
\label{fig:comparisonIntel} 
\end{figure*}

\begin{figure*} 
\begin{centering} 
\subfigure[Barcelona AMD Opteron 8350]{ \includegraphics[width=0.3\textwidth]{images/barcelona} \label{fig:barcelona} } 
\subfigure[Shanghai AMD Opteron 2376]{ \includegraphics[width=0.3\textwidth]{images/shanghai} \label{fig:Shanghai} } 
\subfigure[AMD Phenom II 1045T]{ \includegraphics[width=0.3\textwidth]{images/phenom} \label{fig:Phenom} } 
\end{centering}
\caption{Bilateral filtering kernel on AMD chips. The comparison is with BL\_Naive version, auto optimized by compiler with O3 and SSE flags.} 
\label{fig:comparisonAMD} 
\end{figure*} 
\begin{figure*}[t!]

\begin{centering} 
\subfigure[Harpertown Xeon E5410]{  \includegraphics[width=0.32\textwidth]{images/harpertownfinal} \label{fig:harpertownFinal} } 
\subfigure[Core i7 - 870]{ \includegraphics[width=0.32\textwidth]{images/corei7final} \label{fig:corei7Final} } 
\subfigure[Nehalem-EX Xeon X7650]{ \includegraphics[width=0.32\textwidth]{images/nehalemfinal} \label{fig:NehalemFinal} } 
\end{centering}
\caption{Comparison of combined speedup due to SIMDization and OpenMP parallelization on Intel chips for different algorithms} 
\label{fig:comparisonFinalIntel} 
\end{figure*}

\begin{figure*} 
\begin{centering} 
\subfigure[Barcelona AMD Opteron 8350]{ \includegraphics[width=0.3\textwidth]{images/barcelonafinal} \label{fig:barcelonaFinal} } 
\subfigure[Shanghai AMD Opteron 2376]{ \includegraphics[width=0.3\textwidth]{images/shanghaifinal} \label{fig:ShanghaiFinal} } 
\subfigure[AMD Phenom II 1045T]{ \includegraphics[width=0.3\textwidth]{images/phenomfinal} \label{fig:PhenomFinal} } 
\end{centering}
\caption{Comparison of combined speedup due to SIMDization and OpenMP parallelization on AMD chips for different algorithms.} 
\label{fig:comparisonFinalAMD} 
\end{figure*}

Figure~\ref{fig:comparisonIntel} compares different algorithms on all three $Intel$ chips. In this experiment we are comparing the absolute speedup achieved by the respective algorithm compared to BL\_Naive algorithm. It is important to note that these results are for micro level parallelism (use of 128 bit special purpose registers) only and do not consider speedup due to high level (OpenMP) parallelization. The parallel versions shown here are compared against the parallel version of BL\_Naive algorithm. The speedup for all algorithms is still bounded by the theoretical peak speedup which happens to be four here since we work with 32bit floating point numbers. Interestingly, although we use compiler optimization flags such as -O3 and -xSSE2, i.e., there is automated vectorization taking place in all the algorithms, our hand coded vectorization is superior in many cases.

The speedup shown by BL\_Tiles is mostly due to the computation reduction as discussed in section \ref{subsec:sw} because we do not use special purpose registers in this algorithm. It can be seen from figure~\ref{fig:comparisonIntel} that the parallelism due to SIMDization is limited by four which is the theoretical maximum possible speedup due to SSE registers as discussed in section \ref{sub:SIMD1}.

As seen in figure~\ref{fig:Nehalem}, the maximum speedup achieved on Nehalem architecture only using SSE optimization is 2.29x for the smaller image and 1.33x for the larger image. Smaller image generally performs better because the time to copy data to SSE registers is less as the entire image is always available in the cache memory. The BL\_Naive algorithm outperforms our assembly level implementations but it is slower than our intrinsic based implementations in all cases. The reason for this is that the auto SIMDization by compiler is more efficient than our implementation for Nehalem architecture. Intel Harpertown, which is a relatively older chip, shows an impressive speedup of 3.54x (Figure~\ref{fig:harpertown}) even though compiler optimizations are still in place.

Another interesting observation is the results on Intel Core i7 architecture shown in Figure~\ref{fig:corei7}, where we have the absolute speedup of less than $1$. This indicates the fact that the automated compiler optimizations on newer chips are superior than on older chips. These experiments were run with $SSE4.2$ compiler flag and hence the compiler is able to exploit SIMDization automatically thus making it difficult to surpass it using assembly level coding.

However, our algorithms that use SSE intrinsics and reduction, i.e. BL\_A\_Reduction, BL\_Intrinsic, and BL\_I\_Reduction are able to provide better speedups than those achieved by compiler optimizations as shown in Figure~\ref{fig:comparisonIntel} and Figure~\ref{fig:comparisonAMD}.

Figure~\ref{fig:comparisonAMD} shows the similar comparison across AMD chips. The results are in tune with those for Intel platform. As shown in Figure~\ref{fig:barcelona} for Barcelona, which is a relatively older chip, the maximum speedup is 3.24x for BL\_I\_Reduction algorithm. Similarly, for AMD Shanghai, as shown in Figure~\ref{fig:Shanghai}, the maximum speedup achieved by exploiting SIMDization is 3.02x for BL\_I\_Reduction algorithm. The case of AMD Phenom is shown in Figure~\ref{fig:Phenom}. The auto tuning applied by the compiler outperforms the manual tuning in case of BL\_Assembly and BL\_A\_Reduction. However, BL\_Intrinsic and BL\_I\_Reduction are still able to outperform compiler optimizations. The maximum speedup achieved in this case is 1.30x.

The next set of results demonstrate the total speedup due to OpenMP as well as due to SIMDization. It is interesting to see how speedup due to OpenMP multiplies with the speedup due to SSE optimizations to exhibit super linear speedup.

Figure~\ref{fig:harpertownFinal} shows the performance of Bilateral filtering kernel on Intel Harpertown. This machine has 8 cores and the maximum combined speedup achieved is 20.91x. The speedup achieved only using only SSE optimization was 2.69x and the speedup achieved by BL\_Naive algorithm using 8 cores is 7.73x.

Intel Nehalem chip with 32 cores does not show a super-linear speedup as shown in Figure~\ref{fig:NehalemFinal}. However, the reason behind this is the lack of enough parallelism that can be afforded by the kernel for this image. Both optimizations still multiply their individual speedup to result in the combined speedup. The total combined speedup is 26.46x using 32 cores, the speedup only due to SSE was 1.12x and the speedup for BL\_Naive is 23.47x.

Intel Core i7 chip (Figure~\ref{fig:corei7Final}) also demonstrates the similar behavior. The final speedup of 4.63x, using 4 cores, is a result of speedup due to SSE which is 1.33x and speedup due to OpenMP which is 3.48x.

Figure~\ref{fig:comparisonFinalAMD} shows a similar comparison over various AMD chips. Figure~\ref{fig:barcelonaFinal} shows the speedup for Barcelona chip. This chip has 16 cores with a combined maximum speedup being 38.03x. The speedup achieved using SSE optimizations was 2.48x and the speedup for BL\_Naive was 15.34x and thus these both realize a speedup of 38.03x using 16 cores.

AMD Shanghai similarly shows an impressive speedup of 20.92x (Figure~\ref{fig:ShanghaiFinal}). The interesting case is with AMD Phenom as shown in Figure~\ref{fig:PhenomFinal}. The speedup drops a little as number of cores increases beyond three. The reason behind this is the turbo core technology as discussed in section (\ref{ss:phenom}). When there are a maximum of three cores being used the clock frequency scales up to 3.6 $GH_z$ but as soon as more threads are created the clock frequency steps down to 2.7 $GH_z$.

\subsection{Performance of Bilateral filtering kernel on GPU}
We have performed thorough experiments on the GPU architecture with varying number of blocks and threads for both naive Bilateral filtering algorithm and pair-symmetric Bilateral filtering algorithm.

\begin{figure}[t!] 
\begin{centering} 
\subfigure[Large Image, Filter Width = $10$]{ \includegraphics[width=0.4\textwidth]{images/gpuhw10} \label{fig:hw10} } 
\subfigure[Large Image, Filter Width = $5$]{ \includegraphics[width=0.4\textwidth]{images/gpuhw5} \label{fig:hw5} } 
\subfigure[Smaller Image, Filter Width = $10$]{ \includegraphics[width=0.4\textwidth]{images/gpuhw10light} \label{fig:hw10light} } 
\subfigure[Smaller Image, Filter Width = $5$]{ \includegraphics[width=0.4\textwidth]{images/gpuhw5light} \label{fig:hw5light} } 
\end{centering}
\caption{Performance of Naive and pair-symmetric (P.S.) Bilateral filtering kernel algorithms on GTX 280 NVidia card for both smaller image and larger image.} 
\label{fig:gpuresults} 
\end{figure}

Figure~\ref{fig:gpuresults} represents the performance of both of the algorithms.

\section{Conclusion and future work}
\label{sec:conclusion} 
Modern day scientific applications demand higher computational capabilities than before and performance gains are no longer driven by clock speed. Chip manufacturers are trying to meet this demand by adding multiple cores on a chip equipped with multiple levels of cache hierarchy and enhanced SIMD extensions. In this paper we have tried to optimize Bilateral filtering kernel using both high level and low level parallelism. Our interest is to develop an efficient parallelization method for modern multicore architectures that can be applied to any spatially invariant and inseparable filter kernel and potentially to any compute-intensive regular numeric application. We have also documented why traditional optimization methods for multicores are not as effective for Bilateral filtering kernel compared to memory intensive kernels. It can be seen from Figure~\ref{fig:comparisonIntel}, and Figure~\ref{fig:comparisonAMD} that a stencil kernel could be optimized by using SSE intrinsics even though the compiler optimizations are in place. Moreover, the reduction methods can provide a decent speedup where applicable. These findings will guide the auto optimizers and auto tuners to include SSE intrinsics as an essential optimization approach.

\bibliographystyle{IEEEtran} \bibliography{paper} \end{document}

### 2012-03-10 LateX/paper55.tex

%% LyX 2.0.1 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english,conference]{IEEEtran}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{color}
\usepackage{array}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{esint}
%\usepackage{subscript}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

\floatstyle{ruled}
\newfloat{algorithm}{tbp}{loa}
\providecommand{\algorithmname}{Algorithm}
\floatname{algorithm}{\protect\algorithmname}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\newenvironment{lyxcode}
{\par\begin{list}{}{
\setlength{\rightmargin}{\leftmargin}
\setlength{\listparindent}{0pt}% needed for AMS classes
\raggedright
\setlength{\itemsep}{0pt}
\setlength{\parsep}{0pt}
\normalfont\ttfamily}%
 \item[]}
{\end{list}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.

\usepackage{cite}\@ifundefined{definecolor}
 {\usepackage{color}}{}
\usepackage{amsfonts}\usepackage{mathtools}\usepackage{multirow}\usepackage{subfigure}\usepackage{tikz}\usepackage{pgf}\usepackage{pgfplots}\DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}
%\pgfplotstableread{Experiments/noexp.dat}\noexp

\makeatother

\usepackage{babel}
\begin{document}

\title{Stencil processing on single-core, multicore, and GPU architectures}

\author{\IEEEauthorblockN{Sami Wilf, Dinesh Agarwal, and Abinashi Dhungel}
\IEEEauthorblockA{ Georgia State University\\
 Department of Computer Science\\
 Atlanta, Georgia 30303\\
 } }
\maketitle
\begin{abstract}
This work explores multi-core parallelization and introduces a pair-symmetric
algorithm for stencil codes. Specifically, we use the bilateral filter
as our example stencil code for which we implemented both our pair-symmetric
algorithm and the standard naive stencil code algorithm. The bilateral
filter is a non-iterative image processing filter used to smooth images
without blurring pictorial edges. While it is an embarrassingly parallel
algorithm, it is still extremely compute intensive. In this paper,
we focus on both reducing and mitigating the effects of this computational
intensity and provide methods for improving the performance of bilateral
filtering, which consist of the elimination of redundant calculations
using our pair-symmetric algorithm implemented in Nvidia\textquoteright{}s
Compute Unified Device Architecture (\textquotedblleft{}CUDA\textquotedblright{}),
the use of low level single instruction, multiple data (\textquotedblleft{}SIMD\textquotedblright{})
parallelism, and the use of thread level parallelism and reduction.
We also present empirical data evidencing the performance gains we
achieved across the variety of architectures. The hardware used consists
of Nvidia GTX280, Advanced Micro-devices Devices (``AMD'') Barcelona,
AMD Shanghai, AMD Phenom, Intel Harpertown, Intel Lynnfield (Core
i7) quad core, and Intel Nehalem 32 core machines. The best speedup
achieved was a 235.5x speedup by our CUDA-based implementation of
our pair-symmetric stencil algorithm running on Nvidia's GTX280. Our
CPU multicore implementations resulted in a speedup of up to 38x using
16 cores of AMD Barcelona each with 4-stage vector pipeline, up to
23x using 8 cores for Intel Harpertown compared to a compiler optimized
code exploiting 4-stage vector unit of a core, we further demonstrate
speedup of up to 26x using 32 cores of Intel Nehalem multicore architecture.
\end{abstract}
% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}

\label{sec:intro} The recent advances in parallel computing have
provided a series of avenues for achieving significant increases in
performance. However, unlike the days when applications automatically
became faster with increased clock frequency, multicore chips require
programmers and researchers to revisit their algorithm designs to
harness the power of underlying architectures. Moreover, there is
no well defined list of guidelines that can be followed for each application.
This makes it more challenging to deploy existing applications, even
fundamental ones, to a myriad of available multicore architectures.

As reported in \cite{landscapeofpc} one of the critical questions
for parallel computing community is to identify the common kernels
being used in various applications. One such kernel is the Bilateral
filter, a non-iterative image processing kernel widely used to perform
directionally dependent image smoothing. In this paper we propose
a set of optimizations and a non-naive algorithm for stencil codes.
Specifically, we introduce a pair-symmetric stencil code algorithm
which has the theoretical potential to cut processing time in half.
We also propose architecture specific optimizations, such as by exploiting
the unique capabilities of special registers available in modern multicore
architectures and the rearrangement of data access patterns as per
the computations to exploit special purpose instructions. We also
propose optimizations for stencil code processing on Nvidia's CUDA,
with such optimizations including the utilization of CUDA's implicit
synchronization capability and the maximization of single-instruction
per multiple-thread efficiency.

we use the bilateral filter as our example stencil code\textcolor{blue}{{}
}\textcolor{black}{since it has not only been used for image processing
but also as a benchmarking kernel for various architectures\cite{Kamil2010}.
Moreover, because our optimizations are equally applicable to other
kernels such as 3D heat equation, matrix multiplication, and other
Jacobi iterations, our research will contribute to more than just
image processing applications.}

Our methods are effective over a range of multicore chips with varying
number of cores. In order to assess the effi{}cacy of our optimization
techniques, our experimental testbed includes a comprehensive set
of multicore architecture chips including the Nvidia GTX280, AMD quad-core
2376 (Shanghai), AMD quad-core 8350 (Barcelona), Intel quad-core Xeon
E5410 (Harpertown), Intel eight-core Xeon X7560 (Nehalem-EX), Intel
quad-core Core i7(Lynnfi{}eld) and AMD Phenom II six-core architectures.

\section{Bilateral Filtering Basics}

\label{sec:stencil} 
\begin{figure}[!h]
\centering \includegraphics[width=3.2in]{filter} \caption{Example of a filter at the starting position of the image}

\label{fig:filter} 
\end{figure}

Filtering is a fundamental image processing operation that finds its
application in a variety of tasks, such as filtering out noise, edge
detection, edge highlighting, and pre-processing tasks for higher
level image processing operations such as object recognition and face
recognition, among other things. Filtering is carried out by a relatively
small window, known as the filter (``Filter'' or ``Window''),
which iterates over the source image, performing the same execution
at each step of the iteration. The elements that the Filter's covers
in a given iteration makeup the neighborhood (``Neighborhood'')
and are neighbors (``Neighbors'' or ``Neighboring Pixels'') of
the Filter's center element (``Center'' or ``Center Pixel'').
\ref{fig:filter}illustrates the starting position of the Filter as
commonly seen in standard, naive stencil algorithms; the source image
is on the left and the destination image on the right is used store
the result of the Filter's computation at the respective Center's
position.

Our work explores the bilateral filtering kernel proposed by Tomasi
and Manduchi \cite{Tomasi1998} which is used to smooth an image without
blurring its edges. A faster approximation to the original algorithm
\cite{Paris2009} extends the image to 3D space by treating the pixel
intensity as the third dimension and using a linear shift-invariant
convolution in 3D. A separable implementation \cite{Pham2005} applies
the filter kernel first to one of the image dimensions and the intermediate
result is filtered again on other dimensions. This reduces the complexity
from $O(N*(m^{d}))$ to $O(N*d)$ where $N$ is the total number of
pixels in the image and $d$ is the dimensionality of image. However,
we do not parallelize these faster methods since our interest is to
develop a parallelization method that can be applied, in general,
to any spatially invariant and inseparable stencil code.

\textcolor{black}{Bilateral filtering combines two different kernels---the
spatial kernel and the photometric kernel. With the photometric kernel,
when the Center of the Filter rests on a pixel that boarders or is
a part of a pictorial edge, }\textit{\textcolor{black}{e.g.}}\textcolor{black}{,
an outline of an object or person against a contrasting background,
the Neighbors that make up the other (contrasting) side of the edge
have an insignificant impact on the Center Pixel's corresponding output
pixel, which is stored in the destination array storing the processed
image. Furthermore, when the Center is not bordering, and is not a
part of, a pictorial edge, the Neighbors contribute equally and the
noise is filtered out. Hence, pictorial edges maintain sharpness while
noise in the pictorial regions are filtered, resulting in more appealing
images. Mathematically, the Bilateral filter can be defined in terms
of the spatial kernel and photometric kernel as follows\cite{Tomasi1998}:}\\

Spatial kernel: 
\begin{equation}
h(x)=k_{d}^{-1}(x)\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}f(\xi)c(\xi,x)d\xi\label{spatial}
\end{equation}

\begin{equation}
k_{d}(x)=\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}c(\xi,x)d\xi
\end{equation}
 Photometric kernel: 
\begin{equation}
h(x)=k_{r}^{-1}(x)\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}f(\xi)s(f(\xi),f(x))d\xi\label{photometric}
\end{equation}

\begin{equation}
k_{r}(x)=\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}s(f(\xi),f(x))d\xi
\end{equation}

The combined kernel, which is the bilateral filter, is the product
of spatial kernel and photometric kernel and is described as: 
\begin{equation}
h(x)=k^{-1}(x)\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}f(\xi)c(\xi,x)s(f(\xi),f(x))d\xi\label{combined}
\end{equation}

\begin{equation}
k^{-1}(x)=\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}c(\xi,x)s(f(\xi),f(x))d\xi
\end{equation}

Function $c(\xi,x)$ measures the $geometric$ $closeness$ between
neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$
measures the $photometric$ $similarity$ (similarity in pixel intensity)
between the same. The influence of a Neighbor on the Filter's output
computation for a Center Pixel is calculated as a product of spatial
kernel and the photometric kernel. Thus, Neighbors with similar photometric
properties influence the Center Pixel more than those with less similar
photometric properties. This results in edge-preserved image smoothing,
which is the purpose of the bilateral filter. It should be noted,
however, that the foregoing equations are mathematical representations
for filtering of infinite-precision images. As infinite-precision
is a concept more pertinent to the analog domain and does not exist
in the digital domain, each double integral in equations 1-6 must
be understood as the summation of a finite set of calculations, \textit{i.e.,}
one calculation per Neighbor and one set of calculations per Center
Pixel, or put another way, one set of calculations per Filter iteration.
A formal expression is given below:

\begin{equation}
filtered\: pixil=\frac{\substack{N\\
\sum\\
i=1
}
x_{i}\cdot c(\xi,x)\cdot s(f(\xi),f(x))d\xi}{\substack{N\\
\sum\\
i=1
}
c(\xi,x)\cdot s(f(\xi),f(x))d\xi}
\end{equation}

In addition to restating the bilateral filter's mathematical model
in terms of a finite summation, Equation 7 reveals that data dependency
exists in the bilateral filter. Specifically, Equation 7 shows that
before the filtered pixel can be calculated by dividing the numerator
by the denominator, both the summation in the numerator and summation
in the denominator must first be calculated. Consequently, the bilateral
filter has two data dependencies. As explained later in this paper,
the pair-symmetric algorithm significantly increases the memory consumption
in order to manage data dependencies but pays off by cutting in half
the total number of executions.

\textcolor{black}{Another observation taken into account is that the
values of equation (\ref{spatial}) for the Neighborhood can be computed
in advance as the values depend on a simple Gaussian probability distribution
function. However, the values of equation (\ref{photometric}) and
consequently of equation (\ref{combined}) cannot be known }\textit{\textcolor{black}{a
priori}}\textcolor{black}{{} as they depend on the similarity of signals
between the Center Pixel and the Neighborhood. Moreover, as the size
of the Window increases, which only happens by choice of the user
as the size is user defined, the number of computations per pixel
grows at a rate of $O(FilterWindowWidth^{2})$ where }\textit{\textcolor{black}{FilterWindowWidth}}\textcolor{black}{{}
is the diameter of the Window. By reason of the foregoing, the bulk
of the time spent in processing an imagine using the bilateral filter
is in the arithmetic-intensive computations, because the number of
writes to the destination image is constant as the rate of growth
of Center-Neighbor computations with respect to the Window size increases.}

\section{Stencil Processing}

\subsection{A Naive Approach to Stencil Processing}

\begin{figure}
\centering\includegraphics[scale=0.35]{/Users/Sami/programming/ALGORITHM_1/bilat/the1}\includegraphics[clip,scale=0.35]{/Users/Sami/programming/ALGORITHM_1/bilat/the202}\includegraphics[scale=0.35]{/Users/Sami/programming/ALGORITHM_1/bilat/the2}\includegraphics[scale=0.35]{/Users/Sami/programming/ALGORITHM_1/bilat/the3}

\caption{Example of a naive bilateral filter iterations}
\end{figure}

The naive stencil algorithm is the standard model that is predominantly
used. This is evidenced by the fact that all of the authors of papers
cited herein either overlook or were unable to find improvements to
the naive algorithm itself. Those authors went straight to hardware
dependent implementations and expended their resources on achieving
performance gains that did not match the effort taken. Here, we address
a root problems of the standard naive algorithm by creating a new
algorithm, and then we also develop hardware dependent improvements
for greater speedup. Below is the pseudo-code for the standard naive
stencil kernel.
\begin{algorithm}[!h]
\caption{\label{alg:Example-Algorithm-float}Naive Stencil Kernel Algorithm}

\begin{lyxcode}
for~(each~element~\textit{e}\textsubscript{i,j}~in~source~image~\textit{\footnotesize s})\{~\\
\textit{~Center\textsubscript{i,j}}~=~\textit{e}\textsubscript{i,j}~\\
~for~(each~data~dependency\textit{~n})~\\
~~\textit{set~register~X\textsubscript{n}}to~0~\\
~\textit{/{*}Complete~dependency~computations{*}/}~\\
~Begin:~\\
~for(int~\textit{k~from}~\textit{-filter\_hw~to}~\textit{filter\_hw})~\\
~for(int~\textit{l}~from~\textit{-filter\_hw}~to~\textit{filter\_hw})\{~\\
~~for~(~each~data~dependency~\textit{n})~\\
~~\textit{X\textsubscript{n}}+=Funct\textit{\textsubscript{n}}(\textit{\footnotesize s}{[}\textit{Center\textsubscript{i,j}}{]},\textit{\footnotesize s}{[}\textit{Neighbor\textsubscript{i+k,j+l}}{]})~\\
~\}~\\
~End~\\
~\textit{/{*}Dependency~computations~completed{*}/}~\\
\textit{~/{*}obtain~output~from~dependency~data{*}/}~\\
~outImg{[}\textit{Center\textsubscript{i,j}}{]}=computeOutputPixel(\textit{\&X})~\\
\}

\end{lyxcode}
\end{algorithm}

\subsection{Pair-symmetric Stencil Processing}
\begin{itemize}
\item \textit{\textcolor{black}{Advantages: Reduces by half the amount of
Center-Neighbor computations, and can theoretically cut processing
time in half.}}
\item \textit{\textcolor{black}{Disadvantages: RequIres more memory which
increases with the amount of dependencies.}}
\item \textit{\textcolor{black}{Implementation obstacles: race-conditions}}
\end{itemize}
Our pair-symmetric stencil algorithm is premised on the fact that
each pixel assumes the role of both a Center Pixel and a Neighboring
Pixel and that stencil codes are symmetric with respect to Center-Neighbor
paired computations. For example, in the bilateral filter, $c(\xi,x)$
returns the same value as $c(x,\xi)$ and $s(f(\xi),f(x))$ returns
the same value as $s(f(x),f(\xi))$. It then follows that the naive
stencil algorithm dictated in Algorithm 1 above performs twice as
many calculations as are necessary.

Our pair-symmetric algorithm eliminates this redundancy that is inherent
in the naive stencil code algorithm. Instead of reading from all Neighbors,
our pair-symmetric algorithm reads from only half of the Neighbors.
The result of each Center-Neighbor paired computation is then added
to the \textit{X\textsubscript{n}}register, which eventually obtains
the result of data dependency \textit{n}. This is trivial as the naive
stencil algorithm already does this. What our algorithm does that
is unique is that the result of each Center-Neighbor pair computation
is also added to and stored in the Neighbor's position in a temporary
array (``Map'') that is used for storing the results of a data dependency.
So, there are \textit{N} Maps for \textit{N} number of dependencies.
After the data dependency conditions are met, the function that returns
the output data from the dependencies iterates over the Maps. As applied
to the bilateral filter, the data dependency lies in calculating each
pixel's numerator and denominator that is used to derive the pixel's
output result. After the numerators and denominators are derived,
a function iterates and devised the corresponding elements of the
two arrays to return the output. The pseudocode is given below:
\begin{algorithm}[h]
\noindent \caption{\label{alg:Example-Algorithm-float-1}Pair-symmetric Stencil Algorithm}

\begin{lyxcode}
\noindent \textit{for~(each~element~e\textsubscript{i,j}~in~source~image~}\textit{\footnotesize s}\textit{)\{}~\\
\textit{~~Center\textsubscript{i,j}~=~e\textsubscript{i,j}}~\\
\textit{~~for~(each~data~dependency~n)}~\\
\textit{~~~~float~X\textsubscript{n}=~0}~\\
\textit{~~for(~k~from~0~to~filter\_hw)}~\\
\textit{~~for(~l~from~0~to~filter\_hw)}~\\
\textit{~~for~(~each~data~dependency~n)~\{}~\\
\textit{~~~~temp~=~}

\textit{~~~~funct\textsubscript{n}(}\textit{\footnotesize s}\textit{{[}Center\textsubscript{i,j}{]},}\textit{\footnotesize s}\textit{{[}Neighbor\textsubscript{i+k,j+l}{]})}~\\
\textit{~~~~X\textsubscript{n}+=~temp}~\\
\textit{~~~~dependencyArray\textsubscript{n}{[}Neighbor\textsubscript{k,l}{]}~+=~temp}~\\
\textit{~~\}}~\\
\textit{~~for(~k~from~-1~to~-filter\_hw)}~\\
\textit{~~for(~l~from~-1~to~filter\_hw)}~\\
\textit{~~for~(~each~data~dependency~n)~\{}~\\
\textit{~~~~temp~=~}

\textit{~~~~funct\textsubscript{n}(}\textit{\footnotesize s}\textit{{[}Center\textsubscript{i,j}{]},}\textit{\footnotesize s}\textit{{[}Neighbor\textsubscript{i+k,j+l}{]})}~\\
\textit{~~~~X\textsubscript{n}+=~temp}~\\
\textit{~~~~dependencyArray\textsubscript{n}{[}Neighbor\textsubscript{k,l}{]}~+=~temp}~\\
\textit{~~\}}~\\
\textit{~~for~(each~data~dependency~n)~}~\\
\textit{~~~~dependencyArray\textsubscript{n}{[}Center\textsubscript{i,j}{]}~+=~X\textsubscript{n}}~\\
\textit{\}}~\\
\textit{for~(each~element~e\textsubscript{i,j}~of~output~image)}~\\
\textit{output{[}e\textsubscript{i,j}{]}=~computeOutput(\&dependencyArray)}

\end{lyxcode}
\end{algorithm}

\section{CUDA-based Implementation Details}

\label{sec:GPUoptimizations} As stencil codes are a class of embarrassingly
parallel algorithms, CUDA is well suited for its implementation. However,
in order to \textit{\textcolor{black}{effectively}} implement a stencil
algorithm in CUDA, one must understand how CUDA operates, which is
significantly different from any CPU.

\subsection{Implicit Intra-Warp Synchronization and Memory Access Patterns}

\begin{figure}

\caption{Illustration of CUDA implementation of pair-symmetric processing}

\includegraphics[width=1\columnwidth]{project22.jpg}

\end{figure}
CUDA's intra-warp implicit synchronization capability is the cornerstone
of our CUDA implementation of our pair-symmetric algorithm. Without
implicit synchronization, our pair-symmetric algorithm would render
useless for CUDA because it would require expensive synchronization
calls per Center-Neighbor Filter computation.

In CUDA, threads are organized into groups called warps. Each warp
executes one instruction at a time for its own group of threads. Essentially,
all the threads in a warp share the same program counter. Because
all threads in a warp share the same program counter and are therefore
driven by the same sequential instruction stream, all threads in a
warp are always in sync with each other without requiring explicit
synchronization calls. Therefore, synchronization between threads
in a warp is implicit.

To take full advantage of CUDA's implicit intra-warp synchronization
feature, divergent branching between threads in the same warp must
be avoided. This is so because all the threads in the same warp share
only one program counter. So, a divergent branch will cause some or
most of the threads to be idle as the program counter runs through
the branch and drives only those threads that meet the branch condition
to work. An example of a divergent branch can be a simple if-statement
such as: if (threadIdx.x < 5)\{...\} else\{...\}).

As it relates to stencil codes, divergent branching will commonly
occur in memory access patterns in which global memory operations
are not coalesced and shared memory operations are not free of memory
bank conflicts. For example, if eight threads were to read from eight
non-contiguous segments of global memory--such as that which is commonly
seen in strided access patterns like array{[}threadIdx.x {*} 2{]}--divergent
branching would occur because a warp of threads can only read from
one segment of memory per instructional transaction, and that segment's
word-length is hardware defined. In other words, a memory transaction
will be serialized unless all the threads in the warp to which the
transaction belongs are aligned consecutively and fit within a segment
of memory. Therefore, in order for all the threads of a warp to read/write
from global memory in a single transaction, the threads must correspond
read/write operations to aligned contiguous addresses within a segment
of memory. What this means in the implementation of stencil codes
is that warps should correspond to horizontal rows of pixels for an
i{*}jMax+j representation of a 2d image, for if threads in a warp
were to operate on pixels in a vertical order, there would be a separate
transaction for each thread in the warp, and if, for example, there
were two rows such that half of the threads were reading from memory
on each row, then there would be two transactions instead of one--i.e.,
inefficiency.

While intra-warp synchronization is guaranteed, the rate at which
warps stream through their sequence of instructions is unpredictable.
As a consequence, work must be arranged such that warps can work as
independent from each other as possible in order to minimize explicit
synchronization calls. This pooling of work is achieved by threads
reading from addresses incrementally along the horizontal axis, and
only making an explicit synchronization call per vertical shift to
the next row once the work requiring the warp's access to the instant
row of input data has completed. This design works because horizontal
shifts are implicitly synchronized as there is only one warp per row,
but a vertical shift without an explicit synchronization call will
result in a warp trespassing into the territory of another warp, thereby
causing race conditions as threads from two different warps will consequently
be prone to write to the same memory location.

\subsection{Shared Memory and Tile Processing}

CUDA features three types of memory with read/write capability suitable
for general-purpose computing on GPU: (1) global memory, which is
accessible by all threads in all thread blocks; (2) shared memory,
which is provided separately for each block of threads to share among
themselves; and (3) registers, which are separate and distinct for
each thread and of which each is only accessible by the thread to
which it belongs. Of these three types of memory, the slowest is global
memory, which unfortunately is where the host must first send the
input data before the GPU can begin its work. So, in order to see
any meaningful performance in CUDA, data stored in global memory must
be transferred to faster memory--i.e., shared memory, which is hundreds
of times faster. However, the trade-off is that the shared memory
size is much smaller than the global memory size, and this fact compels
a tile-based approach to stencil processing.

Our tile-based stencil processing approach is self-explanatory. Equal
sized partitions of the input data are processed, and then any remaining
data on and between the edge and the last partition that is closest
to (but not falling over) the edge boundary of the 2-dimensional data
representation is processed. With respect to the naive stencil code
algorithm, tile-based stencil processing is trivial and needs no further
explanation because simply partitioning the space and processing it
will suffice. However, for the pair-symmetric algorithm that we propose,
tile-based stencil processing requires adjustments in the order in
which tiles are processed in order to guarantee the exclusion of race
conditions.

\begin{figure}
\caption{\label{fig:Tiling-for-Pair-Symmetric}Tiling for Pair-Symmetric Stencil
Processing:}

\includegraphics[width=1\columnwidth]{\string"stacked tiles\string".jpg}
\end{figure}

If processes were assigned to adjacent tiles, race conditions would
occur because the edge cells of the tiles, a by product of the Filter's
half-width, would overlap over each other, as illustrated in \ref{fig:Tiling-for-Pair-Symmetric}.
To get around this problem, simultaneously processed tiles can be
spaced apart in strides such that no two tiles being processed at
the same time intersect with one another. An example of such a technique
is illustrated in \ref{fig:An-iteration-of}(the tiles are flipped
sideways in this rendering).

\begin{figure}
\includegraphics[width=1\columnwidth]{lenatiles}\caption{\label{fig:An-iteration-of}An iteration of processed tiles of Lena}
\end{figure}

\section{CPU-based Implementation Details}

\label{sec:optimizations} For regular numerical scientific computations
performance improvements in general can be achieved by parallelizing
the code either at low-level (micro-level or instruction level) or
at high-level (thread-level or task-based). For instance, $Intel$'s
Streaming SIMD Extension (SSE) is a technique to exploit micro-level
parallelism in x86 based processors using SIMD capabilities. SSE can
be ideal for tasks that involve repetitive computation such as the
3D heat equation defined in section \ref{sec:stencil} or any other
stencil kernel such as Jacobi iterations.

Researchers paid attention to the use of SSE instructions for scientific
applications almost a decade ago. Fundamental scientific kernels were
implemented to utilize SSE instructions which lead to some speedup.
For instance the matrix multiplication kernel was implemented in \cite{matrix-multi,MatrixMultiFloating}
for matrices with integers and floating point numbers. SSE instructions
have also been used for image or template matching algorithms such
as normalized cross correlation (``NCC'') \cite{NCCmatching}.

Task level parallelism can be achieved by utilizing multiple cores
in a system by using standard parallelization libraries such as Threaded
Building Block (``TBB''), OpenMP etc. Other traditional optimization
techniques such as loop unrolling, and prefetching etc. help to transform
the code into instruction level parallelism friendly code. Although,
the modern compilers usually try to improve the performance by automatically
vectorizing, unrolling loops, prefetching, and SIMDization, we were
still able to get significant performance improvements alongside compiler
optimizations.

\subsection{SIMD optimizations}

\label{sub:SIMD1} Modern day processors support SIMD optimizations
where the programmer can operate on multiple data elements in parallel
by storing the data into special purpose registers formerly known
as MMX\cite{MMX}, were introduced by Intel in 1996 to enhance multimedia
applications by exploiting their inherent parallelism. MMX has 8 64-bit
wide registers, which are consecutively named from MM0 to MM7. These
registers can hold 8 different bytes such as characters, or 4 words
such as short integers, 2 double words such as long integer, or single
precision float, or a single quad word such as double precision float.
A single MMX instruction can simultaneously process information stored
in these MMX registers, and therefore have the potential to provide
a speedup of up to 8 depending on the data type loaded into the registers.

Modern day CPUs use SSE that is an extension of MMX technology and
is fundamentally similar to MMX. It has 8 128-bit registers named
XMM0 through XMM7. SSE instructions can also operate in parallel on
data packed in these registers thus providing a theoretically maximum
speedup of 16. There are other extensions of SSE such as SSE2, SSE3,
and SSE4.2. Moreover, recent extension of SSE named Advanced Vector
Extensions (``AVX''), uses 256-bit data registers and thus a theoretical
peak speedup of 32 is possible.

Our implementation explores use of special purpose registers using
both assembly level programming and SSE intrinsics. The advantage
of using assembly language is that a programmer can have explicit
control of underlying hardware. However, programmer productivity is
affected and an in-depth knowledge of underlying hardware is essential.
Although, SSE intrinsics are wrappers for assembly language, it is
preferred to use SSE intrinsics due to ease of coding and also due
to efficient utilization of available assembly level instructions
without knowing the low level details of underlying architecture.
For instance, instructions that operate on aligned data usually perform
better than instructions that operate on unaligned data. A programmer,
working with intrinsics, can specify the intrinsic and the appropriate
instruction will automatically replace this intrinsic during compilation.
Moreover, there are some operations for which there are no assembly
instructions and hence in case of assembly level programming the code
size can grow to such an extent that readability gets compromised.

\subsection{Reduction Methods}

\label{sub:reduction} Sum by reduction is a technique to compute
sum of multiple elements usually stored in contiguous memory elements.
There are assembly instructions that can add horizontally packed,
single precision, floating point numbers. We used $haddps$ assembly
instruction that takes two (source and destination) 128 bit operands,
and adds the single precision floating point numbers stored in first
and second dword of destination and source operand and stores it in
first and third dword of the destination operand respectively. Similarly,
the third and fourth dword of destination and source operand are added
and stored in second and fourth dword of the destination operand respectively.
Both source and destination values must be stored in MMX registers.
The SSE intrinsic to perform horizontal add is $\_mm\_hadd\_ps$.
It takes two 128 bit operands, each of which is treated as four 32-bit
floating point elements. The result of this operation on operand X(X0,
X1, X2, X3) and Y(Y0, Y1, Y2, Y3) is (Y3 + Y2, Y1 + Y0, X3 + X2, X1
+ X0).

\section{Experimental Testbed}

\label{sec:expt}

\begin{table*}[t]
\begin{tabular}{|l|p{1.2cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{3.5cm}|l|p{2cm}|l|p{0.9cm}|}
\hline 
\textbf{Model}  & \textbf{Clock frequency}  & \textbf{cores per chip}  & \textbf{chips per node}  & \textbf{cores per node}  & \textbf{SIMD support}  & \textbf{Cache Memory}  & Compiler &  & \tabularnewline
\hline 
Intel Xeon E5410  & 2.33 GHz  & 4  & 2  & 8  & SSE, SSE2  & 12 MB  & icpc 11.1 &  & \tabularnewline
\hline 
Intel Xeon X7560  & 2.27 GHz  & 32  & 1  & 32  & SSE, SSE2  & 24 MB & icpc 11.1 &  & \tabularnewline
\hline 
Intel Core i7-870  & 2.93 GHz  & 4  & 1  & 4  & SSE, SSE2, SSE3, SSE4.2  & 8MB & icpc 12.0.2  &  & \tabularnewline
\hline 
AMD Opteron 2376  & 2.3 GHz  & 4  & 2  & 8  & SSE, SSE2  & 6MB  & icpc 11.1 &  & \tabularnewline
\hline 
AMD Opteron 8350  & 2.0 GHz  & 4  & 4  & 16  & SSE, SSE2  & 2MB  & icpc 11.1 &  & \tabularnewline
\hline 
AMD Phenom II X6 1045T  & 2.7 GHz  & 3  & 1  & 6  & SSE, SSE2, SSE4a  & 6MB  & icpc 12.0.2  &  & \tabularnewline
\hline 
\end{tabular}\caption{Architectural details of multicore chips employed for experiments}

\label{tab:archs} 
\end{table*}

In order to evaluate our implementation we have tried it on a range
of multicore chips. Table \ref{tab:archs} summarizes architectural
details of systems we used during our experiments.

\subsection{Hardware}

\label{subsec:hw}

\subsubsection{Intel Xeon}

\label{ss:xeon} We have used two different Xeon chips, E5410 (Harpertown)
and X7560 (Nehalem EX). Harpertown is a dual-socket, dual-processor
relatively old system with a clock frequency of 2.33 GHz. Our computing
node has two Harpertown chips with total number of cores being eight.
It does not have Intel's hyperthreading (HT) technology which means
the maximum number of threads is equal to the number of cores available.
Intel Nehalem EX is considerably newer chip having eight-cores per
processor with each core having clock frequency of 2.27 GHz. We had
access to a Linux cluster from Intel's Manycore Testing Lab (MTL)
containing four sockets of Nehalem EX chips thereby making 32 cores
available to experiment with. The processors are connected to each
other via Intel QPI (Quickpath interconnect) links that can transfer
data at 6.4 GT/s. Both of these machines support 64 bit instructions.

\subsubsection{AMD Opteron}

\label{ss:opteron} AMD Opteron 2376 (Shanghai) chip has private L1
and L2 caches of size 128KB (64 D + 64 I) and 512KB respectively per
core along with a 6MB cache shared among all of the cores. It has
a clock frequency of 2.30 GHz and it is a quad-core architecture.
Our compute node has two of Shanghai chips making it a total of eight
cores to be used for experiments. AMD Opteron 8350 (Barcelona) is
similar to Shanghai in many aspects but it has a clock frequency of
2.0 GHz and the shared cache is only 2MB. Our compute node contains
4 of Barcelona chips and thus we are able to run 16 threads on this
node.

\subsubsection{Intel Core i7}

\label{ss:corei7}Intel Core i7 - 870 code named Lynnfield supports
SSE 4.2 and has AVX technology. The clock frequency is an impressive
2.93 GHz and it can operate at a maximum of 3.6 GHz thanks to Intel's
power boost technology. Lynnfield supports hyperthreading which lets
an application to execute 8 threads. It contains 2 memory channels
that provide a theoretical maximum memory bandwidth of 21 GB/s.

\subsubsection{AMD Phenom II}

\label{ss:phenom} AMD Phenom is a six core processor with each core
capable of operating at 2.7 GHz. If only three cores are in operation
it can run at a maximum of 3.2 GHz. It boasts 128 KB (64 D + 64 I)
L1 cache, and 512 KB L2 cache per core in addition to a 6MB L3 cache
shared among all six cores. The cache latencies are 3, 13, and 47
clock cycles for L1, L2, and L3 caches respectively. The cache line
size is 64 bytes, L1 caches are 2 way set associative, L2 caches are
16-way set associative and L3 cache is 48-way set associative.

\subsection{Software Framework}

\label{subsec:sw} Our implementation is written in $C++$ and uses
OpenMP for parallelization. To enrich supported image formats by our
implementation we use $CImg$ library. Intel Vtunes Amplifier XE was
used to find out hotspots in our code.

\subsubsection{Algorithms}

We have implemented a number of variations of bilateral filtering
kernel, results of which are reported in section \ref{sec:results}.
These algorithms are labeled as follows: 
\begin{itemize}
\item $BL\_Naive$ is the sequential version of the code, as described by
\cite{Tomasi1998}. 
\item $BL\_Naive\_No\_Exp$ is similar to sequential version in all aspects,
except the fact the most time consuming $exp$ instruction, discussed
after this list of algorithms, is commented out. 
\item $BL\_Tiles$ is the version that employs data blocking while processing
the input image. 
\item $BL\_Tiles\_No\_Exp$ is to $BL\_Tiles$ what $BL\_Naive\_No\_Exp$
is to $BL\_Naive$. 
\item $BL\_Assembly$ is the version of bilateral filtering implemented
using assembly level code. 
\item $BL\_A\_Reduction$ is the version of bilateral filtering that uses
assembly language as well as reduction optimizations. 
\item $BL\_Intrinsic$ version uses SSE intrinsics instead of assembly language. 
\item $BL\_I\_Reduction$ is the extended version of $BL\_Intrinsic$ that
uses the intrinsics that can compute sum of two 128 bit operands in
parallel. 
\end{itemize}
$BL\_Naive$ has a redundant calculation inside the innermost loop
which has been moved out of loop for all other versions. The original
instruction involving this calculation looks like\\
 $gaussian\_ph=exp(-(pd)/(2*(sigma\_ph*sigma\_ph)));$\\
 After moving the redundant calculation out of the loop the same instruction
becomes\\
 $gaussian\_ph=exp(pd*pre\_calc\_sigma\_sq));$\\

\subsubsection{Input Image}

In order to analyze the behavior of our kernel with varying image
sizes, we have executed our kernel with two different images. One
that fits into cache easily and the other that does not fit into cache
but fits into main memory. Both of the images were converted to single
spectrum images represented by single precision numbers for each pixel.
The smaller image was 512 pixels wide and 512 pixels high, thus requiring
1MB memory to store it. As shown in table \ref{tab:archs} all of
the systems have cache memories bigger than 1 MB. The second image
was 3000 pixels wide and 3000 pixels high and hence requires little
more than 34 MB of memory to store it. None of the cache memories
were such big to store this image entirely.

\section{Performance Results}

\label{sec:results}

\begin{figure*}[t!]

\begin{centering}
\subfigure[Small Image]{ \includegraphics[width=3.2in,height=2in]{noexpsmall}
\label{fig:smallnoexp} } \subfigure[Large Image]{ \includegraphics[width=3.2in,height=2in]{noexplarge}
\label{fig:largenoexp} } 
\par\end{centering}

\caption{Percentage of time taken by sequential kernel when run without $exp$
instruction, $filter$ $radius$=10.}

\label{fig:noexp} 
\end{figure*}

For all of our experiments the filter radius was kept at 10, subsequently
making the Window of size 21 x 21. The spatial spread and the photometric
spread were also set to 10. The optimization flag O3 was used for
all experiments in addition to SSE2 for SIMDization. For Core i7 chip
compiler flag SSE4.2 was used instead of SSE2. We do not use hyperthreading
which means there are no more than one thread per core. 
\begin{figure*}[t!]

\begin{centering}
\subfigure[Nehalem-EX Xeon X7650]{ \includegraphics[width=3.2in,height=2.3in]{nehalem}
\label{fig:Nehalem} } \subfigure[Harpertown Xeon E5410]{ \includegraphics[width=3.2in,height=2.3in]{harpertown}
\label{fig:harpertown} } \subfigure[Core i7 - 870]{ \includegraphics[width=3.2in,height=2.3in]{corei7}
\label{fig:corei7} } 
\par\end{centering}

\caption{Performance of different algorithms for bilateral filtering kernel
on Intel chips for both small image and large image. Sequential version
(Small\_Sequential, Large\_Sequential) does not use OpenMP while Parallel
version (Small\_Parallel, Large\_Parallel) uses all available cores.
The comparison is with BL\_Naive version, auto optimized by compiler
with -O3 and -xSSE2 flags. }

\label{fig:comparisonIntel} 
\end{figure*}

\begin{figure*}
\begin{centering}
\subfigure[Barcelona AMD Opteron 8350]{ \includegraphics[width=3.2in,height=2.3in]{barcelona}
\label{fig:barcelona} } \subfigure[Shanghai AMD Opteron 2376]{
\includegraphics[width=3.2in,height=2.3in]{shanghai} \label{fig:Shanghai}
} \subfigure[AMD Phenom II 1045T]{ \includegraphics[width=3.2in,height=2.3in]{phenom}
\label{fig:Phenom} } 
\par\end{centering}

\caption{Performance of different algorithms for bilateral filtering kernel
on AMD chips for both smaller image and larger image. Sequential version
(Small\_Sequential, Large\_Sequential) does not use OpenMP while Parallel
version (Small\_Parallel, Large\_Parallel) uses all available cores.
The comparison is with BL\_Naive version, auto optimized by compiler
with -O3 and -xSSE2 flags.}

\label{fig:comparisonAMD} 
\end{figure*}

\begin{figure*}[t!]

\begin{centering}
\subfigure[Harpertown Xeon E5410]{ \includegraphics[width=3.2in,height=2.5in]{harpertownfinal}
\label{fig:harpertownFinal} } \subfigure[Core i7 - 870]{ \includegraphics[width=3.2in,height=2.5in]{corei7final}
\label{fig:corei7Final} } \subfigure[Nehalem-EX Xeon X7650]{ \includegraphics[width=4.2in,height=3.5in]{nehalemfinal}
\label{fig:NehalemFinal} } 
\par\end{centering}

\caption{Comparison of combined speedup due to SIMDization and OpenMP parallelization
on Intel chips for different algorithms}

\label{fig:comparisonFinalIntel} 
\end{figure*}

\begin{figure*}
\begin{centering}
\subfigure[Barcelona AMD Opteron 8350]{ \includegraphics[width=3.2in,height=2.5in]{barcelonafinal}
\label{fig:barcelonaFinal} } \subfigure[Shanghai AMD Opteron 2376]{
\includegraphics[width=3.2in,height=2.5in]{shanghaifinal} \label{fig:ShanghaiFinal}
} \subfigure[AMD Phenom II 1045T]{ \includegraphics[width=3.2in,height=2.5in]{phenomfinal}
\label{fig:PhenomFinal} } 
\par\end{centering}

\caption{Comparison of combined speedup due to SIMDization and OpenMP parallelization
on AMD chips for different algorithms.}

\label{fig:comparisonFinalAMD} 
\end{figure*}

Running the hotspot analysis it was realized that the most time consuming
instruction was the instruction involving computation of exponential
function for the photometric kernel. Figure~\ref{fig:noexp} shows
the percentage of time taken by $BL\_Naive\_No\_Exp$ and $BL\_Tiles\_No\_Exp$
on different platforms for both smaller image (figure~\ref{fig:smallnoexp})
and the larger image (figure~\ref{fig:largenoexp}). It can be seen
that after taking the $exp$ instruction out of the computation the
total sequential execution time is consistently less than 2\% of the
execution time with $exp$ instruction in place. For the smaller image
it is strictly less that 1\% of the time with $exp$ instruction in
place. To be precise it is 1.39\% maximum on Intel Nehalem for the
larger image and 0.65\% for the smaller image on Core i7 processor.
The lower memory transfer time for smaller image is due to the fact
that the smaller image sits into the cache memory entirely and hence
there are no repeated data transfers between processor and main memory.

Since the $exp$ instruction does not involve transfer of data from
main memory, it is logical to deduce from this experiment that memory
reuse and memory access pattern related optimizations are not of much
help to speed up bilateral filtering kernel. What Amdahl's law claims
for inherently sequential portion of the code can be applied here.
\\
 \textit{The total speedup achieved by optimizations that deal with
memory management is limited by the amount of time spent in memory
transfers}. Therefore, in this case traditional optimizations such
as tiling/blocking, cache efficient initializations, prefetching,
and other NUMA optimizations can only result in a speed up of at most
2\%.

Figure~\ref{fig:comparisonIntel} compares different algorithms on
different Intel chips. In this experiment we are comparing the speedup
achieved by the respective algorithm compared to BL\_Naive algorithm.
It is important to note that these results are for micro level parallelism
(use of 128 bit special purpose registers) only and do not consider
speedup due to high level (OpenMP) parallelization. We have used parallel
version, that utilizes all available cores on that chip, only to show
that the speedup is still bounded by the theoretical peak speedup
which happens to be four here since we work with 32bit floating point
numbers. Interestingly, since we use compiler optimization flags such
as -O3 and -xSSE2 there is automated vectorization taking place in
all the algorithms. However, our hand coded vectorization is superior
in most of the cases.

The speedup shown by BL\_Tiles is mostly due to the computation reduction
as discussed in section \ref{subsec:sw} because we do not use special
purpose registers in this algorithm. It can be seen from figure~\ref{fig:comparisonIntel}
that the parallelism due to SIMDization is limited by four which is
the theoretical maximum possible speedup due to SSE registers as discussed
in section \ref{sub:SIMD1}.

As seen in figure~\ref{fig:Nehalem} the maximum speedup achieved
on Nehalem architecture only using SSE optimization is 2.29x for the
smaller image and 1.33x for the larger image. Smaller image generally
performs better because the time to copy data to SSE registers is
less as the entire image is always available in the cache memory.
The BL\_Naive algorithm outperforms our assembly level implementations
but it is slower than our intrinsic based implementations in all cases.
The reason for this is that the auto SIMDization by compiler is more
efficient than our implementation for Nehalem architecture. Intel
Harpertown, which is a relatively older chip, shows an impressive
speedup of 3.54x (figure~\ref{fig:harpertown}) even though compiler
optimizations are still in place. Another interesting observation
is the results on Intel Core i7 architecture (figure~\ref{fig:corei7}).
The reason for this behavior is the newer compiler. These experiments
were run with $SSE4.2$ compiler flag and hence the compiler is able
to exploit SIMDization automatically thus making it difficult to surpass
it using assembly level coding. However, our algorithms that use SSE
intrinsics and reduction, i.e. BL\_A\_Reduction, BL\_Intrinsic, and
BL\_I\_Reduction are able to provide greater speedups than those achieved
by compiler optimizations as shown in figures~\ref{fig:comparisonIntel},
and \ref{fig:comparisonAMD}.

Figure~\ref{fig:comparisonAMD} shows the same comparison across
AMD chips. The results are in tune with those for Intel platform.
As shown in figure~\ref{fig:barcelona} for Barcelona, which is a
relatively older chip, the maximum speedup is 3.24x for BL\_I\_Reduction
algorithm. Similarly, for AMD Shanghai, as shown in figure~\ref{fig:Shanghai},
the maximum speedup achieved by exploiting SIMDization is 3.02x for
BL\_I\_Reduction algorithm. The case of AMD Phenom is shown in figure~\ref{fig:Phenom}.
It can be seen that similar to Intel Core i7 and Intel Nehalem architecture
the newer compiler by Intel affects the performance here too. The
auto tuning applied by the compiler outperforms the manual tuning
in case of BL\_Assembly and BL\_A\_Reduction. However, BL\_Intrinsic
and BL\_I\_Reduction are still able to outperform compiler optimizations.
The maximum speedup achieved in this case is 1.30x.

Figure~\ref{fig:comparisonFinalIntel} shows the speedup with increasing
number of OpenMP threads for the larger image. It is interesting to
see how speedup due to OpenMP multiplies with the speedup due to SSE
optimizations to exhibit super linear performance. Although the input
images are not very large to provide enough grain to a machine with
32 cores, the kernel still scales well. For other architectures the
kernel demonstrates super-linear behavior.

Figure~\ref{fig:harpertownFinal} shows the performance of bilateral
filtering kernel on Intel Harpertown. This machine has 8 cores and
the maximum combined speedup achieved is 20.91x. The speedup achieved
only using only SSE optimization was 2.69x and the speedup achieved
by BL\_Naive algorithm using 8 cores is 7.73x.

Intel Nehalem chip with 32 cores does not show a super-linear speedup
as shown in figure~\ref{fig:NehalemFinal}. However, the reason behind
this is the lack of enough parallelism that can be afforded by the
kernel for this image. Both optimizations still multiply their individual
speedup to result in the combined speedup. The total combined speedup
is 26.46x using 32 cores, the speedup only due to SSE was 1.12x and
the speedup for BL\_Naive is 23.47x.

Intel Core i7 chip (figure~\ref{fig:corei7Final}) demonstrates the
similar behavior too. The final speedup of 4.63x, using 4 cores, is
a result of speedup due to SSE which is 1.33x and speedup due to OpenMP
which is 3.48x. Figure~\ref{fig:comparisonFinalAMD} shows a similar
comparison over various AMD chips. Figure~\ref{fig:barcelonaFinal}
shows the speedup for Barcelona chip. This chip has 16 cores with
a combined maximum speedup being 38.03x. The speedup achieved using
SSE optimizations was 2.48x and the speedup for BL\_Naive was 15.34x
and thus these both realize a speedup of 38.03x using 16 cores.

AMD Shanghai similarly shows an impressive speedup of 20.92x (figure~\ref{fig:ShanghaiFinal}).
The interesting case is with AMD Phenom as shown in figure~\ref{fig:PhenomFinal}.
The speedup drops a little as number of cores increases beyond three.
The reason behind this is the turbo core technology as discussed in
section (\ref{ss:phenom}). When there are a maximum of three cores
being used the clock frequency scales up to 3.6 GHz but as soon as
more threads are created the clock frequency steps down to 2.7 GHz.

\section{Conclusion and future work}

\label{sec:conclusion} Modern day scientific applications demand
higher computational capabilities than before and performance gains
are no longer driven by clock speed. Chip manufacturers are trying
to meet this demand by adding multiple cores on a chip equipped with
multiple levels of cache hierarchy and enhanced SIMD extensions. In
this paper we have tried to optimize bilateral filtering kernel using
both high level and low level parallelism. Our interest is to develop
an efficient parallelization method for modern multicore architectures
that can be applied to any spatially invariant and inseparable filter
kernel and potentially to any compute-intensive regular numeric application.
We have also documented why traditional optimization methods for multicores
are not as effective for bilateral filtering kernel compared to memory
intensive kernels. It can be seen from figure~\ref{fig:comparisonIntel},
and figure~\ref{fig:comparisonAMD} that a stencil kernel could be
optimized by using SSE intrinsics even though the compiler optimizations
are in place. Moreover, the reduction methods can provide a decent
speedup where applicable. These findings will guide the auto optimizers
and auto tuners to include SSE intrinsics as an essential optimization
approach.

\begin{figure}[h!]
\includegraphics[width=3.2in]{Gpu} \caption{Execution time of bilateral filtering on GPU with increasing number
of threads per block for larger image, $filter$ $radius$=10, $spatial$
$spread$=10, $photometric$ $spread$=10}

\label{fig:gpu} 
\end{figure}

We have also ported bilateral filtering to GPU. As shown in figure~\ref{fig:gpu}
we are able to use up to 90,000 threads per block with total number
of blocks being 10. Each thread is processing one pixel of the larger
image which has a height of 3000 pixels and width of 3000 pixels too.
The results are promising and we are exploring the software controlled
memory hierarchy and registers in an efficient ways as future work
of this research.

\bibliographystyle{IEEEtran}
\bibliography{paper}

\end{document}

### 2012-03-10 trunk/LateX/paper8.tex
test

--- 2012-03-10 LateX/paper5.tex
@@
+test
+
 \documentclass{IEEEtran}
@@
 \begin{document}
-
 \title{Acceleration of Bilateral filtering algorithm for manycore and multicore architectures}
-
 \author{\IEEEauthorblockN{Dinesh Agarwal, Sami Wilf, Abinashi Dhungel, and Sushil K. Prasad}\\
@@
  Atlanta, Georgia 30303\\
- } }
+ }}
 \maketitle
 \begin{abstract}
-This work explores multicore and manycore acceleration for embarrassingly parallel compute-intensive Bilateral filtering kernel.
+This work explores multicore and manycore acceleration for the embarrassingly parallel, compute-intensive Bilateral filtering kernel.
 For manycore architectures, we have created a pair-symmetric algorithm to avoid redundant calculations and for multicore architectures we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads.
@@
 
-As reported in \cite{landscapeofpc} one of the critical questions for parallel computing community is to identify the common kernels being used in various applications.
-One such kernel is the Bilateral filter, a non-iterative image processing kernel widely used to perform directionally dependent image smoothing.
-In this paper we propose a set of optimizations and a non-naive algorithm for Bilateral filtering kernel.
-Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to cut processing time in half.
-We propose architecture specific optimizations, such as exploiting the unique capabilities of special registers available in modern multicore architectures and the rearrangement of data access patterns as per the computations to exploit special purpose instructions.
-We also propose optimizations for Nvidia's CUDA, including utilization of CUDA's implicit synchronization capability and the maximization of single-instruction-multiple-thread efficiency.
+Our work explores efficient parallel implementation of Bilateral filtering kernel\footnotemark[1] proposed by Tomasi and Manduchi \cite{Tomasi1998}, on multicore and manycore architectures.
+Bilateral filtering kernel is used to smooth an image without blurring its edges.
+It is ubiquitous in various image processing applications in various contexts such as denoising\cite{zhang2008}, texture editing and relighting\cite{eisemann2004}, tone management\cite{Bae2006}, stylization\cite{DeCarlo2002}, optical-flow estimations\cite{Xiao2006}, and demosaicking\cite{Ramanath2003} etc.
 
-Our work explores the Bilateral filtering kernel proposed by Tomasi and Manduchi \cite{Tomasi1998}, which is used to smooth an image without blurring its edges.
-Bilateral filtering is ubiquitous in various image processing applications in various contexts such as denoising\cite{zhang2008}, texture editing and relighting\cite{eisemann2004}, tone management\cite{Bae2006}, stylization\cite{DeCarlo2002}, optical-flow estimations\cite{Xiao2006}, and demosaicking\cite{Ramanath2003} etc.
 A faster approximation to the original algorithm \cite{Paris2009} extends the image to 3D space by treating the pixel intensity as the third dimension and using a linear shift-invariant convolution in 3D.
-
 A separable implementation \cite{Pham2005} applies the filter kernel first to one of the image dimensions and the intermediate result is filtered again on other dimensions.
-This reduces the complexity from $O(N*(m^{d}))$ to $O(N*d)$ where $N$ is the total number of pixels in the image and $d$ is the dimensions of the image.
-Yang et.al.\cite{yang2009}, have reported a constant time algorithm by dividing the Bilateral filtering kernel into a constant number of time spatial kernels.
+This reduces the complexity from $O(N*(m^{d}))$ to $O(N*d)$, where $N$ is the total number of pixels in the image and $d$ is the dimensions of the image.
+Yang et al.\cite{yang2009} have reported a constant time algorithm by dividing the Bilateral filtering kernel into a constant number of time spatial kernels.
 However, we do not parallelize these methods since our interest is to develop a parallelization method that can be applied, in general, to any spatially invariant and inseparable filtering algorithm.
 
-Our methods are effective over a range of multicore chips with varying number of cores.
+In this paper, we propose a set of optimizations and a non-naive algorithm for Bilateral filtering kernel.
+Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to reduce processing time to half.
+We propose architecture specific optimizations, such as exploiting the unique capabilities of special registers available in modern multicore architectures and the rearrangement of data access patterns as per the computations to exploit special purpose instructions.
+We also propose optimizations pertinent to Nvidia's CUDA, including utilization of CUDA's implicit synchronization capability and the maximization of single-instruction-multiple-thread efficiency.
+
 In order to assess the efficacy of our optimization techniques, our experimental testbed includes a comprehensive set of multicore and manycore architecture chips including the Nvidia GTX280, AMD quad-core 2376 (Shanghai), AMD quad-core 8350 (Barcelona), Intel quad-core Xeon E5410 (Harpertown), Intel eight-core Xeon X7560 (Nehalem-EX), Intel quad-core Core i7, and AMD Phenom II six-core architectures.
-Our code is available under GNU GPL license from \url{http://code.google.com/p/blfilter/}.
+\footnotetext[1]{Our code is available under GNU GPL license from \url{http://code.google.com/p/blfilter/}.}
 
@@
 
-Bilateral filtering combines two different kernels---the spatial kernel and the photometric kernel.
+Bilateral filtering combines two different kernels - the spatial kernel and the photometric kernel.
 With the photometric kernel, when the Center of the Filter rests on a pixel that boarders or is a part of a pictorial edge, e.g., an outline of an object or person against a contrasting background, the Neighbors that make up the other (contrasting) side of the edge have an insignificant impact on the Center Pixel's corresponding output pixel, which is stored in the destination array storing the processed image.
@@
 \begin{equation}
-k^{-1}(x)=\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}c(\xi,x)s(f(\xi),f(x))d\xi
+k(x)=\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}c(\xi,x)s(f(\xi),f(x))d\xi
 \end{equation}
@@
 Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same.
-The influence of a Neighbor on the Filter's output computation for a Center Pixel is calculated as a product of spatial kernel and the photometric kernel.
+The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel.
 Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties.
@@
 It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images.
-As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration.
+As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e.
+one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration.
 A formal expression is given below:
@@
 \label{eq:depen}
-filtered\: pixel=\frac{\substack{N\\
+filtered\: pixel\: x=\frac{\substack{N\\
 \sum\\
@@
 }
-x_{i}\cdot c(\xi,x)\cdot s(f(\xi),f(x))d\xi}{\substack{N\\
+f(\xi_{i})\cdot c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}{\substack{N\\
 \sum\\
@@
 }
-c(\xi,x)\cdot s(f(\xi),f(x))d\xi}
+c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}
 \end{equation}
@@
 In addition to restating the Bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the Bilateral filter.
-Specifically, Equation~\ref{eq:depen} shows that before the filtered pixel can be calculated by dividing the numerator by the denominator, both the summation in the numerator and summation in the denominator must first be calculated.
+Specifically, Equation~\ref{eq:depen} shows that before the filtered pixel can be calculated by dividing the numerator by the denominator, both the summation in the numerator and denominator must first be calculated.
 Consequently, the Bilateral filter has two data dependencies.
@@
 
-Another observation taken into account is that the values of equation (\ref{spatial}) for the Neighborhood can be computed in advance as the values depend on a simple Gaussian probability distribution function.
-However, the values of equation (\ref{photometric}) and consequently of equation (\ref{combined}) cannot be known a priori as they depend on the similarity of signals between the Center Pixel and the Neighborhood.
+Another observation taken into account is that the values of Equation~\ref{spatial}, for the Neighborhood, can be computed in advance as the values depend on a simple Gaussian probability distribution function.
+However, the values of Equation~\ref{photometric} and consequently of Equation~\ref{combined} cannot be known a priori as they depend on the similarity of signals between the current Center Pixel and the Neighborhood.
 Moreover, as the size of the Window increases, which only happens by choice of the user as the size is user defined, the number of computations per pixel grows at a rate of $O(FilterWindowWidth^{2})$ where FilterWindowWidth is the diameter of the Window.
@@
 
-\subsection{A Naive Approach to Bilateral filetering Processing}
+\subsection{A Naive Approach to Bilateral filtering kernel}
 
@@
 The naive stencil algorithm is the standard model that is predominantly used in Bilateral filtering kernel.
-Previous implementations\cite{Paris2009,Pham2005} of Bilateral filtering kernel improves the hardware dependent aspects of Bilateral filtering kernel and hence expend resources on achieving performance gains that can not match the effort taken.
-Here, we address a root problem of the standard naive algorithm by creating a new algorithm, and then we also develop hardware dependent improvements for greater speedup.
-Below is the pseudo-code for the standard naive stencil kernel.
+Previous implementations\cite{Paris2009,Pham2005} of Bilateral filtering kernel improves the hardware dependent aspects and hence expend resources on achieving performance gains that can not match the effort taken.
+Here, we address a root problem of the standard naive algorithm by proposing a new algorithm, and then develop hardware dependent improvements for further speedup.
+Algorithm~\ref{alg:Example-Algorithm-float} is the pseudo-code for the standard naive Bilateral filtering kernel.
+
+%\begin{algorithm}[!h]
+%\caption{Generalized Naive Stencil Kernel Algorithm}
+%\label{alg:Example-Algorithm-float}
+%\begin{algorithmic}
+%\FORALL{element  $s_{i,j}$ in source image s}
+%\FORALL{data dependency $n$}
+%\STATE{$X_n \leftarrow 0$}
+%\ENDFOR
+%\FOR{$k=-filter\_hw$ \TO $filter\_hw$}
+%\FOR {$l  = -filter\_hw$ \TO $filter\_hw$}
+%\FORALL{data dependency $n$}
+%\STATE{$X_n \leftarrow X_n + f_n(s[i,j],s[i+k,j+l])$}
+%\ENDFOR
+%\ENDFOR
+%\ENDFOR
+%\STATE{output[$i,j] \leftarrow f_{output}(X_{1}, \dots, X_{last})$}
+%\ENDFOR
+%\end{algorithmic}
+%\end{algorithm}
 
 \begin{algorithm}[!h]
-\caption{\label{alg:Example-Algorithm-float}Naive Stencil Kernel Algorithm}
+\caption{Naive Bilateral Filter Algorithm}
+\label{alg:Example-Algorithm-float}
 \begin{algorithmic}
-\FORALL{element  $e_{i,j}$ in source image  s}
-\STATE{$C_{i,j}$ = $e_{i,j}$}, /* C is the center pixel */
-\FOR{$k=-filter\_hw \to filter\_hw$}
-\FOR {$l  = -filter\_hw \to filter\_hw$}
-\FORALL{data dependency $n$}
-\STATE{$X_n$ += $f_n(s[C_{i,j}],s[N_{i+k,j+l}])$}
-\STATE /* N is the neighboring pixel */
+\FORALL{element  $s_{i,j}$ in source image s}
+\STATE{$X_{1} \leftarrow 0$}
+\STATE{$X_{2} \leftarrow 0$}
+\FOR{$k=-filter\_hw$ \TO $filter\_hw$}
+\FOR {$l  = -filter\_hw$ \TO $filter\_hw$}
+\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i,j+l])^{2}}$}
+\STATE{$X_1 \leftarrow X_1 + temp \cdot s[i+k,j+l]$}
+\STATE{$X_2 \leftarrow X_2 + temp$}
 \ENDFOR
 \ENDFOR
+\STATE{output[$i,j] \leftarrow X_{1} / X_{2}$}
 \ENDFOR
-\ENDFOR
-\STATE $output[C_{i,j}] = f_{output}(X)$
 \end{algorithmic}
@@
 
-\subsection{Pair-symmetric Bilateral kernel Processing}
-The pair-symmetric algorithm reduces the amount of Center-Neighbor computations by half and thus can theoretically cut processing time in half at the cost of increased memory requirement which increases with the amount of dependencies.
-Its implementation is non-trivial due to potential race-conditions if not done carefully.
+\subsection{Pair-symmetric Bilateral filtering kernel}
+The pair-symmetric algorithm, shown in Algorithm~\ref{alg:Example-Algorithm-float-1}, reduces the amount of Center-Neighbor computations by half.
+Thus, it can theoretically cut the processing time in half at the cost of increased memory requirement, which increases with the amount of dependencies.
+Its implementation is non-trivial due to potential race-conditions.
 
-Our pair-symmetric Bilateral filtering algorithm is premised on the fact that each pixel assumes the role of both a Center Pixel and a Neighboring Pixel and that the codes are symmetric with respect to Center-Neighbor paired computations.
+Our pair-symmetric Bilateral filtering algorithm is premised on the fact that each pixel assumes the role of both a Center Pixel and a Neighboring Pixel, and that the codes are symmetric with respect to Center-Neighbor paired computations.
 For example, in the Bilateral filter, $c(\xi,x)$ returns the same value as $c(x,\xi)$ and $s(f(\xi),f(x))$ returns the same value as $s(f(x),f(\xi))$.
@@
 
-Pair-symmetric algorithm eliminates this redundancy that is inherent in the naive Bilateral filtering algorithm.
-Instead of reading from all Neighbors, pair-symmetric algorithm reads from only half of the Neighbors.
+The Pair-symmetric algorithm eliminates this redundancy that is inherent in the naive Bilateral filtering algorithm.
+Instead of reading from all Neighbors, our pair-symmetric algorithm reads from only half of the Neighbors.
 The result of each Center-Neighbor paired computation is then added to the $X_n$ register, which eventually obtains the result of data dependency $n$.
 This is trivial as the naive Bilateral filtering algorithm already does this.
-What pair-symmetric algorithm does unique is that the result of each Center-Neighbor pair computation is also added to and stored in the Neighbor's position in a temporary array (``Map'') that is used for storing the results of a data dependency.
+What our pair-symmetric algorithm does unique is that the result of each Center-Neighbor paired computation is also added to and stored in the Neighbor's position in a temporary array ("Map'') that is used for storing the results of a data dependency.
 So, there are $n$ Maps for $n$ number of dependencies.
@@
 After the numerators and denominators are derived, a function iterates and divides the corresponding elements of the two arrays to return the output.
-The pseudocode is given below:
+
+%\begin{algorithm}[h]
+%\caption{Pair-symmetric Stencil Algorithm}
+%\label{alg:Example-Algorithm-float-1}
+%\begin{algorithmic}
+%\STATE /* D is the set of dependency arrays */
+%\FORALL{data dependency $n$}
+%\FORALL{element $s_{i,j}$ in source image s}
+%\STATE{$D_n[i,j] \leftarrow 0$}
+%\ENDFOR
+%\ENDFOR
+%\FORALL{element $s_{i,j}$ in source image $s$}
+%\FOR{$k=1$ \TO $filter\_hw$}
+%\FOR{$l = -filter\_hw$ \TO $filter\_hw$}
+%\FORALL{data dependency $n$}
+%\STATE $temp \leftarrow f_n(s[i,j],s[i+k,j+l])$
+%\STATE $D_n[i,j] \leftarrow D_n[i,j] + temp$
+%\STATE $D_n[i+k,j+l] \leftarrow D_n[i+k,j+l] + temp$
+%\ENDFOR
+%\ENDFOR
+%\ENDFOR
+%\FOR{$l = -filter\_hw$ \TO $0$}
+%\FORALL{data dependency $n$}
+%\STATE $temp \leftarrow f_n(s[i,j],s[i,j+l])$
+%\STATE $D_n[i,j] \leftarrow D_n[i,j] + temp$
+%\STATE $D_n[i,j+l] \leftarrow D_n[i,j+l] + temp$
+%\ENDFOR
+%\ENDFOR
+%\ENDFOR
+%\FORALL{element $s_{i,j}$ in source image s}
+%\STATE{$output[i,j] \leftarrow f_{output}(D_{1}[i,j], \dots, D_{last}[i,j])$}
+%\ENDFOR
+%\end{algorithmic}
+%\end{algorithm}
+
+%\begin{algorithm}[h]
+%\caption{Pair-symmetric Bilateral Filtering kernel}
+%\label{alg:Example-Algorithm-float-1}
+%\begin{algorithmic}
+%\STATE /* D is the set of dependency arrays */
+%\FORALL{element $s_{i,j}$ in source image s}
+%\STATE{$D_1[i,j] \leftarrow 0$}
+%\STATE{$D_2[i,j] \leftarrow 0$}
+%\ENDFOR
+%\FORALL{element $s_{i,j}$ in source image $s$}
+%\FOR{$k=1$ \TO $filter\_hw$}
+%\FOR{$l = -filter\_hw$ \TO $filter\_hw$}
+%\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i+k,j+l])^{2}}$}
+%\STATE{$D_1[i,j] \leftarrow D_1[i,j] + temp \cdot s[i+k,j+l]$}
+%\STATE{$D_2[i,j] \leftarrow D_2[i,j] + temp$}
+%\STATE{$D_1[i+k,j+l] \leftarrow D_1[i+k,j+l] + temp \cdot s[i,j]$}
+%\STATE{$D_2[i+k,j+l] \leftarrow D_2[i+k,j+l] + temp$}
+%\ENDFOR
+%\ENDFOR
+%\FOR{$l = -filter\_hw$ \TO $0$}
+%\STATE{$temp \leftarrow e^{C_{1}l^2 + C_2(s[i,j] - s[i,j+l])^{2}}$}
+%\STATE{$D_1[i,j] \leftarrow D_1[i,j] + temp \cdot s[i,j+l]$}
+%\STATE{$D_2[i,j] \leftarrow D_2[i,j] + temp$}
+%\STATE{$D_1[i,j+l] \leftarrow D_1[i,j+l] + temp \cdot s[i,j]$}
+%\STATE{$D_2[i,j+l] \leftarrow D_2[i,j+l] + temp$}
+%\ENDFOR
+%\ENDFOR
+%\FORALL{element $s_{i,j}$ in source image s}
+%\STATE{$output[i,j] \leftarrow D_{1}[i,j] / D_{2}[i,j]$}
+%\ENDFOR
+%\end{algorithmic}
+%\end{algorithm}
 
 \begin{algorithm}[h]
-\caption{\label{alg:Example-Algorithm-float-1}Pair-symmetric Bilateral Filtering Algorithm}
+\caption{Pair-symmetric Bilateral Filtering Algorithm}
+\label{alg:Example-Algorithm-float-1}
 \begin{algorithmic}
-\FORALL{element $e_{i,j}$ in source image $s$}
-\STATE $C_{i,j}$ = $e_{i,j}$, /* C is the center pixel */
-\STATE{$X_n$= 0}
-\FOR{$k=0 \to filter\_hw$}
-\FOR{$l = 0 \to filter\_hw$}
-\FORALL{data dependency $n$}
-\STATE $temp$ = $f_n$ ($s$[$C_{i,j}$],$s$[$N_{i+k,j+l}$])
-\STATE /* N is the neighboring pixel */
-\STATE $X_n$ += $temp$
-\STATE $D_n$[$N_{k,l}$] += $temp$
-\STATE /* D is the dependency array */
-\FOR{$k =-1 \to -filter\_hw$}
-\FOR{$l = -1 \to -filter\_hw$}
-\STATE $temp$ = $f_n(s[C_{i,j}],s[N_{i+k,j+l}]$
-\STATE $X_n$+= $temp$
-\STATE $D_n$[$N_{k,l}$] += $temp$
-\STATE $D_n$[$C_{i,j}$] += $X_n$
-\FORALL{element $e_{i,j}$ of output image}
-\STATE{output[$e_{i,j}$]=$f_{output}(D)$}
+\STATE /* D is the set of dependency arrays */
+\FORALL{element $s_{i,j}$ in source image s}
+\STATE{$D_1[i,j] \leftarrow 0$}
+\STATE{$D_2[i,j] \leftarrow 0$}
 \ENDFOR
+\FORALL{element $s_{i,j}$ in source image $s$}
+\STATE{$X_1 \leftarrow 0$}
+\STATE{$X_2 \leftarrow 0$}
+\FOR{$k=1$ \TO $filter\_hw$}
+\FOR{$l = -filter\_hw$ \TO $filter\_hw$}
+\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i+k,j+l])^{2}}$}
+\STATE{$X_1 \leftarrow X_1 + temp \cdot s[i+k,j+l]$}
+\STATE{$X_2 \leftarrow X_2 + temp$}
+\STATE{$D_1[i+k,j+l] \leftarrow D_1[i+k,j+l] + temp \cdot s[i,j]$}
+\STATE{$D_2[i+k,j+l] \leftarrow D_2[i+k,j+l] + temp$}
 \ENDFOR
+\STATE{syncthreads();}
 \ENDFOR
+\FOR{$l = -filter\_hw$ \TO $0$}
+\STATE{$temp \leftarrow e^{C_{1} l^2 + C_2(s[i,j] - s[i,j+l])^{2}}$}
+\STATE{$X_1 \leftarrow X_1 + temp \cdot s[i,j+l]$}
+\STATE{$X_2 \leftarrow X_2 + temp$}
+\STATE{$D_1[i,j+l] \leftarrow D_1[i,j+l] + temp \cdot s[i,j]$}
+\STATE{$D_2[i,j+l] \leftarrow D_2[i,j+l] + temp$}
 \ENDFOR
+\STATE $D_1[i,j] \leftarrow D_1[i,j] + X_1$
+\STATE $D_2[i,j] \leftarrow D_2[i,j] + X_2$
 \ENDFOR
-\ENDFOR
+\FORALL{element $s_{i,j}$ in source image s}
+\STATE{$output[i,j] \leftarrow D_{1}[i,j] / D_{2}[i,j]$}
 \ENDFOR
@@
 
-CUDA's intra-warp implicit synchronization capability is the cornerstone of our CUDA implementation of pair-symmetric algorithm.
-Without implicit synchronization, pair-symmetric algorithm would suffer significantly over CUDA because it would require expensive synchronization calls per Center-Neighbor Filter computation.
+CUDA's intra-warp implicit synchronization capability is the cornerstone of our CUDA implementation of our pair-symmetric algorithm.
+Without implicit synchronization, our pair-symmetric algorithm would suffer significantly over CUDA because it would require expensive synchronization calls per Center-Neighbor Filter computation.
 
@@
 This is so because all the threads in the same warp share only one program counter.
-So, a divergent branch will cause some or most of the threads to be idle as the program counter runs through the branch and drives only those threads that meet the branch condition to work.
-An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)\{...\} else\{...\})$.
+A divergent branch will cause some or most of the threads to be idle as the program counter runs through the branch and drives only those threads that meet the branch condition to work.
+An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.
 
-As it relates to stencil codes in general, divergent branching will commonly occur in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts.
-For example, if eight threads were to read from eight non-contiguous segments of global memory--such as that which is commonly seen in strided access patterns like array{[}threadIdx.x {*} 2{]}--divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined.
+Divergent branching will commonly occur in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts.
+For example, if eight threads were to read from eight non-contiguous segments of global memory - commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]} - divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined.
 
@@
 Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory.
-What this means in the implementation of Bilateral filtering algorithm is that warps should correspond to horizontal rows of pixels for an i{*}jMax+j representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp, and if, for example, there were two rows such that half of the threads were reading from memory on each row, then there would be two transactions instead of one which is inefficient.
+For Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*jMax + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.
 
 While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable.
-As a consequence, work must be arranged such that warps can work as independent from each other as possible in order to minimize explicit synchronization calls as shown in Figure~\ref{fig:pspro}.
-This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the instant row of input data has completed.
+As a consequence, work must be arranged such that warps can work as independent from each other as possible, in order to minimize explicit synchronization calls as shown in Figure~\ref{fig:pspro}.
+This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the current row of input data has completed.
 This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call will result in a warp trespassing into the territory of another warp, thereby causing race conditions as threads from two different warps will consequently be prone to write to the same memory location.
@@
 \subsection{Shared Memory and Tile Processing}
+\begin{figure}
+\includegraphics[width = 0.48 \columnwidth]{images/project24.pdf}
+\caption{Illustration of our CUDA implementation of pair-symmetric processing using implicit intra-wap synchronization for iterating through the pixel-cells for each thread-block's assigned tile.}
+\label{fig:pspro}
+\end{figure}
 
 \begin{figure}
-\includegraphics[width = 1 \columnwidth]{images/project22}
-\caption{Illustration of CUDA implementation of pair-symmetric processing}
+\includegraphics[width = 0.48  \columnwidth]{images/tiles19.pdf}
+\caption{Illustration of our CUDA implemented pair-symmetric algorithm processing a 24 x 12 pixel image using 2 x 2 thread-blocks each with 4 x 3 threads and with the half-width parameter set to 3 pixels.}
+\label{fig:pspro}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width = 0.48  \columnwidth]{images/smallfaces.pdf}
+\caption{An illustrated overview of the bilateral filtering process.
+The six raster images of processed tiles reflect the successive work of the four concurrent thread-blocks iteratively carried out over the padded source-image.}
 \label{fig:pspro}
@@
 
-If processes were assigned to adjacent tiles, race conditions would occur because the edge cells of the tiles, a by-product of the Filter's half-width, would overlap each other, as illustrated in Figure~\ref{fig:Tiling-for-Pair-Symmetric}.
+If processes were assigned to adjacent tiles, race conditions would occur because the edge cells of the tiles would overlap each other due to intersecting half-width periphery regions, as illustrated in Figure~\ref{fig:Tiling-for-Pair-Symmetric}.
 To get around this problem, simultaneously processed tiles can be spaced apart in strides such that no two tiles being processed at the same time intersect with one another.
@@
 \subsection{Performance of Bilateral filtering kernel on GPU}
-We have performed thorough experiments on the GPU architecture with varying number of blocks and threads for both naive Bilateral filtering algorithm and pair-symmetric Bilateral filtering algorithm.
+
+We used the NVIDIA GTX 280 for benchmarking the performance of our CUDA implementations.
+Our implementations' variable parameters were the image size, the filter window size, and the grid dimensions (\# blocks horizontal * \# of blocks vertical concurrently executed) and block dimensions (\# of threads horizontal * \# of threads vertical in each block).
+The tests used two images: a large image with 3000 x 3000 pixels and a small image with 768 x 512 pixels.
+
+On the GPU, our pair-symmetric algorithm outperformed our naive algorithm for every set of tests.
+For example, Figure~\ref{fig:hw10} shows the pair-symmetric algorithm's best timing to be roughly 75\% of the naive algorithm's best timing for processing a 3000 x 3000 image using a 21 x 21 pixel filter window.
+The test that gave the closest difference in timing as shown in Figure~\ref{fig:hw10light} was processing a 768 x 712 pixel image using a 11 x 11 window size.
+In this test, the pair-symmetric algorithm's best timing was at a virtual tie with the naive algorithm's best timing at 0.00390 seconds, but upon further testing we found the pair-symmetric algorithm's optimal timing to be 0.00388 seconds when changing the grid dimension to 16 x 16 concurrent blocks, thus slightly outperforming the naive algorithm.
+Figures \ref{fig:hw10} and \ref{fig:hw10light} show exponential rises in execution time for both algorithms as block vertical dimension decreased.
+Due to runtime complexity of the bilateral filter being the square of the filter width, Figures (b) and (d), which reflect tests that used a 5 pixel filter width, do not exhibit the steep curvature that Figures (a) and (c) exhibit.
+
+While our tests returned a maximum 25\% reduction in execution time, our pair-symmetric algorithm has the potential to reach a near 50\% reduction in execution time.
+Furthermore, the benchmarks of our pair-symmetric algorithm reveal minor irregularities in the trajectory of performance derived from execution timings versus respective incremental adjustments of parameter values such as CUDA block and grid dimensions.
+The main cause of both of these observations is shared memory.
+To better understand the impact of the additional shared memory transactions that our pair-symmetric algorithm requires, whereas the naive algorithm does not require shared memory for output storage, we implemented a single-threaded CPU-based version of both the naive bilateral filter algorithm and pair-symmetric algorithm so that native-memory (RAM, L1, or L2 cache) could substitute for the GPU's shared memory.
+The results were remarkably illustrative of our pair-symmetric algorithm's potential.
+The naive algorithm running on the CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 1206.7 seconds while our pair-symmetric algorithm took 611.8 seconds - a 49.23\% reduction in execution time.
+Moreover, when replacing shared memory transactions with dummy register transactions in our CUDA implementation of the pair-symmetric algorithm, we observed a consistent 40\% reduction in execution time.
+From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., achieving timings no less than 75\% of the naive algorithm's timings.
+To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code - CUDA's quasi-assembly language - of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions.
+Not only were no optimizations being made but the PTX code showed that the shared memory transactions required additional load and store instructions, further evidencing the cost of using shared memory.
+
+Another challenge in interpreting our results was porting the concept of "speed-up" to general purpose GPU programming.
+Since speed-up is necessarily with respect to a reference timing, we used the same reference timing that our CPU optimizations used---that is how we calculated the 235.5x speed-up figure stated in the abstract of this paper---and we also made another reference timing by running our GPU implementations using a single thread.
+For example, using a single thread operating with filter width of 10 pixels to process a 768 x 512 pixel image, the time taken for the GTX280 to complete this execution was an average of 177.9 seconds for both of our algorithms; using a filter width of 5 pixels, the average execution time was 50.5 seconds for both algorithms.
+With respect to these reference timings, the speed-up is roughly 20,500x for the 10 pixel filter width and 12,440x for the 5 pixel filter width.
 
@@
 \end{figure}
-
-Figure~\ref{fig:gpuresults} represents the performance of both of the algorithms.
 

### 2012-03-10 trunk/LateX/paper8.tex
changed using smallfaces.pdf to smallfaces5.pdf
modified the first part about cue and intra-warp syncing

@@
-test
-
 \documentclass{IEEEtran}
@@
 
-In CUDA, threads are organized into groups called warps.
+In CUDA, threads are organized into groups called warps, and warps are launched in groups called thread-blocks, in which each thread-block can share among its warps an allocated portion of shared-memory.
 Each warp executes one instruction at a time for its own group of threads.
@@
 This is so because all the threads in the same warp share only one program counter.
-A divergent branch will cause some or most of the threads to be idle as the program counter runs through the branch and drives only those threads that meet the branch condition to work.
+A divergent branch would cause some or most of the threads to be idle as the program counter runs through the branch and drives only those threads that meet the branch condition to work.
 An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.
 
-Divergent branching will commonly occur in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts.
-For example, if eight threads were to read from eight non-contiguous segments of global memory - commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]} - divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined.
-
+Divergent branching commonly occur in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts.
+For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined.
 In other words, a memory transaction will be serialized unless all the threads in the warp to which the transaction belongs are aligned consecutively and fit within a segment of memory.
@@
 \begin{figure}
-\includegraphics[width = 0.48  \columnwidth]{images/smallfaces.pdf}
+\includegraphics[width = 0.48  \columnwidth]{images/smallfaces5.pdf}
 \caption{An illustrated overview of the bilateral filtering process.

### 2012-03-10 LateX/paper8.tex
test
editing wording of the paragraph about divergent branching. e.g. changed will to would etc

--- 2012-03-10 trunk/LateX/paper8.tex
@@
 CUDA's intra-warp implicit synchronization capability is the cornerstone of our CUDA implementation of our pair-symmetric algorithm.
-Without implicit synchronization, our pair-symmetric algorithm would suffer significantly over CUDA because it would require expensive synchronization calls per Center-Neighbor Filter computation.
+Without implicit synchronization, the performance of our CUDA-implemented pair-symmetric algorithm would be significantly reduced because it would require expensive synchronization calls per Center-Neighbor Filter computation.
 
-In CUDA, threads are organized into groups called warps.
+In CUDA, threads are organized into groups called warps, and warps are launched in groups called thread-blocks, in which each thread-block can share among its warps an allocated portion of shared-memory.
 Each warp executes one instruction at a time for its own group of threads.
 Because all threads in a warp share the same program counter and are therefore driven by the same sequential instruction stream, all threads in a warp are always in sync with each other without requiring explicit synchronization calls.
-Therefore, synchronization between threads in a warp is implicit.
+Hence, synchronization between threads in a warp is implicit.
 
-To take full advantage of CUDA's implicit intra-warp synchronization feature, divergent branching between threads in the same warp must be avoided.
-This is so because all the threads in the same warp share only one program counter.
-A divergent branch will cause some or most of the threads to be idle as the program counter runs through the branch and drives only those threads that meet the branch condition to work.
+To take full advantage of CUDA's implicit intra-warp synchronization, divergent branching between threads in the same warp must be avoided.
+This is because all the threads in the same warp share only one program counter.
+A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that meet the branch condition to work.
 An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.
 
-Divergent branching will commonly occur in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts.
-For example, if eight threads were to read from eight non-contiguous segments of global memory - commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]} - divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined.
-
-In other words, a memory transaction will be serialized unless all the threads in the warp to which the transaction belongs are aligned consecutively and fit within a segment of memory.
+Divergent branching commonly occurs in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts.
+For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined.
+In other words, a memory transaction would be serialized if all the threads in the warp to which the transaction belonged were not aligned consecutively or did not fit within a segment of memory.
 Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory.
-For Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*jMax + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.
+For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*j_pitch + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.
 

### 2012-03-10 trunk/LateX/paper8.tex
more minor changes

@@
 This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter.
-It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images.
-As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e.
-one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration.
+It should be noted, however, that the foregoing equations are mathematical representations for filtering infinite-resolution images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image.
+Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration.
 A formal expression is given below:
@@
 Consequently, the Bilateral filter has two data dependencies.
-As explained later in this paper, the pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.
+As explained later in this paper, our pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.
 
@@
 CUDA's intra-warp implicit synchronization capability is the cornerstone of our CUDA implementation of our pair-symmetric algorithm.
-Without implicit synchronization, our pair-symmetric algorithm would suffer significantly over CUDA because it would require expensive synchronization calls per Center-Neighbor Filter computation.
+Without implicit synchronization, the performance of our CUDA-implemented pair-symmetric algorithm would be significantly reduced because it would require expensive synchronization calls per Center-Neighbor Filter computation.
 
@@
 
-To take full advantage of CUDA's implicit intra-warp synchronization feature, divergent branching between threads in the same warp must be avoided.
+To take full advantage of CUDA's implicit intra-warp synchronization, divergent branching between threads in the same warp must be avoided.
 This is so because all the threads in the same warp share only one program counter.
-A divergent branch would cause some or most of the threads to be idle as the program counter runs through the branch and drives only those threads that meet the branch condition to work.
+A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that met the branch condition to work.
 An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.
 
-Divergent branching commonly occur in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts.
+Divergent branching commonly occurs in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts.
 For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined.
-In other words, a memory transaction will be serialized unless all the threads in the warp to which the transaction belongs are aligned consecutively and fit within a segment of memory.
+In other words, a memory transaction would be serialized if all the threads in the warp to which the transaction belonged were not aligned consecutively or did not fit within a segment of memory.
 Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory.
-For Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*jMax + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.
+For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*width + j] representation of a 2D image, for if threads in a warp were to read from or write to array elements (representing pixels) in a vertical order, there would be a separate transaction for each thread in the warp.
 
 While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable.
-As a consequence, work must be arranged such that warps can work as independent from each other as possible, in order to minimize explicit synchronization calls as shown in Figure~\ref{fig:pspro}.
-This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the current row of input data has completed.
-This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call will result in a warp trespassing into the territory of another warp, thereby causing race conditions as threads from two different warps will consequently be prone to write to the same memory location.
+As a consequence, work should be arranged such that warps can work as independently from each other as possible in order to minimize explicit synchronization calls, as illustrated in Figure~\ref{fig:pspro}.
+This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work of each warp on its current row of input data has completed.
+This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call would result in warps trespassing into the territory of other warps, thereby causing race conditions as threads from different warps would be consequently prone to writing to the same memory location.
 

### 2012-03-10 trunk/LateX/paper8.tex
more minor changes

@@
 It should be noted, however, that the foregoing equations are mathematical representations for filtering infinite-resolution images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image.
-Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration.
+Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set of calculations per Center Pixel.
 A formal expression is given below:

### 2012-03-11 trunk/LateX/paper8.tex
more minor changes

@@
 It should be noted, however, that the foregoing equations are mathematical representations for filtering infinite-resolution images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image.
-Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set of calculations per Center Pixel.
+Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set neighbors per Center Pixel.
 A formal expression is given below:
@@
 Figures \ref{fig:hw10} and \ref{fig:hw10light} show exponential rises in execution time for both algorithms as block vertical dimension decreased.
-Due to runtime complexity of the bilateral filter being the square of the filter width, Figures (b) and (d), which reflect tests that used a 5 pixel filter width, do not exhibit the steep curvature that Figures (a) and (c) exhibit.
+Due to runtime complexity of the bilateral filter being the square of the filter width, Figures (b) and (d), which reflect our tests that used a 5 pixel filter width, do not exhibit the steep curvature that Figures (a) and (c) exhibit.
 
@@
 Moreover, when replacing shared memory transactions with dummy register transactions in our CUDA implementation of the pair-symmetric algorithm, we observed a consistent 40\% reduction in execution time.
-From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., achieving timings no less than 75\% of the naive algorithm's timings.
-To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code - CUDA's quasi-assembly language - of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions.
+From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., reducing timings by no more than than 25\%.
+To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code, CUDA's quasi-assembly language, of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions.
 Not only were no optimizations being made but the PTX code showed that the shared memory transactions required additional load and store instructions, further evidencing the cost of using shared memory.

### 2012-03-11 trunk/LateX/paper8.tex
changed manycore to many-core.

@@
 \begin{document}
-\title{Acceleration of Bilateral filtering algorithm for manycore and multicore architectures}
+\title{Accelerated Bilateral Filtering for Many-core and Multi-core Architectures}
 \author{\IEEEauthorblockN{Dinesh Agarwal, Sami Wilf, Abinashi Dhungel, and Sushil K. Prasad}\\
@@
 \begin{abstract}
-This work explores multicore and manycore acceleration for the embarrassingly parallel, compute-intensive Bilateral filtering kernel.
-For manycore architectures, we have created a pair-symmetric algorithm to avoid redundant calculations and for multicore architectures we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads.
+This work explores multicore and many-core acceleration for the embarrassingly parallel, compute-intensive Bilateral filtering kernel.
+For many-core architectures, we have created a pair-symmetric algorithm to avoid redundant calculations and for multicore architectures we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads.
 We present empirical data evidencing the performance gains we achieved over variety of hardware architectures including Nvidia GTX280, AMD Barcelona, AMD Shanghai, Intel Harpertown, AMD Phenom, Intel Core i7 quad core, and Intel Nehalem 32 core machines.
@@
 \begin{figure}
-\includegraphics[width = 0.48 \columnwidth]{images/project24.pdf}
+\includegraphics[width = 1\columnwidth]{images/project23_.pdf}
 \caption{Illustration of our CUDA implementation of pair-symmetric processing using implicit intra-wap synchronization for iterating through the pixel-cells for each thread-block's assigned tile.}
@@
 \begin{figure}
-\includegraphics[width = 0.48  \columnwidth]{images/tiles19.pdf}
+\includegraphics[width = 1\columnwidth]{images/tiles19.pdf}
 \caption{Illustration of our CUDA implemented pair-symmetric algorithm processing a 24 x 12 pixel image using 2 x 2 thread-blocks each with 4 x 3 threads and with the half-width parameter set to 3 pixels.}
@@
 \begin{figure}
-\includegraphics[width = 0.48  \columnwidth]{images/smallfaces5.pdf}
+\includegraphics[width = 1\columnwidth]{images/smallfaces5.pdf}
 \caption{An illustrated overview of the bilateral filtering process.
@@
 
-\begin{figure}
-\includegraphics[width=0.48\textwidth]{images/stackedtiles}
-\caption{Tiling for Pair-Symmetric Stencil Processing with each tile having a unique color and overlapping tiles have a resultant color that results from mixing colors of overlapping tiles.
+%\begin{figure}
+%\includegraphics[width=0.48\textwidth]{images/stackedtiles}
+%\caption{Tiling for Pair-Symmetric Stencil Processing with each tile having a unique color and overlapping tiles have a resultant color that results from mixing colors of overlapping tiles.
 The pair-symmetric thread access pattern ensures exclusive memory writes.}
-\label{fig:Tiling-for-Pair-Symmetric}
-\end{figure}
+%\label{fig:Tiling-for-Pair-Symmetric}
+%\end{figure}
 
-\begin{figure}
-\includegraphics[width=0.48\textwidth]{images/lenatiles}
-\caption{Input image after one iteration of pair-symmetric algorithm. Tiles do not have perfect boundaries due to overlaps.}
-\label{fig:An-iteration-of}
-\end{figure}
+%\begin{figure}
+%\includegraphics[width=0.48\textwidth]{images/lenatiles}
+%\caption{Input image after one iteration of pair-symmetric algorithm. Tiles do not have perfect boundaries due to overlaps.}
+%\label{fig:An-iteration-of}
+%\end{figure}
 
@@
 For example, using a single thread operating with filter width of 10 pixels to process a 768 x 512 pixel image, the time taken for the GTX280 to complete this execution was an average of 177.9 seconds for both of our algorithms; using a filter width of 5 pixels, the average execution time was 50.5 seconds for both algorithms.
-With respect to these reference timings, the speed-up is roughly 20,500x for the 10 pixel filter width and 12,440x for the 5 pixel filter width.
+With respect to these reference timings, the speed-up is roughly 21,090x for the 10 pixel filter width and 12,440x for the 5 pixel filter width.
 

### 2012-03-11 LateX/paper8.tex
changed manycore to many-core. changing title to be more specific.

@@
 \begin{document}
-\title{Acceleration of Bilateral filtering algorithm for manycore and multicore architectures}
+\title{Accelerated Bilateral Filtering for Multi-core Architectures and Code Unified Device Architecture}
 \author{\IEEEauthorblockN{Dinesh Agarwal, Sami Wilf, Abinashi Dhungel, and Sushil K. Prasad}\\
@@
 This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter.
-It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images.
-As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e.
-one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration.
+It should be noted, however, that the foregoing equations are mathematical representations for filtering infinite-resolution images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image.
+Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set neighbors per Center Pixel.
 A formal expression is given below:
@@
 Because all threads in a warp share the same program counter and are therefore driven by the same sequential instruction stream, all threads in a warp are always in sync with each other without requiring explicit synchronization calls.
-Hence, synchronization between threads in a warp is implicit.
+Therefore, synchronization between threads in a warp is implicit.
 
 To take full advantage of CUDA's implicit intra-warp synchronization, divergent branching between threads in the same warp must be avoided.
-This is because all the threads in the same warp share only one program counter.
-A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that meet the branch condition to work.
+This is so because all the threads in the same warp share only one program counter.
+A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that met the branch condition to work.
 An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.
@@
 Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory.
-For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*j_pitch + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.
+For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*width + j] representation of a 2D image, for if threads in a warp were to read from or write to array elements (representing pixels) in a vertical order, there would be a separate transaction for each thread in the warp.
 
 While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable.
-As a consequence, work must be arranged such that warps can work as independent from each other as possible, in order to minimize explicit synchronization calls as shown in Figure~\ref{fig:pspro}.
-This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the current row of input data has completed.
-This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call will result in a warp trespassing into the territory of another warp, thereby causing race conditions as threads from two different warps will consequently be prone to write to the same memory location.
+As a consequence, work should be arranged such that warps can work as independently from each other as possible in order to minimize explicit synchronization calls, as illustrated in Figure~\ref{fig:pspro}.
+This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work of each warp on its current row of input data has completed.
+This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call would result in warps trespassing into the territory of other warps, thereby causing race conditions as threads from different warps would be consequently prone to writing to the same memory location.
 

### 2012-03-11 LateX/paper8.tex
Updated text and arrangement

--- 2012-03-10 LateX/paper8.tex
@@
 \usepackage{pgfplots}
+\usepackage{wrapfig}
 \usepackage{url}
@@
 \begin{abstract}
-This work explores multicore and manycore acceleration for the embarrassingly parallel, compute-intensive Bilateral filtering kernel.
+This work explores multicore and manycore acceleration for the embarrassingly parallel, compute-intensive bilateral filtering kernel.
 For manycore architectures, we have created a pair-symmetric algorithm to avoid redundant calculations and for multicore architectures we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads.
@@
 
-Our work explores efficient parallel implementation of Bilateral filtering kernel\footnotemark[1] proposed by Tomasi and Manduchi \cite{Tomasi1998}, on multicore and manycore architectures.
-Bilateral filtering kernel is used to smooth an image without blurring its edges.
+Our work explores efficient parallel implementation of bilateral filtering kernel\footnotemark[1] proposed by Tomasi and Manduchi \cite{Tomasi1998}, on multicore and manycore architectures.
+bilateral filtering kernel is used to smooth an image without blurring its edges.
 It is ubiquitous in various image processing applications in various contexts such as denoising\cite{zhang2008}, texture editing and relighting\cite{eisemann2004}, tone management\cite{Bae2006}, stylization\cite{DeCarlo2002}, optical-flow estimations\cite{Xiao2006}, and demosaicking\cite{Ramanath2003} etc.
@@
 This reduces the complexity from $O(N*(m^{d}))$ to $O(N*d)$, where $N$ is the total number of pixels in the image and $d$ is the dimensions of the image.
-Yang et al.\cite{yang2009} have reported a constant time algorithm by dividing the Bilateral filtering kernel into a constant number of time spatial kernels.
+Yang et al.\cite{yang2009} have reported a constant time algorithm by dividing the bilateral filtering kernel into a constant number of time spatial kernels.
 However, we do not parallelize these methods since our interest is to develop a parallelization method that can be applied, in general, to any spatially invariant and inseparable filtering algorithm.
 
-In this paper, we propose a set of optimizations and a non-naive algorithm for Bilateral filtering kernel.
+In this paper, we propose a set of optimizations and a non-naive algorithm for bilateral filtering kernel.
 Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to reduce processing time to half.
@@
 
-Rest of the paper is organized as following: Section~\ref{sec:stencil} defines the Bilateral filtering kernel and introduces a naive algorithm and our pair-symmetric algorithm for Bilateral filtering.
+Rest of the paper is organized as following: Section~\ref{sec:stencil} defines the bilateral filtering kernel and introduces a naive algorithm and our pair-symmetric algorithm for bilateral filtering.
 Section~\ref{sec:GPUoptimizations} introduces the optimizations employed to devise the pair-symmetric algorithm for CUDA architectures.
-Section~\ref{sec:optimizations} introduces the optimizations for CPU architectures and establishes the rationale behind ignoring traditional memory related optimizations for Bilateral filtering kernel.
-Our hardware and software testbed for experiments is detailed in Section~\ref{sec:expt}, immediately followed by Section~\ref{sec:results} that reports the performance of Bilateral filtering kernel on both multicore and manycore architectures.
+Section~\ref{sec:optimizations} introduces the optimizations for CPU architectures and establishes the rationale behind ignoring traditional memory related optimizations for bilateral filtering kernel.
+Our hardware and software testbed for experiments is detailed in Section~\ref{sec:expt}, immediately followed by Section~\ref{sec:results} that reports the performance of bilateral filtering kernel on both multicore and manycore architectures.
 Section~\ref{sec:conclusion} concludes this paper with remarks on general applicability of our optimizations.
@@
 Hence, pictorial edges maintain sharpness while noise in the pictorial regions are filtered, resulting in more appealing images.
-Mathematically, the Bilateral filter can be defined in terms of the spatial kernel and photometric kernel as follows\cite{Tomasi1998}:\\
+Mathematically, the bilateral filter can be defined in terms of the spatial kernel and photometric kernel as follows\cite{Tomasi1998}:\\
 
@@
 
-The combined kernel, which is the Bilateral filter, is the product
+The combined kernel, which is the bilateral filter, is the product
 of spatial kernel and photometric kernel and is described as: 
@@
 Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties.
-This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter.
+This results in edge-preserved image smoothing, which is the purpose of the bilateral filter.
 It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images.
@@
 A formal expression is given below:
-
 \begin{equation}
@@
 
-In addition to restating the Bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the Bilateral filter.
+In addition to restating the bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the bilateral filter.
 Specifically, Equation~\ref{eq:depen} shows that before the filtered pixel can be calculated by dividing the numerator by the denominator, both the summation in the numerator and denominator must first be calculated.
-Consequently, the Bilateral filter has two data dependencies.
+Consequently, the bilateral filter has two data dependencies.
 As explained later in this paper, the pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.
@@
 Moreover, as the size of the Window increases, which only happens by choice of the user as the size is user defined, the number of computations per pixel grows at a rate of $O(FilterWindowWidth^{2})$ where FilterWindowWidth is the diameter of the Window.
-By reason of the foregoing, the bulk of the time spent in processing an image using the Bilateral filter is in the arithmetic-intensive computations, because the number of writes to the destination image is constant as the rate of growth of Center-Neighbor computations with respect to the Window size increase.
+By reason of the foregoing, the bulk of the time spent in processing an image using the bilateral filter is in the arithmetic-intensive computations, because the number of writes to the destination image is constant as the rate of growth of Center-Neighbor computations with respect to the Window size increase.
 
-\subsection{A Naive Approach to Bilateral filtering kernel}
+\subsection{A Naive Approach to bilateral filtering kernel}
 
@@
 \includegraphics[clip,scale=0.35]{images/the202}\includegraphics[scale=0.35]{images/the2}\includegraphics[scale=0.35]{images/the3}
-\caption{Example of a naive Bilateral filter iterations}
+\caption{Example of a naive bilateral filter iterations}
 \end{figure}
 
-The naive stencil algorithm is the standard model that is predominantly used in Bilateral filtering kernel.
-Previous implementations\cite{Paris2009,Pham2005} of Bilateral filtering kernel improves the hardware dependent aspects and hence expend resources on achieving performance gains that can not match the effort taken.
+The naive stencil algorithm is the standard model that is predominantly used in bilateral filtering kernel.
+Previous implementations\cite{Paris2009,Pham2005} of bilateral filtering kernel improves the hardware dependent aspects and hence expend resources on achieving performance gains that can not match the effort taken.
 Here, we address a root problem of the standard naive algorithm by proposing a new algorithm, and then develop hardware dependent improvements for further speedup.
-Algorithm~\ref{alg:Example-Algorithm-float} is the pseudo-code for the standard naive Bilateral filtering kernel.
-
-%\begin{algorithm}[!h]
-%\caption{Generalized Naive Stencil Kernel Algorithm}
-%\label{alg:Example-Algorithm-float}
-%\begin{algorithmic}
-%\FORALL{element  $s_{i,j}$ in source image s}
-%\FORALL{data dependency $n$}
-%\STATE{$X_n \leftarrow 0$}
-%\ENDFOR
-%\FOR{$k=-filter\_hw$ \TO $filter\_hw$}
-%\FOR {$l  = -filter\_hw$ \TO $filter\_hw$}
-%\FORALL{data dependency $n$}
-%\STATE{$X_n \leftarrow X_n + f_n(s[i,j],s[i+k,j+l])$}
-%\ENDFOR
-%\ENDFOR
-%\ENDFOR
-%\STATE{output[$i,j] \leftarrow f_{output}(X_{1}, \dots, X_{last})$}
-%\ENDFOR
-%\end{algorithmic}
-%\end{algorithm}
+Algorithm~\ref{alg:Example-Algorithm-float} is the pseudo-code for the standard naive bilateral filtering kernel.
 
 \begin{algorithm}[!h]
-\caption{Naive Bilateral Filter Algorithm}
+\caption{Naive bilateral Filter Algorithm}
 \label{alg:Example-Algorithm-float}
@@
 
-\subsection{Pair-symmetric Bilateral filtering kernel}
+\subsection{Pair-symmetric bilateral filtering kernel}
 The pair-symmetric algorithm, shown in Algorithm~\ref{alg:Example-Algorithm-float-1}, reduces the amount of Center-Neighbor computations by half.
@@
 
-Our pair-symmetric Bilateral filtering algorithm is premised on the fact that each pixel assumes the role of both a Center Pixel and a Neighboring Pixel, and that the codes are symmetric with respect to Center-Neighbor paired computations.
-For example, in the Bilateral filter, $c(\xi,x)$ returns the same value as $c(x,\xi)$ and $s(f(\xi),f(x))$ returns the same value as $s(f(x),f(\xi))$.
-It then follows that the naive Bilateral filtering algorithm dictated in Algorithm~\ref{alg:Example-Algorithm-float} above, although simple, performs twice as many calculations than necessary.
+Our pair-symmetric bilateral filtering algorithm is premised on the fact that each pixel assumes the role of both a Center Pixel and a Neighboring Pixel, and that the codes are symmetric with respect to Center-Neighbor paired computations.
+For example, in the bilateral filter, $c(\xi,x)$ returns the same value as $c(x,\xi)$ and $s(f(\xi),f(x))$ returns the same value as $s(f(x),f(\xi))$.
+It then follows that the naive bilateral filtering algorithm dictated in Algorithm~\ref{alg:Example-Algorithm-float} above, although simple, performs twice as many calculations than necessary.
 
-The Pair-symmetric algorithm eliminates this redundancy that is inherent in the naive Bilateral filtering algorithm.
+The Pair-symmetric algorithm eliminates this redundancy that is inherent in the naive bilateral filtering algorithm.
 Instead of reading from all Neighbors, our pair-symmetric algorithm reads from only half of the Neighbors.
 The result of each Center-Neighbor paired computation is then added to the $X_n$ register, which eventually obtains the result of data dependency $n$.
-This is trivial as the naive Bilateral filtering algorithm already does this.
+This is trivial as the naive bilateral filtering algorithm already does this.
 What our pair-symmetric algorithm does unique is that the result of each Center-Neighbor paired computation is also added to and stored in the Neighbor's position in a temporary array ("Map'') that is used for storing the results of a data dependency.
@@
 After the data dependency conditions are met, the function that returns the output data from the dependencies iterates over the Maps.
-As applied to the Bilateral filter, the data dependency lies in calculating each pixel's numerator and denominator that is used to derive the pixel's output result.
+As applied to the bilateral filter, the data dependency lies in calculating each pixel's numerator and denominator that is used to derive the pixel's output result.
 After the numerators and denominators are derived, a function iterates and divides the corresponding elements of the two arrays to return the output.
-
-%\begin{algorithm}[h]
-%\caption{Pair-symmetric Stencil Algorithm}
-%\label{alg:Example-Algorithm-float-1}
-%\begin{algorithmic}
-%\STATE /* D is the set of dependency arrays */
-%\FORALL{data dependency $n$}
-%\FORALL{element $s_{i,j}$ in source image s}
-%\STATE{$D_n[i,j] \leftarrow 0$}
-%\ENDFOR
-%\ENDFOR
-%\FORALL{element $s_{i,j}$ in source image $s$}
-%\FOR{$k=1$ \TO $filter\_hw$}
-%\FOR{$l = -filter\_hw$ \TO $filter\_hw$}
-%\FORALL{data dependency $n$}
-%\STATE $temp \leftarrow f_n(s[i,j],s[i+k,j+l])$
-%\STATE $D_n[i,j] \leftarrow D_n[i,j] + temp$
-%\STATE $D_n[i+k,j+l] \leftarrow D_n[i+k,j+l] + temp$
-%\ENDFOR
-%\ENDFOR
-%\ENDFOR
-%\FOR{$l = -filter\_hw$ \TO $0$}
-%\FORALL{data dependency $n$}
-%\STATE $temp \leftarrow f_n(s[i,j],s[i,j+l])$
-%\STATE $D_n[i,j] \leftarrow D_n[i,j] + temp$
-%\STATE $D_n[i,j+l] \leftarrow D_n[i,j+l] + temp$
-%\ENDFOR
-%\ENDFOR
-%\ENDFOR
-%\FORALL{element $s_{i,j}$ in source image s}
-%\STATE{$output[i,j] \leftarrow f_{output}(D_{1}[i,j], \dots, D_{last}[i,j])$}
-%\ENDFOR
-%\end{algorithmic}
-%\end{algorithm}
-
-%\begin{algorithm}[h]
-%\caption{Pair-symmetric Bilateral Filtering kernel}
-%\label{alg:Example-Algorithm-float-1}
-%\begin{algorithmic}
-%\STATE /* D is the set of dependency arrays */
-%\FORALL{element $s_{i,j}$ in source image s}
-%\STATE{$D_1[i,j] \leftarrow 0$}
-%\STATE{$D_2[i,j] \leftarrow 0$}
-%\ENDFOR
-%\FORALL{element $s_{i,j}$ in source image $s$}
-%\FOR{$k=1$ \TO $filter\_hw$}
-%\FOR{$l = -filter\_hw$ \TO $filter\_hw$}
-%\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i+k,j+l])^{2}}$}
-%\STATE{$D_1[i,j] \leftarrow D_1[i,j] + temp \cdot s[i+k,j+l]$}
-%\STATE{$D_2[i,j] \leftarrow D_2[i,j] + temp$}
-%\STATE{$D_1[i+k,j+l] \leftarrow D_1[i+k,j+l] + temp \cdot s[i,j]$}
-%\STATE{$D_2[i+k,j+l] \leftarrow D_2[i+k,j+l] + temp$}
-%\ENDFOR
-%\ENDFOR
-%\FOR{$l = -filter\_hw$ \TO $0$}
-%\STATE{$temp \leftarrow e^{C_{1}l^2 + C_2(s[i,j] - s[i,j+l])^{2}}$}
-%\STATE{$D_1[i,j] \leftarrow D_1[i,j] + temp \cdot s[i,j+l]$}
-%\STATE{$D_2[i,j] \leftarrow D_2[i,j] + temp$}
-%\STATE{$D_1[i,j+l] \leftarrow D_1[i,j+l] + temp \cdot s[i,j]$}
-%\STATE{$D_2[i,j+l] \leftarrow D_2[i,j+l] + temp$}
-%\ENDFOR
-%\ENDFOR
-%\FORALL{element $s_{i,j}$ in source image s}
-%\STATE{$output[i,j] \leftarrow D_{1}[i,j] / D_{2}[i,j]$}
-%\ENDFOR
-%\end{algorithmic}
-%\end{algorithm}
 
@@
 \label{sec:GPUoptimizations} 
-As Bilateral filtering algorithm belongs to the class of embarrassingly parallel algorithms, CUDA is well suited for its implementation. However,
+As bilateral filtering algorithm belongs to the class of embarrassingly parallel algorithms, CUDA is well suited for its implementation. However,
 in order to effectively implement such an algorithm in CUDA, one must understand how CUDA operates, which is significantly different from any CPU.
@@
 Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory.
-For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*j_pitch + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.
+For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*width + j] representation of a 2D image, for if threads in a warp were to read from or write to array elements (representing pixels) in a vertical order, there would be a separate transaction for each thread in the warp.
+
+In other words, a memory transaction will be serialized unless all the threads in the warp to which the transaction belongs are aligned consecutively and fit within a segment of memory.
+Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory.
+For bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*jMax + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.
 
 While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable.
-As a consequence, work must be arranged such that warps can work as independent from each other as possible, in order to minimize explicit synchronization calls as shown in Figure~\ref{fig:pspro}.
+As a consequence, work must be arranged such that warps can work as independent from each other as possible, in order to minimize explicit synchronization calls.
 This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the current row of input data has completed.
@@
 
-\subsection{Shared Memory and Tile Processing}
 \begin{figure}
-\includegraphics[width = 0.48 \columnwidth]{images/project24.pdf}
-\caption{Illustration of our CUDA implementation of pair-symmetric processing using implicit intra-wap synchronization for iterating through the pixel-cells for each thread-block's assigned tile.}
-\label{fig:pspro}
+\subfigure[Each warp is assigned a row in filter window]{\includegraphics[width = 0.48 \textwidth]{images/project24.pdf}
+}\label{fig:pspro}
+\subfigure[2 x 2 thread-blocks each with 4 x 3 threads]{\includegraphics[width = 0.48  \textwidth]{images/tiles19.pdf}
+}\label{fig:pspro2}
+\subfigure[An illustrated overview of the bilateral filtering process]{\includegraphics[width = 0.48  \textwidth]{images/smallfaces.pdf}}
+\label{fig:pspro3}
+\caption{Memory access patterns for CUDA that exploit implicit intra-warp synchronization}
+\label{fig:intrawarp}
 \end{figure}
 
-\begin{figure}
-\includegraphics[width = 0.48  \columnwidth]{images/tiles19.pdf}
-\caption{Illustration of our CUDA implemented pair-symmetric algorithm processing a 24 x 12 pixel image using 2 x 2 thread-blocks each with 4 x 3 threads and with the half-width parameter set to 3 pixels.}
-\label{fig:pspro}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width = 0.48  \columnwidth]{images/smallfaces5.pdf}
-\caption{An illustrated overview of the bilateral filtering process.
-The six raster images of processed tiles reflect the successive work of the four concurrent thread-blocks iteratively carried out over the padded source-image.}
-\label{fig:pspro}
-\end{figure}
+\subsection{Shared Memory and Tile Processing}
 
@@
 Of these three types of memory, the slowest is global memory, which unfortunately is where the host must first send the input data before the GPU can begin its work.
-So, in order to see any meaningful performance in CUDA, data stored in global memory must be transferred to faster memory--i.e., shared memory, which is hundreds of times faster.
-However, the trade-off is that the shared memory size is much smaller than the global memory size, and this fact compels a tile-based approach to the implementation of Bilateral filtering algorithm.
+So, in order to see any meaningful performance in CUDA, data stored in global memory must be transferred to faster memory, i.e.
+shared memory, which is hundreds of times faster.
+However, the trade-off is that the shared memory size is much smaller than the global memory size, and this fact compels a tile-based approach to the implementation of bilateral filtering algorithm.
 
-Our tile-based processing approach is self-explanatory.
-Equal sized partitions of the input data are processed, and then any remaining data on and between the edge and the last partition that is closest to (but not falling over) the edge boundary of the 2-dimensional data representation is processed.
-With respect to the naive Bilateral filtering algorithm, tile-based processing is trivial and needs no further explanation because simply partitioning the space and processing it will suffice.
-However, for the pair-symmetric algorithm that we propose, tile-based Bilateral filtering algorithm requires adjustments in the order in which tiles are processed in order to guarantee the exclusion of race conditions.
+Figure~\ref{fig:Tiling-for-Pair-Symmetric} illustrates how tiles are processed in parallel without explicit mutual exclusion. 
+\begin{wrapfigure}{r}{0.25\textwidth}
+\centering
+\includegraphics[width=0.24\textwidth]{images/stackedtiles}
+\caption{Tiling for Pair-Symmetric bilateral filtering with each tile having a unique color. Overlapping tiles have a resultant color.}
+\vspace{0mm}
+\label{fig:Tiling-for-Pair-Symmetric}
+\end{wrapfigure}
+In tile based algorithm, equal sized partitions of the input data are processed, and then any remaining data on and between the edge and the last partition that is closest to (but not falling over) the edge boundary of the 2-dimensional data representation is processed.
+Each thread only updates (writes to) only half of the Neighbors.
+Due to this access pattern, greater number of threads can work in parallel.
+With respect to the naive bilateral filtering algorithm, tile-based processing is trivial and needs no further explanation because simply partitioning the space and processing it will suffice.
+However, for the pair-symmetric algorithm that we propose, tile-based bilateral filtering algorithm requires adjustments in the order in which tiles are processed in order to guarantee the exclusion of race conditions.
 
-\begin{figure}
-\includegraphics[width=0.48\textwidth]{images/stackedtiles}
-\caption{Tiling for Pair-Symmetric Stencil Processing with each tile having a unique color and overlapping tiles have a resultant color that results from mixing colors of overlapping tiles.
-The pair-symmetric thread access pattern ensures exclusive memory writes.}
-\label{fig:Tiling-for-Pair-Symmetric}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.48\textwidth]{images/lenatiles}
+\begin{wrapfigure}{r}{0.25\textwidth}
+\includegraphics[width=0.24\textwidth]{images/lenatiles}
 \caption{Input image after one iteration of pair-symmetric algorithm. Tiles do not have perfect boundaries due to overlaps.}
 \label{fig:An-iteration-of}
-\end{figure}
-
+\end{wrapfigure}
 If processes were assigned to adjacent tiles, race conditions would occur because the edge cells of the tiles would overlap each other due to intersecting half-width periphery regions, as illustrated in Figure~\ref{fig:Tiling-for-Pair-Symmetric}.
 To get around this problem, simultaneously processed tiles can be spaced apart in strides such that no two tiles being processed at the same time intersect with one another.
-An example of such a technique is illustrated in Figure~\ref{fig:An-iteration-of} (the tiles are flipped sideways in this rendering).
+An example of such a technique is illustrated in Figure~\ref{fig:An-iteration-of} - the tiles are flipped sideways in this rendering.
 
@@
 
-\subsection{Memory to computation ratio of Bilateral filtering kernel}
+\subsection{Memory to computation ratio of bilateral filtering kernel}
 \begin{figure}[h!]
 \begin{centering}
-\subfigure{ \includegraphics[width=0.48\textwidth]{images/noexplarge}} 
-\par\end{centering}
+\includegraphics[width=0.48\textwidth]{images/noexplarge}
+\end{centering}
 \caption{Percentage of time taken by sequential kernel when run without $exp$ instruction, $filter$ $radius$=10.}
@@
 
-Unlike traditional regular embarrassingly parallel stencil kernel codes where memory management plays a pivotal role in realizing performance, Bilateral filtering kernel is extremely computationally intensive with only a meager amount of time spent in memory accesses.
-Bilateral filtering kernel requires computation of $exponential$ function ($exp$ function) which is a major component of time spent in processing.
-Figure~\ref{fig:noexp} shows the percentage of time taken by Naive Bilateral filtering algorithm\cite{Tomasi1998} and an improved algorithm employing tiled-memory-access pattern on different platforms for a 512x512 pixels small image and a 3000x3000 pixels large image after the $exp$ function was commented out from the computation.
-The only purpose of this experiment is to justify the fact that traditional memory-access-pattern related optimizations can not improve the performance of Bilateral filtering kernel.
+Unlike traditional regular embarrassingly parallel stencil kernel codes where memory management plays a pivotal role in realizing performance, bilateral filtering kernel is extremely computationally intensive with only a meager amount of time spent in memory accesses.
+Bilateral filtering kernel requires computation of $exponential$ function ($exp$ function), as shown in Algorithm~\ref{alg:Example-Algorithm-float}, which constitutes a major component of time spent in processing.
+Figure~\ref{fig:noexp} shows the percentage of time taken by Naive bilateral filtering algorithm and a modified algorithm, employing tiled-memory-access pattern, on different multicore platforms for a 512x512 pixels and another 3000x3000 pixels image, after the $exp$ function was commented out from the algorithms.
+The only purpose of this experiment is to justify the fact that traditional memory-access-pattern related optimizations can not improve the performance of bilateral filtering kernel.
 
@@
 
-Since the $exp$ instruction does not involve transfer of data from main memory, it follows from this experiment that memory reuse and memory access pattern related optimizations are not of much help to speed up Bilateral filtering kernel.
+Since the $exp$ instruction does not involve transfer of data from main memory, it follows from this experiment that memory reuse and memory access pattern related optimizations are not of much help to speed up bilateral filtering kernel.
 What Amdahl's law claims, for inherently sequential portion of a parallel algorithm, can be applied here.
 \\ \textit{The total speedup achieved by optimizations that deal with memory management is limited by the amount of time spent in memory transfers}.
-Therefore, in this case, traditional optimizations such as tiling/blocking, cache efficient initializations, prefetching, and other NUMA optimizations can only result in a speed up of at most 2\% and hence we have concentrated on optimizations that deal with the complexities of computational aspects of Bilateral filtering kernel.
+Therefore, in this case, traditional optimizations such as tiling/blocking, cache efficient initializations, prefetching, and other NUMA optimizations can only result in a speed up of at most 2\% and hence we have concentrated on optimizations that deal with the complexities of computational aspects of bilateral filtering kernel.
 
-\subsection{SIMD optimizations}
-\label{sub:SIMD1} 
-$Intel$'s Streaming SIMD Extension (SSE) is a technique to exploit micro-level parallelism in x86 based processors using SIMD capabilities.
-SSE can be ideal for tasks that involve repetitive computation such as the 3D heat equation or any other stencil kernels such as Jacobi iterations\cite{kdatta2008}.
-The matrix multiplication kernel was implemented in \cite{matrix-multi,MatrixMultiFloating} for matrices with integers and floating point numbers.
-SSE instructions have also been used for image or template matching algorithms such as normalized cross correlation (``NCC'') \cite{NCCmatching}.
-
-Modern day CPUs use SSE that is an extension of MMX technology and is fundamentally similar to MMX.
-It has 8, 128-bit registers named XMM0 through XMM7.
-SSE instructions can operate in parallel on data packed in these registers thus providing a theoretically maximum speedup of 16 when working with 1 Byte data structures or 8 when working with 2 Bytes long data structures and so on.
-There are other extensions of SSE such as SSE2, SSE3, and SSE4.2.
-Moreover, recent extension of SSE, named Advanced Vector Extensions (``AVX''), uses 256-bit data registers and thus a theoretical peak speedup of 32 is possible.
-
-Our CPU based implementation of Bilateral filtering algorithm explores the use of special purpose registers using both assembly level programming and SSE intrinsics.
-The advantage of using assembly language is that a programmer can have explicit control of underlying hardware.
-However, programmer productivity is affected and an in-depth knowledge of underlying hardware is essential.
-
-Although SSE intrinsics are wrappers for assembly language, it is preferred to use SSE intrinsics due to ease of coding and also due to efficient utilization of available assembly level instructions without knowing the low level details of underlying architecture.
-A programmer, working with intrinsics, can specify the intrinsic and the appropriate instruction will automatically replace this intrinsic during compilation.
-Moreover, there are some operations for which there are no assembly instructions and hence in case of assembly level programming the code size can grow to such an extent that readability gets compromised.
-
-Although the modern compilers usually try to improve the performance by automatically vectorizing, unrolling loops, prefetching, and SIMDization, we were still able to get significant performance improvements alongside compiler optimizations.
-
-\subsection{Reduction Methods}
-
-\label{sub:reduction} Sum by reduction is a technique to compute sum of multiple elements usually stored in contiguous memory locations.
-There are assembly instructions that can add horizontally packed, single precision, floating point numbers.
-We used $haddps$ assembly instruction that takes two (source and destination) 128 bit operands, and adds the single precision floating point numbers stored in first and second $dword$ of destination and source operand and stores it in first and third $dword$ of the destination operand respectively.
-Similarly, the third and fourth $dword$ of destination and source operand are added and stored in second and fourth $dword$ of the destination operand respectively.
-Both source and destination values must be stored in MMX registers.
-The SSE intrinsic to perform horizontal add is $\_mm\_hadd\_ps$.
-It takes two 128 bit operands, each of which is treated as four 32-bit floating point elements.
-The result of this operation on operand X(X0, X1, X2, X3) and Y(Y0, Y1, Y2, Y3) is (Y3 + Y2, Y1 + Y0, X3 + X2, X1 + X0).
-
-\section{Experimental Testbed}
-\label{sec:expt}
-
-\begin{table*}[t]
+\begin{table*}[t!]
 \begin{tabular}{|l|p{1.2cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{3.5cm}|l|p{2cm}|l|p{0.9cm}|}
@@
 \hline
-\end{tabular}\caption{Architectural details of multicore chips employed for experiments}
+\end{tabular}
+\caption{Architectural details of multicore chips employed for experiments}
 \label{tab:archs} 
@@
 
-In order to evaluate our implementation we have tried it on a range of multicore and manycore chips. Table \ref{tab:archs} summarizes architectural details of systems we used during our experiments.
+\subsection{SIMD optimizations}
+\label{sub:SIMD1} 
+$Intel$'s Streaming SIMD Extension (SSE) is a technique to exploit micro-level parallelism in x86 based processors using SIMD capabilities.
+SSE can be ideal for tasks that involve repetitive computation such as the 3D heat equation or any other stencil kernels such as Jacobi iterations\cite{kdatta2008}.
+The matrix multiplication kernel was implemented in \cite{matrix-multi,MatrixMultiFloating} for matrices with integers and floating point numbers.
+SSE instructions have also been used for image or template matching algorithms such as normalized cross correlation (``NCC'') \cite{NCCmatching}.
 
-\subsection{Hardware}
-\label{subsec:hw} 
-\subsubsection{Intel Xeon}
-\label{ss:xeon} We have used two different Xeon chips, E5410 (Harpertown) and X7560 (Nehalem EX).
-Harpertown is a dual-socket, dual-processor relatively old system with a clock frequency of 2.33 $GH_z$.
-Harpertown does not have Intel's hyperthreading (HT) technology which means the maximum number of threads is equal to the number of cores available.
-Intel Nehalem EX is considerably newer chip having eight-cores processing at a clock frequency of 2.27 $GH_z$.
-We had access to a Linux cluster from Intel's Manycore Testing Lab (MTL) containing four sockets of Nehalem EX chips thereby making 32 cores available to experiment with.
-The processors are connected to each other via Intel QPI (Quickpath interconnect) links that can transfer data at 6.4 GT/s.
-Both of these machines support 64 bit instructions.
+Modern day CPUs use SSE that is an extension of MMX technology and is fundamentally similar to MMX.
+It has 8, 128-bit registers named XMM0 through XMM7.
+SSE instructions can operate in parallel on data packed in these registers thus providing a theoretically maximum speedup of 16 when working with 1 Byte data structures or 8 when working with 2 Bytes long data structures and so on.
+There are other extensions of SSE such as SSE2, SSE3, and SSE4.2.
+Moreover, recent extension of SSE, named Advanced Vector Extensions (``AVX''), uses 256-bit data registers and thus a theoretical peak speedup of 32 is possible.
 
-\subsubsection{AMD Opteron}
-\label{ss:opteron} AMD Opteron 2376 (Shanghai) chip has private L1 and L2 caches of size 128KB (64 D + 64 I) and 512KB respectively per core along with a 6MB cache shared among all of the cores.
-It has a clock frequency of 2.30 $GH_z$ and it is a quad-core architecture.
-Our compute node has two of Shanghai chips making it a total of eight cores to be used for experiments.
-AMD Opteron 8350 (Barcelona) is similar to Shanghai in many aspects but it has a clock frequency of 2.0 $GH_z$ and the shared cache is only 2MB.
-Our compute node contains 4 of Barcelona chips and thus we are able to run 16 threads on this node.
+Our CPU based implementation of bilateral filtering algorithm explores the use of special purpose registers using both assembly level programming and SSE intrinsics.
+The advantage of using assembly language is that a programmer can have explicit control of underlying hardware.
+However, programmer productivity is affected and an in-depth knowledge of underlying hardware is essential.
 
-\subsubsection{Intel Core i7}
+Although SSE intrinsics are wrappers for assembly language, it is preferred to use SSE intrinsics due to ease of coding and also due to efficient utilization of available assembly level instructions without knowing the low level details of underlying architecture.
+A programmer, working with intrinsics, can specify the intrinsic and the appropriate instruction will automatically replace this intrinsic during compilation.
+Moreover, there are some operations for which there are no assembly instructions and hence in case of assembly level programming the code size can grow to such an extent that readability gets compromised.
 
-\label{ss:corei7}Intel Core i7 - 870 supports SSE 4.2 and has AVX technology.
-The clock frequency is an impressive 2.93 $GH_z$ and it can operate at a maximum of 3.6 $GH_z$ supported by $Intel$'s power boost technology.
-Its architecture supports hyperthreading which lets an application to execute 8 threads in total; however, we only used one thread per core.
-It contains 2 memory channels that provide a theoretical maximum memory bandwidth of 21 GB/s.
+Although the modern compilers usually try to improve the performance by automatically vectorizing, unrolling loops, prefetching, and SIMDization, we were still able to get significant performance improvements alongside compiler optimizations.
 
-\subsubsection{AMD Phenom II}
-\label{ss:phenom} AMD Phenom is a six core processor with each core capable of operating at 2.7 $GH_z$.
-If only three cores are in operation it can run at a maximum of 3.2 $GH_z$.
-It boasts 128 KB (64 D + 64 I) L1 cache, and 512 KB L2 cache per core in addition to a 6MB L3 cache shared among all six cores.
-The cache latencies are 3, 13, and 47 clock cycles for L1, L2, and L3 caches respectively.
-The cache line size is 64 bytes, L1 caches are 2 way set associative, L2 caches are 16-way set associative and L3 cache is 48-way set associative.
+\subsection{Reduction Methods}
+\label{sub:reduction} Sum by reduction is a technique to compute sum of multiple elements usually stored in contiguous memory locations.
+There are assembly instructions that can add horizontally packed, single precision, floating point numbers.
+We used $haddps$ assembly instruction that takes two (source and destination) 128 bit operands, and adds the single precision floating point numbers stored in first and second $dword$ of destination and source operand and stores it in first and third $dword$ of the destination operand respectively.
+Similarly, the third and fourth $dword$ of destination and source operand are added and stored in second and fourth $dword$ of the destination operand respectively.
+Both source and destination values must be stored in MMX registers.
+The SSE intrinsic to perform horizontal add is $\_mm\_hadd\_ps$.
+It takes two 128 bit operands, each of which is treated as four 32-bit floating point elements.
+The result of this operation on operand X(X0, X1, X2, X3) and Y(Y0, Y1, Y2, Y3) is (Y3 + Y2, Y1 + Y0, X3 + X2, X1 + X0).
 
-\subsubsection{NVidia GeForce GTX 280}
+\section{Experimental Testbed}
+\label{sec:expt}
 
-GTX 280 by NVidia is a GPU chip with 240 CUDA cores with 602 $MH_z$ graphics clock and 1296 $MH_z$ processor clock.
-GTX 280 is equipped with a 1 GB standard memory with 1107 $MH_z$ memory clock and 141.7 GB/s memory bandwidth.
-GTX 280 boasts a memory interface width of 512-bit with texture fill rate of 48.2 billion/second.
-
-\subsection{Software Framework}
-
-\label{subsec:sw} Our implementation for multicores is written in $C++$ and uses OpenMP for parallelization; the GPU implementation uses CUDA.
+In order to evaluate our implementation we have tried it on a range of multicore and manycore chips.
+Table \ref{tab:archs} summarizes architectural details of systems we used during our experiments.
+Our implementation for multicores is written in $C++$ and uses OpenMP for parallelization; the GPU implementation uses CUDA.
 To enrich supported image formats, we use $CImg$ library.
-\subsubsection{Algorithm variations on multicore architectures}
-We have implemented a number of variations of Bilateral filtering kernel for multicore chips, results of which are reported in section \ref{sec:results}.
-Our GPU implementation uses naive algorithm and the pair symmetric algorithm.
-For multicore chips we have created multiple variations of naive algorithm.
-These algorithms are labeled as follows:
-\begin{itemize}
-\item $BL\_Naive$ is the sequential version of the code, as described by Tomasi\cite{Tomasi1998}. 
-\item $BL\_Tiles$ is an improved version of Bilateral filtering kernel that employs data blocking to leverage the availability of multi-level cache hierarchy in the modern multicore CPUs.   
-\item $BL\_Assembly$ is the version of Bilateral filtering kernel implemented using assembly level code. 
-\item $BL\_A\_Reduction$ is the version of Bilateral filtering that uses assembly language as well as reduction optimizations. 
-\item $BL\_Intrinsic$ version uses SSE intrinsics instead of assembly language.
-\item $BL\_I\_Reduction$ is the extended version of $BL\_Intrinsic$ that uses the intrinsics that can compute sum of two 128 bit operands in parallel by exploiting the vector array processing. 
-\end{itemize}
-$BL\_Naive$ has a redundant calculation inside the innermost loop which has been moved out of loop for all other versions. The original instruction involving this calculation looks like
- $$gaussian\_ph = exp\left( \frac{-pd} {2 * sigma\_ph*sigma\_ph}\right);$$
-After moving the redundant calculation out of the loop the same instruction becomes
- $$gaussian\_ph = exp( pd * pre\_calc\_sigma\_sq );$$
-
-Our Bilateral filtering code is open-source licensed under GPL and available from $Google~Code$ repository \url{http://code.google.com/p/blfilter} through SVN version management system. 
-\subsubsection{Input Image}
-In order to analyze the behavior of our kernel with varying image sizes, we have executed our kernel with two different images.
-One that fits into cache easily and the other that does not fit into cache but fits into main memory.
-Both of the images were converted to single spectrum images represented by single precision numbers for each pixel.
-The smaller image was 512 pixels wide and 512 pixels high, thus requiring 1MB memory to store it.
-As shown in Table~\ref{tab:archs} all of the multicore systems have cache memories bigger than 1 MB.
-The second image was 3000 pixels wide and 3000 pixels high and hence requires little more than 34 MB of memory to store it.
-None of the multicore cache memories were big enough to store this image entirely.
-
-\section{Performance Results}
-\label{sec:results}
-\subsection{Performance of Bilateral filtering kernel on CPU}
-For all of our experiments the filter radius was kept at 10, subsequently making the filter Window of size 21 x 21.
-The spatial spread and the photometric spread were also set to 10.
-The optimization flag $O3$ was used for all experiments in addition to SSE2 for SIMDization.
-For Core i7 chip compiler flag SSE4.2 was used instead of SSE2.
-We do not use hyperthreading, i.e., there are no more than one thread per core.
 
@@
 \end{figure*} 
+%
+%\subsection{Hardware}
+%\label{subsec:hw} 
+%\subsubsection{Intel Xeon}
+%\label{ss:xeon} We have used two different Xeon chips, E5410 (Harpertown) and X7560 (Nehalem EX).
+Harpertown is a dual-socket, dual-processor relatively old system with a clock frequency of 2.33 $GH_z$.
+Harpertown does not have Intel's hyperthreading (HT) technology which means the maximum number of threads is equal to the number of cores available.
+Intel Nehalem EX is considerably newer chip having eight-cores processing at a clock frequency of 2.27 $GH_z$.
+We had access to a Linux cluster from Intel's Manycore Testing Lab (MTL) containing four sockets of Nehalem EX chips thereby making 32 cores available to experiment with.
+The processors are connected to each other via Intel QPI (Quickpath interconnect) links that can transfer data at 6.4 GT/s.
+Both of these machines support 64 bit instructions.
+%
+%\subsubsection{AMD Opteron}
+%\label{ss:opteron} AMD Opteron 2376 (Shanghai) chip has private L1 and L2 caches of size 128KB (64 D + 64 I) and 512KB respectively per core along with a 6MB cache shared among all of the cores.
+It has a clock frequency of 2.30 $GH_z$ and it is a quad-core architecture.
+Our compute node has two of Shanghai chips making it a total of eight cores to be used for experiments.
+AMD Opteron 8350 (Barcelona) is similar to Shanghai in many aspects but it has a clock frequency of 2.0 $GH_z$ and the shared cache is only 2MB.
+Our compute node contains 4 of Barcelona chips and thus we are able to run 16 threads on this node.
+%
+%\subsubsection{Intel Core i7}
+%
+%\label{ss:corei7}Intel Core i7 - 870 supports SSE 4.2 and has AVX technology.
+The clock frequency is an impressive 2.93 $GH_z$ and it can operate at a maximum of 3.6 $GH_z$ supported by $Intel$'s power boost technology.
+Its architecture supports hyperthreading which lets an application to execute 8 threads in total; however, we only used one thread per core.
+It contains 2 memory channels that provide a theoretical maximum memory bandwidth of 21 GB/s.
+%
+%\subsubsection{AMD Phenom II}
+%\label{ss:phenom} AMD Phenom is a six core processor with each core capable of operating at 2.7 $GH_z$.
+If only three cores are in operation it can run at a maximum of 3.2 $GH_z$.
+It boasts 128 KB (64 D + 64 I) L1 cache, and 512 KB L2 cache per core in addition to a 6MB L3 cache shared among all six cores.
+The cache latencies are 3, 13, and 47 clock cycles for L1, L2, and L3 caches respectively.
+The cache line size is 64 bytes, L1 caches are 2 way set associative, L2 caches are 16-way set associative and L3 cache is 48-way set associative.
+%
+%\subsubsection{NVidia GeForce GTX 280}
+%
+%GTX 280 by NVidia is a GPU chip with 240 CUDA cores with 602 $MH_z$ graphics clock and 1296 $MH_z$ processor clock.
+GTX 280 is equipped with a 1 GB standard memory with 1107 $MH_z$ memory clock and 141.7 GB/s memory bandwidth.
+GTX 280 boasts a memory interface width of 512-bit with texture fill rate of 48.2 billion/second.
 
-\begin{figure*} 
-\begin{centering} 
-\subfigure[Barcelona AMD Opteron 8350]{ \includegraphics[width=0.3\textwidth]{images/barcelona} \label{fig:barcelona} } 
-\subfigure[Shanghai AMD Opteron 2376]{ \includegraphics[width=0.3\textwidth]{images/shanghai} \label{fig:Shanghai} } 
-\subfigure[AMD Phenom II 1045T]{ \includegraphics[width=0.3\textwidth]{images/phenom} \label{fig:Phenom} } 
-\end{centering}
-\caption{Bilateral filtering kernel on AMD chips. The comparison is with BL\_Naive version, auto optimized by compiler with O3 and SSE flags.} 
-\label{fig:comparisonAMD} 
-\end{figure*} 
-\begin{figure*}[t!]
+We have implemented a number of variations of bilateral filtering kernel\footnote[2]{Our bilateral filtering code is open-source licensed under GPL and available from online repository \url{http://code.google.com/p/blfilter}.}  for multicore chips, results of which are reported in section \ref{sec:results}.
+These algorithms are labeled as follows:
+\begin{itemize}
+\item $BL\_Naive$ is the sequential version of the naive bilateral filtering algorithm, as described by Tomasi\cite{Tomasi1998}. 
+\item $BL\_Tiles$ is an improved version of bilateral filtering kernel that employs data blocking to leverage the availability of multi-level cache hierarchy in the modern multicore CPUs. 
+\item $BL\_Assembly$ is the version of naive bilateral filtering kernel implemented using assembly level code. 
+\item $BL\_A\_Reduction$ is the version of bilateral filtering kernel that uses assembly language as well as reduction optimizations discussed in Section~\ref{sub:reduction}.
+\item $BL\_Intrinsic$ version is an extension of $BL\_Assembly$ version where SSE intrinsics have replaced assembly level instructions.
+\item $BL\_I\_Reduction$ is the extended version of $BL\_Intrinsic$ that employs intrinsics for reduction methods.
+These intrinsics can compute the sum of two 128 bit operands in parallel by exploiting the vector array processing.
+\end{itemize}
+$BL\_Naive$ has a redundant calculation inside the innermost loop which has been moved out of loop for all other versions. The original instruction involving this calculation looks like
+ $$gaussian\_ph = exp\left( \frac{-pd} {2 * sigma\_ph*sigma\_ph}\right);$$
+After moving the redundant calculation out of the loop the same instruction becomes
+ $$gaussian\_ph = exp( pd * pre\_calc\_sigma\_sq );$$
 
-\begin{centering} 
-\subfigure[Harpertown Xeon E5410]{  \includegraphics[width=0.32\textwidth]{images/harpertownfinal} \label{fig:harpertownFinal} } 
-\subfigure[Core i7 - 870]{ \includegraphics[width=0.32\textwidth]{images/corei7final} \label{fig:corei7Final} } 
-\subfigure[Nehalem-EX Xeon X7650]{ \includegraphics[width=0.32\textwidth]{images/nehalemfinal} \label{fig:NehalemFinal} } 
-\end{centering}
-\caption{Comparison of combined speedup due to SIMDization and OpenMP parallelization on Intel chips for different algorithms} 
-\label{fig:comparisonFinalIntel} 
-\end{figure*}
-
-\begin{figure*} 
-\begin{centering} 
-\subfigure[Barcelona AMD Opteron 8350]{ \includegraphics[width=0.3\textwidth]{images/barcelonafinal} \label{fig:barcelonaFinal} } 
-\subfigure[Shanghai AMD Opteron 2376]{ \includegraphics[width=0.3\textwidth]{images/shanghaifinal} \label{fig:ShanghaiFinal} } 
-\subfigure[AMD Phenom II 1045T]{ \includegraphics[width=0.3\textwidth]{images/phenomfinal} \label{fig:PhenomFinal} } 
-\end{centering}
-\caption{Comparison of combined speedup due to SIMDization and OpenMP parallelization on AMD chips for different algorithms.} 
-\label{fig:comparisonFinalAMD} 
-\end{figure*}
+\subsubsection{Input Image}
+In order to analyze the behavior of our kernel with varying image sizes, we have executed our kernel with two different images.
+One that fits into cache easily and the other that does not fit into cache but fits into main memory.
+Both of the images were converted to single spectrum images represented by single precision numbers for each pixel.
+The smaller image was 512 pixels wide and 512 pixels high, thus requiring 1MB memory to store it.
+As shown in Table~\ref{tab:archs} all of the multicore systems have cache memories bigger than 1 MB.
+The second image was 3000 pixels wide and 3000 pixels high and hence requires little more than 34 MB of memory to store it.
+None of the multicore cache memories were big enough to store this image entirely.
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Performance Results}
+\label{sec:results}
+\subsection{Performance of bilateral filtering kernel on CPU}
+For all of our experiments the filter radius was kept at 10, subsequently making the filter Window of size 21 x 21.
+The spatial spread and the photometric spread were also set to 10.
+The optimization flag $-O3$ was used for all experiments in addition to SSE2 for SIMDization.
+For Core i7 chip compiler flag SSE4.2 was used instead of SSE2.
+We do not use hyperthreading, i.e.
+there are no more than one thread per core.
 
@@
 In this experiment we are comparing the absolute speedup achieved by the respective algorithm compared to BL\_Naive algorithm.
-It is important to note that these results are for micro level parallelism (use of 128 bit special purpose registers) only and do not consider speedup due to high level (OpenMP) parallelization.
+It is important to note that these results are for micro level parallelism (use of 128 bit special purpose registers) only; the results do not consider speedup due to high level (OpenMP) parallelization.
 The parallel versions shown here are compared against the parallel version of BL\_Naive algorithm.
-The speedup for all algorithms is still bounded by the theoretical peak speedup which happens to be four here since we work with 32bit floating point numbers.
-Interestingly, although we use compiler optimization flags such as -O3 and -xSSE2, i.e., there is automated vectorization taking place in all the algorithms, our hand coded vectorization is superior in many cases.
+The speedup for all algorithms is still bounded by the theoretical peak speedup which happens to be four here, since we work with 32bit floating point numbers.
+Interestingly, although we use compiler optimization flags such as -O3 and -xSSE2, i.e.
+there is automated vectorization taking place in all the algorithms, our hand coded vectorization is superior in many cases.
 
-The speedup shown by BL\_Tiles is mostly due to the computation reduction as discussed in section \ref{subsec:sw} because we do not use special purpose registers in this algorithm.
+The speedup shown by BL\_Tiles is mostly due to the computation reduction, as discussed in section \ref{sec:expt}, because we do not use special purpose registers in this algorithm.
 It can be seen from figure~\ref{fig:comparisonIntel} that the parallelism due to SIMDization is limited by four which is the theoretical maximum possible speedup due to SSE registers as discussed in section \ref{sub:SIMD1}.
@@
 
+\begin{figure*} [t!]
+\begin{centering} 
+\subfigure[Barcelona AMD Opteron 8350]{ \includegraphics[width=0.3\textwidth]{images/barcelona} \label{fig:barcelona} } 
+\subfigure[Shanghai AMD Opteron 2376]{ \includegraphics[width=0.3\textwidth]{images/shanghai} \label{fig:Shanghai} } 
+\subfigure[AMD Phenom II 1045T]{ \includegraphics[width=0.3\textwidth]{images/phenom} \label{fig:Phenom} } 
+\end{centering}
+\caption{Bilateral filtering kernel on AMD chips. The comparison is with BL\_Naive version, auto optimized by compiler with O3 and SSE flags.} 
+\label{fig:comparisonAMD} 
+\end{figure*}
+
+\begin{figure*}[t!] 
+\begin{centering} 
+\subfigure[Harpertown Xeon E5410]{  \includegraphics[width=0.3\textwidth]{images/harpertownfinal} \label{fig:harpertownFinal} } 
+\subfigure[Core i7 - 870]{ \includegraphics[width=0.3\textwidth]{images/corei7final} \label{fig:corei7Final} } 
+\subfigure[Nehalem-EX Xeon X7650]{ \includegraphics[width=0.3\textwidth]{images/nehalemfinal} \label{fig:NehalemFinal} } 
+\end{centering}
+\caption{Comparison of combined speedup due to SIMDization and OpenMP parallelization on Intel chips for different algorithms} 
+\label{fig:comparisonFinalIntel} 
+\end{figure*}
+
+\begin{figure*}[t!]
+\begin{centering} 
+\subfigure[Barcelona AMD Opteron 8350]{ \includegraphics[width=0.3\textwidth]{images/barcelonafinal} \label{fig:barcelonaFinal} } 
+\subfigure[Shanghai AMD Opteron 2376]{ \includegraphics[width=0.3\textwidth]{images/shanghaifinal} \label{fig:ShanghaiFinal} } 
+\subfigure[AMD Phenom II 1045T]{ \includegraphics[width=0.3\textwidth]{images/phenomfinal} \label{fig:PhenomFinal} } 
+\end{centering}
+\caption{Comparison of combined speedup due to SIMDization and OpenMP parallelization on AMD chips for different algorithms.} 
+\label{fig:comparisonFinalAMD} 
+\end{figure*}
+
 Figure~\ref{fig:comparisonAMD} shows the similar comparison across AMD chips.
@@
 
-Figure~\ref{fig:harpertownFinal} shows the performance of Bilateral filtering kernel on Intel Harpertown.
+Figure~\ref{fig:harpertownFinal} shows the performance of bilateral filtering kernel on Intel Harpertown.
 This machine has 8 cores and the maximum combined speedup achieved is 20.91x.
@@
 The speedup achieved using SSE optimizations was 2.48x and the speedup for BL\_Naive was 15.34x and thus these both realize a speedup of 38.03x using 16 cores.
+\begin{figure}
+\begin{centering} 
+\subfigure[Large Image, Filter Width = $10$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw10} \label{fig:hw10}} 
+\subfigure[Large Image, Filter Width = $5$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw5} \label{fig:hw5} } 
+\subfigure[Smaller Image, Filter Width = $10$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw10light} \label{fig:hw10light} } 
+\subfigure[Smaller Image, Filter Width = $5$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw5light} \label{fig:hw5light} } 
+\end{centering}
+\caption{Performance of Naive and pair-symmetric (P.S.) bilateral filtering kernel algorithms on GTX 280 NVidia card for both smaller image and larger image.} 
+\label{fig:gpuresults} 
+\end{figure}
 
@@
 The speedup drops a little as number of cores increases beyond three.
-The reason behind this is the turbo core technology as discussed in section (\ref{ss:phenom}).
-When there are a maximum of three cores being used the clock frequency scales up to 3.6 $GH_z$ but as soon as more threads are created the clock frequency steps down to 2.7 $GH_z$.
+The reason behind this is the turbo core technology; when there are a maximum of three cores being used, the clock frequency scales up to 3.6 $GH_z$ but as soon as more threads are created the clock frequency steps down to 2.7 $GH_z$.
 
-\subsection{Performance of Bilateral filtering kernel on GPU}
-
+\subsection{Performance of bilateral filtering kernel on GPU}
 We used the NVIDIA GTX 280 for benchmarking the performance of our CUDA implementations.
@@
 The results were remarkably illustrative of our pair-symmetric algorithm's potential.
-The naive algorithm running on the CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 1206.7 seconds while our pair-symmetric algorithm took 611.8 seconds - a 49.23\% reduction in execution time.
+The naive algorithm running on the CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 1206.7 seconds, while our pair-symmetric algorithm took 611.8 seconds - a 49.23\% reduction in execution time.
 Moreover, when replacing shared memory transactions with dummy register transactions in our CUDA implementation of the pair-symmetric algorithm, we observed a consistent 40\% reduction in execution time.
@@
 
-Another challenge in interpreting our results was porting the concept of "speed-up" to general purpose GPU programming.
-Since speed-up is necessarily with respect to a reference timing, we used the same reference timing that our CPU optimizations used---that is how we calculated the 235.5x speed-up figure stated in the abstract of this paper---and we also made another reference timing by running our GPU implementations using a single thread.
+Another challenge in interpreting our results was porting the concept of \lq{}\lq{}speed-up" to general purpose GPU programming.
+Since speed-up is necessarily with respect to a reference timing, we used the same reference timing that our CPU optimizations used - that is how we calculated the 235.5x speed-up figure stated in the abstract of this paper - and we also made another reference timing by running our GPU implementations using a single thread.
 For example, using a single thread operating with filter width of 10 pixels to process a 768 x 512 pixel image, the time taken for the GTX280 to complete this execution was an average of 177.9 seconds for both of our algorithms; using a filter width of 5 pixels, the average execution time was 50.5 seconds for both algorithms.
@@
 
-\begin{figure}[t!] 
-\begin{centering} 
-\subfigure[Large Image, Filter Width = $10$]{ \includegraphics[width=0.4\textwidth]{images/gpuhw10} \label{fig:hw10} } 
-\subfigure[Large Image, Filter Width = $5$]{ \includegraphics[width=0.4\textwidth]{images/gpuhw5} \label{fig:hw5} } 
-\subfigure[Smaller Image, Filter Width = $10$]{ \includegraphics[width=0.4\textwidth]{images/gpuhw10light} \label{fig:hw10light} } 
-\subfigure[Smaller Image, Filter Width = $5$]{ \includegraphics[width=0.4\textwidth]{images/gpuhw5light} \label{fig:hw5light} } 
-\end{centering}
-\caption{Performance of Naive and pair-symmetric (P.S.) Bilateral filtering kernel algorithms on GTX 280 NVidia card for both smaller image and larger image.} 
-\label{fig:gpuresults} 
-\end{figure}
-
 \section{Conclusion and future work}
 \label{sec:conclusion} 
-Modern day scientific applications demand higher computational capabilities than before and performance gains are no longer driven by clock speed.
-Chip manufacturers are trying to meet this demand by adding multiple cores on a chip equipped with multiple levels of cache hierarchy and enhanced SIMD extensions.
-In this paper we have tried to optimize Bilateral filtering kernel using both high level and low level parallelism.
-Our interest is to develop an efficient parallelization method for modern multicore architectures that can be applied to any spatially invariant and inseparable filter kernel and potentially to any compute-intensive regular numeric application.
-We have also documented why traditional optimization methods for multicores are not as effective for Bilateral filtering kernel compared to memory intensive kernels.
+Modern day scientific applications demand higher computational capabilities than before.
+The performance gains are no longer driven by clock speed.
+Chip manufacturers are trying to meet this demand by adding multiple cores on a chip equipped with multiple levels of cache hierarchy.
+In this paper we have tried to optimize bilateral filtering kernel using both high level and low level parallelism for multicore architectures as well as for manycore GPU chips.
+Our interest is to develop an efficient parallelization method for modern multicore and manycore architectures that can be applied to any spatially invariant and inseparable filter kernel and potentially to any compute-intensive regular numeric application.
+We have also documented why traditional optimization methods for multicores are not as effective for bilateral filtering kernel compared with memory intensive kernels.
 It can be seen from Figure~\ref{fig:comparisonIntel}, and Figure~\ref{fig:comparisonAMD} that a stencil kernel could be optimized by using SSE intrinsics even though the compiler optimizations are in place.
 Moreover, the reduction methods can provide a decent speedup where applicable.
-These findings will guide the auto optimizers and auto tuners to include SSE intrinsics as an essential optimization approach.
+These findings will guide the auto optimizers and auto tuners to include SSE intrinsics and reduction methods as an essential optimization approach.
 

### 2012-03-11 trunk/LateX/paper8.tex
Added and changed few plots, rearranged a few sections.

--- 2012-03-10 trunk/LateX/paper8.tex
@@
 Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties.
-This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter.
-It should be noted, however, that the foregoing equations are mathematical representations for filtering infinite-resolution images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image.
-Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set of calculations per Center Pixel.
+This results in edge-preserved image smoothing, which is the purpose of the bilateral filter.
+It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images.
+As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e.
+one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration.
 A formal expression is given below:
-
 \begin{equation}
@@
 
-In addition to restating the Bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the Bilateral filter.
+In addition to restating the bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the bilateral filter.
 Specifically, Equation~\ref{eq:depen} shows that before the filtered pixel can be calculated by dividing the numerator by the denominator, both the summation in the numerator and denominator must first be calculated.
-Consequently, the Bilateral filter has two data dependencies.
-As explained later in this paper, our pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.
+Consequently, the bilateral filter has two data dependencies.
+As explained later in this paper, the pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.
 
@@
 
+In other words, a memory transaction will be serialized unless all the threads in the warp to which the transaction belongs are aligned consecutively and fit within a segment of memory.
+Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory.
+For bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*jMax + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.
+
 While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable.
-As a consequence, work should be arranged such that warps can work as independently from each other as possible in order to minimize explicit synchronization calls, as illustrated in Figure~\ref{fig:pspro}.
-This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work of each warp on its current row of input data has completed.
-This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call would result in warps trespassing into the territory of other warps, thereby causing race conditions as threads from different warps would be consequently prone to writing to the same memory location.
+As a consequence, work must be arranged such that warps can work as independent from each other as possible, in order to minimize explicit synchronization calls.
+This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the current row of input data has completed.
+This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call will result in a warp trespassing into the territory of another warp, thereby causing race conditions as threads from two different warps will consequently be prone to write to the same memory location.
+
+\begin{figure}
+\subfigure[Each warp is assigned a row in filter window]{\includegraphics[width = 0.48 \textwidth]{images/project24.pdf}
+}\label{fig:pspro}
+\subfigure[2 x 2 thread-blocks each with 4 x 3 threads]{\includegraphics[width = 0.48  \textwidth]{images/tiles19.pdf}
+}\label{fig:pspro2}
+\subfigure[An illustrated overview of the bilateral filtering process]{\includegraphics[width = 0.48  \textwidth]{images/smallfaces.pdf}}
+\label{fig:pspro3}
+\caption{Memory access patterns for CUDA that exploit implicit intra-warp synchronization}
+\label{fig:intrawarp}
+\end{figure}
 
 \subsection{Shared Memory and Tile Processing}
-\begin{figure}
-\includegraphics[width = 0.48 \columnwidth]{images/project24.pdf}
-\caption{Illustration of our CUDA implementation of pair-symmetric processing using implicit intra-wap synchronization for iterating through the pixel-cells for each thread-block's assigned tile.}
-\label{fig:pspro}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width = 0.48  \columnwidth]{images/tiles19.pdf}
-\caption{Illustration of our CUDA implemented pair-symmetric algorithm processing a 24 x 12 pixel image using 2 x 2 thread-blocks each with 4 x 3 threads and with the half-width parameter set to 3 pixels.}
-\label{fig:pspro}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width = 0.48  \columnwidth]{images/smallfaces5.pdf}
-\caption{An illustrated overview of the bilateral filtering process.
-The six raster images of processed tiles reflect the successive work of the four concurrent thread-blocks iteratively carried out over the padded source-image.}
-\label{fig:pspro}
-\end{figure}
 
@@
 
-\begin{figure*} 
+%
+%\subsection{Hardware}
+%\label{subsec:hw} 
+%\subsubsection{Intel Xeon}
+%\label{ss:xeon} We have used two different Xeon chips, E5410 (Harpertown) and X7560 (Nehalem EX).
+Harpertown is a dual-socket, dual-processor relatively old system with a clock frequency of 2.33 $GH_z$.
+Harpertown does not have Intel's hyperthreading (HT) technology which means the maximum number of threads is equal to the number of cores available.
+Intel Nehalem EX is considerably newer chip having eight-cores processing at a clock frequency of 2.27 $GH_z$.
+We had access to a Linux cluster from Intel's Manycore Testing Lab (MTL) containing four sockets of Nehalem EX chips thereby making 32 cores available to experiment with.
+The processors are connected to each other via Intel QPI (Quickpath interconnect) links that can transfer data at 6.4 GT/s.
+Both of these machines support 64 bit instructions.
+%
+%\subsubsection{AMD Opteron}
+%\label{ss:opteron} AMD Opteron 2376 (Shanghai) chip has private L1 and L2 caches of size 128KB (64 D + 64 I) and 512KB respectively per core along with a 6MB cache shared among all of the cores.
+It has a clock frequency of 2.30 $GH_z$ and it is a quad-core architecture.
+Our compute node has two of Shanghai chips making it a total of eight cores to be used for experiments.
+AMD Opteron 8350 (Barcelona) is similar to Shanghai in many aspects but it has a clock frequency of 2.0 $GH_z$ and the shared cache is only 2MB.
+Our compute node contains 4 of Barcelona chips and thus we are able to run 16 threads on this node.
+%
+%\subsubsection{Intel Core i7}
+%
+%\label{ss:corei7}Intel Core i7 - 870 supports SSE 4.2 and has AVX technology.
+The clock frequency is an impressive 2.93 $GH_z$ and it can operate at a maximum of 3.6 $GH_z$ supported by $Intel$'s power boost technology.
+Its architecture supports hyperthreading which lets an application to execute 8 threads in total; however, we only used one thread per core.
+It contains 2 memory channels that provide a theoretical maximum memory bandwidth of 21 GB/s.
+%
+%\subsubsection{AMD Phenom II}
+%\label{ss:phenom} AMD Phenom is a six core processor with each core capable of operating at 2.7 $GH_z$.
+If only three cores are in operation it can run at a maximum of 3.2 $GH_z$.
+It boasts 128 KB (64 D + 64 I) L1 cache, and 512 KB L2 cache per core in addition to a 6MB L3 cache shared among all six cores.
+The cache latencies are 3, 13, and 47 clock cycles for L1, L2, and L3 caches respectively.
+The cache line size is 64 bytes, L1 caches are 2 way set associative, L2 caches are 16-way set associative and L3 cache is 48-way set associative.
+%
+%\subsubsection{NVidia GeForce GTX 280}
+%
+%GTX 280 by NVidia is a GPU chip with 240 CUDA cores with 602 $MH_z$ graphics clock and 1296 $MH_z$ processor clock.
+GTX 280 is equipped with a 1 GB standard memory with 1107 $MH_z$ memory clock and 141.7 GB/s memory bandwidth.
+GTX 280 boasts a memory interface width of 512-bit with texture fill rate of 48.2 billion/second.
+
+We have implemented a number of variations of bilateral filtering kernel\footnote[2]{Our bilateral filtering code is open-source licensed under GPL and available from online repository \url{http://code.google.com/p/blfilter}.}  for multicore chips, results of which are reported in section \ref{sec:results}.
+These algorithms are labeled as follows:
+\begin{itemize}
+\item $BL\_Naive$ is the sequential version of the naive bilateral filtering algorithm, as described by Tomasi\cite{Tomasi1998}. 
+\item $BL\_Tiles$ is an improved version of bilateral filtering kernel that employs data blocking to leverage the availability of multi-level cache hierarchy in the modern multicore CPUs. 
+\item $BL\_Assembly$ is the version of naive bilateral filtering kernel implemented using assembly level code. 
+\item $BL\_A\_Reduction$ is the version of bilateral filtering kernel that uses assembly language as well as reduction optimizations discussed in Section~\ref{sub:reduction}.
+\item $BL\_Intrinsic$ version is an extension of $BL\_Assembly$ version where SSE intrinsics have replaced assembly level instructions.
+\item $BL\_I\_Reduction$ is the extended version of $BL\_Intrinsic$ that employs intrinsics for reduction methods.
+These intrinsics can compute the sum of two 128 bit operands in parallel by exploiting the vector array processing.
+\end{itemize}
+$BL\_Naive$ has a redundant calculation inside the innermost loop which has been moved out of loop for all other versions. The original instruction involving this calculation looks like
+ $$gaussian\_ph = exp\left( \frac{-pd} {2 * sigma\_ph*sigma\_ph}\right);$$
+After moving the redundant calculation out of the loop the same instruction becomes
+ $$gaussian\_ph = exp( pd * pre\_calc\_sigma\_sq );$$
+
+\subsubsection{Input Image}
+In order to analyze the behavior of our kernel with varying image sizes, we have executed our kernel with two different images.
+One that fits into cache easily and the other that does not fit into cache but fits into main memory.
+Both of the images were converted to single spectrum images represented by single precision numbers for each pixel.
+The smaller image was 512 pixels wide and 512 pixels high, thus requiring 1MB memory to store it.
+As shown in Table~\ref{tab:archs} all of the multicore systems have cache memories bigger than 1 MB.
+The second image was 3000 pixels wide and 3000 pixels high and hence requires little more than 34 MB of memory to store it.
+None of the multicore cache memories were big enough to store this image entirely.
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Performance Results}
+\label{sec:results}
+\subsection{Performance of bilateral filtering kernel on CPU}
+For all of our experiments the filter radius was kept at 10, subsequently making the filter Window of size 21 x 21.
+The spatial spread and the photometric spread were also set to 10.
+The optimization flag $-O3$ was used for all experiments in addition to SSE2 for SIMDization.
+For Core i7 chip compiler flag SSE4.2 was used instead of SSE2.
+We do not use hyperthreading, i.e.
+there are no more than one thread per core.
+
+\begin{figure*} [t!]
 \begin{centering} 
@@
 \begin{figure*}[t!] 
-
 \begin{centering} 
-\subfigure[Harpertown Xeon E5410]{  \includegraphics[width=0.32\textwidth]{images/harpertownfinal} \label{fig:harpertownFinal} } 
-\subfigure[Core i7 - 870]{ \includegraphics[width=0.32\textwidth]{images/corei7final} \label{fig:corei7Final} } 
-\subfigure[Nehalem-EX Xeon X7650]{ \includegraphics[width=0.32\textwidth]{images/nehalemfinal} \label{fig:NehalemFinal} } 
+\subfigure[Harpertown Xeon E5410]{  \includegraphics[width=0.3\textwidth]{images/harpertownfinal} \label{fig:harpertownFinal} } 
+\subfigure[Core i7 - 870]{ \includegraphics[width=0.3\textwidth]{images/corei7final} \label{fig:corei7Final} } 
+\subfigure[Nehalem-EX Xeon X7650]{ \includegraphics[width=0.3\textwidth]{images/nehalemfinal} \label{fig:NehalemFinal} } 
 \end{centering}
-\caption{Comparison of combined speedup due to SIMDization and OpenMP parallelization on Intel chips for different algorithms} 
+\caption{Speedup on Intel Chips using OpenMP along with SSE} 
 \label{fig:comparisonFinalIntel} 
 \end{figure*} 
-
-\begin{figure*} 
+\begin{figure*}[t!]
 \begin{centering} 
@@
 \end{centering}
-\caption{Comparison of combined speedup due to SIMDization and OpenMP parallelization on AMD chips for different algorithms.} 
+\caption{Speedup on AMD chips  using OpenMP along with SSE} 
 \label{fig:comparisonFinalAMD} 
@@
 The next set of results demonstrate the total speedup due to OpenMP as well as due to SIMDization.
-It is interesting to see how speedup due to OpenMP multiplies with the speedup due to SSE optimizations to exhibit super linear speedup.
+It is interesting to see how speedup due to OpenMP, demonstrated by BL\_Naive algorithm,  multiplies with the speedup due to SSE optimizations to exhibit super linear speedup.
 
-Figure~\ref{fig:harpertownFinal} shows the performance of Bilateral filtering kernel on Intel Harpertown.
-This machine has 8 cores and the maximum combined speedup achieved is 20.91x.
+Figure~\ref{fig:harpertownFinal} shows the performance of bilateral filtering kernel on Intel Harpertown.
+This machine has 8 cores and the maximum combined speedup achieved is 23.07x.
 The speedup achieved only using only SSE optimization was 2.69x and the speedup achieved by BL\_Naive algorithm using 8 cores is 7.73x.
 
-Intel Nehalem chip with 32 cores does not show a super-linear speedup as shown in Figure~\ref{fig:NehalemFinal}.
-However, the reason behind this is the lack of enough parallelism that can be afforded by the kernel for this image.
+\begin{wrapfigure}{r}{0.25\textwidth}
+\centering
+\includegraphics[width=0.25\textwidth]{images/nehalemtime}
+\caption{Time taken by Intel Nehalem architecture}
+\vspace{0mm}
+\label{fig:nehalemtime}
+\end{wrapfigure}
+Intel Nehalem chip with 32 cores shows an impressive speedup as shown in Figure~\ref{fig:NehalemFinal}.
+The reason behind not showing a super-linear speedup in this case is the lack of enough parallelism that can be afforded by the kernel for this image.
+Figure~\ref{fig:nehalemtime} shows the absolute timings for the entire process, the timings tend to saturate with 32 processors as the total processing time approaches one second.
 Both optimizations still multiply their individual speedup to result in the combined speedup.
-The total combined speedup is 26.46x using 32 cores, the speedup only due to SSE was 1.12x and the speedup for BL\_Naive is 23.47x.
+The final speedup is 26.46x using 32 cores; the speedup only due to SSE was 1.12x and the speedup for BL\_Naive algorithm was 23.47x.
+\begin{figure*}[t!]
+\begin{centering} 
+\subfigure[Large 3000X3000 Image, Filter Width = $10$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw10} \label{fig:hw10}} 
+\subfigure[Large 3000X3000 Image, Filter Width = $5$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw5} \label{fig:hw5} } 
+\subfigure[Smaller 768X512 Image, Filter Width = $10$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw10light} \label{fig:hw10light} } 
+\subfigure[Smaller 768X512 Image, Filter Width = $5$]{ \includegraphics[width=0.48\textwidth]{images/gpuhw5light} \label{fig:hw5light} } 
+\end{centering}
+\caption{Performance of Naive and pair-symmetric (P.S.) bilateral filtering kernel algorithms on GTX 280 NVidia card for both smaller image and larger image.} 
+\label{fig:gpuresults} 
+\end{figure*}
 
 Intel Core i7 chip (Figure~\ref{fig:corei7Final}) also demonstrates the similar behavior.
-The final speedup of 4.63x, using 4 cores, is a result of speedup due to SSE which is 1.33x and speedup due to OpenMP which is 3.48x.
+The final speedup of 4.63x, using 4 cores, is a result of speedup due to SSE which is 1.33x and speedup due to OpenMP, using BL\_Naive algorithm, which is 3.48x.
 
@@
 The interesting case is with AMD Phenom as shown in Figure~\ref{fig:PhenomFinal}.
-The speedup drops a little as number of cores increases beyond three.
-The reason behind this is the turbo core technology as discussed in section (\ref{ss:phenom}).
-When there are a maximum of three cores being used the clock frequency scales up to 3.6 $GH_z$ but as soon as more threads are created the clock frequency steps down to 2.7 $GH_z$.
+The speedup drops a little as the number of cores increases beyond three.
+The reason behind this is the turbo core technology; when there are a maximum of three cores being used, the clock frequency scales up to 3.6 $GH_z$ but as soon as more threads are created the clock frequency steps down to 2.7 $GH_z$.
 
-\subsection{Performance of Bilateral filtering kernel on GPU}
+\subsection{Performance of bilateral filtering kernel on GPU}
 
@@
 The test that gave the closest difference in timing as shown in Figure~\ref{fig:hw10light} was processing a 768 x 712 pixel image using a 11 x 11 window size.
+
 In this test, the pair-symmetric algorithm's best timing was at a virtual tie with the naive algorithm's best timing at 0.00390 seconds, but upon further testing we found the pair-symmetric algorithm's optimal timing to be 0.00388 seconds when changing the grid dimension to 16 x 16 concurrent blocks, thus slightly outperforming the naive algorithm.
 Figures \ref{fig:hw10} and \ref{fig:hw10light} show exponential rises in execution time for both algorithms as block vertical dimension decreased.
-Due to runtime complexity of the bilateral filter being the square of the filter width, Figures (b) and (d), which reflect tests that used a 5 pixel filter width, do not exhibit the steep curvature that Figures (a) and (c) exhibit.
+Due to runtime complexity of the bilateral filter being the square of the filter width, Figures~\ref{fig:hw5} and \ref{fig:hw5}, which reflect our tests that used a 5 pixel filter width, do not exhibit the steep curvature that Figures~\ref{fig:hw10} and \ref{fig:hw10light} exhibit.
 
@@
 The main cause of both of these observations is shared memory.
+
 To better understand the impact of the additional shared memory transactions that our pair-symmetric algorithm requires, whereas the naive algorithm does not require shared memory for output storage, we implemented a single-threaded CPU-based version of both the naive bilateral filter algorithm and pair-symmetric algorithm so that native-memory (RAM, L1, or L2 cache) could substitute for the GPU's shared memory.
 The results were remarkably illustrative of our pair-symmetric algorithm's potential.
-The naive algorithm running on the CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 1206.7 seconds while our pair-symmetric algorithm took 611.8 seconds - a 49.23\% reduction in execution time.
+The naive algorithm running on the CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 1206.7 seconds, while our pair-symmetric algorithm took 611.8 seconds - a 49.23\% reduction in execution time.
+
 Moreover, when replacing shared memory transactions with dummy register transactions in our CUDA implementation of the pair-symmetric algorithm, we observed a consistent 40\% reduction in execution time.
 From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., achieving timings no less than 75\% of the naive algorithm's timings.
+
 To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code - CUDA's quasi-assembly language - of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions.

### 2012-03-11 src/symmetric.c
Added symmetric code to repo

@@ -1,0 +1,280 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <math.h>
+#include <time.h>
+#include <sys/time.h>
+
+
+#define HALF_WIDTH 10
+#define SIGMA 25.0
+#define SIGSP 50.0
+#define FULLWIDTH (2*HALF_WIDTH+1)
+
+
+struct uchar4{
+	unsigned char x, y, z, w;
+};
+
+
+struct inputVariables {
+	 int 
+		jMax, 	//hieght of the matrix
+		iMax,	//Width of the matrix
+		size,
+		halfWidth,
+		width;
+};
+struct hostVariables {
+	struct uchar4 pix;
+	unsigned char	* pixelArray;
+	float * gaussian_sp_WINDOW_ARRAY;
+};
+struct deviceVariables {
+	unsigned char	* pixelArray;
+	float
+			* Array1,
+			* Array2;
+	float * gaussian_sp_WINDOW_ARRAY;
+};
+//"struct timer" and functions "start," "stop," and "startNewTimer" are for easily setting up timers.
+
+#pragma pack(1)
+struct BMPHeader{
+    short type;
+    int size;
+    short reserved1;
+    short reserved2;
+    int offset;
+};
+#pragma pack(1)
+struct BMPInfoHeader{
+    unsigned int size;
+    int width;
+    int height;
+    unsigned short planes;
+    unsigned short int bitsPerPixel;
+    unsigned int compression;
+    unsigned int imageSize;
+    int xPelsPerMeter;
+    int yPelsPerMeter;
+    unsigned int clrUsed;
+    unsigned int clrImportant;
+};
+void * memset ( void * ptr, int value, size_t num );
+unsigned char *LoadBMPFile(BMPHeader *hdr, BMPInfoHeader *infoHdr, struct uchar4 **dst, int *width, int *height, const char *name){
+
+    int x, y;
+    FILE *fd;
+    printf("Loading %s...\n", name);
+    if( !(fd = fopen(name,"rb")) ){
+        printf("***BMP load error: file access denied***\n");
+        exit(0);
+    }
+
+    fread(hdr, sizeof(*hdr), 1, fd);
[…]

### 2012-03-11 trunk/LateX/paper8.tex
latest draft after changes from Dr. Prasad

@@
 \begin{document}
-\title{Acceleration of Bilateral filtering algorithm for manycore and multicore architectures}
+\title{Acceleration of bilateral filtering algorithm for manycore and multicore architectures}
 \author{\IEEEauthorblockN{Dinesh Agarwal, Sami Wilf, Abinashi Dhungel, and Sushil K. Prasad}\\
@@
 This work explores multicore and manycore acceleration for the embarrassingly parallel, compute-intensive bilateral filtering kernel.
-For manycore architectures, we have created a pair-symmetric algorithm to avoid redundant calculations and for multicore architectures we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads.
-We present empirical data evidencing the performance gains we achieved over variety of hardware architectures including Nvidia GTX280, AMD Barcelona, AMD Shanghai, Intel Harpertown, AMD Phenom, Intel Core i7 quad core, and Intel Nehalem 32 core machines.
-The best speedup achieved was a 235.5x speedup by our CUDA-based implementation of our pair-symmetric algorithm running on Nvidia's GTX280.
-Our CPU multicore implementations resulted in a speedup of up to 38x using 16 cores of AMD Barcelona each with 4-stage vector pipeline, up to 23x using 8 cores for Intel Harpertown, and up to 26x using 32 cores of Intel Nehalem multicore machines compared to a compiler-optimized code.
+For manycore architectures, we have created a pair-symmetric algorithm to avoid redundant calculations.
+For multicore architectures, we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads.
+We propose architecture specific optimizations, such as exploiting the unique capabilities of special registers available in modern multicore architectures and the rearrangement of data access patterns as per the computations to exploit special purpose instructions.
+We also propose optimizations pertinent to Nvidia's CUDA, including utilization of CUDA's implicit synchronization capability and the maximization of single-instruction-multiple-thread efficiency.
+We present empirical data on the performance gains we achieved over variety of hardware architectures including Nvidia GTX280, AMD Barcelona, AMD Shanghai, Intel Harpertown, AMD Phenom, Intel Core i7 quad core, and Intel Nehalem 32 core machines.
+The best speedup achieved was (i) 235.5x speedup by our CUDA-based implementation of our pair-symmetric algorithm running on Nvidia's GTX280 GPU and (ii) up to 38x using 16 cores of AMD Barcelona each with 4-stage vector pipeline compared to a compiler-optimized code.
 \end{abstract}
@@
 
-Our work explores efficient parallel implementation of bilateral filtering kernel\footnotemark[1] proposed by Tomasi and Manduchi \cite{Tomasi1998}, on multicore and manycore architectures.
-bilateral filtering kernel is used to smooth an image without blurring its edges.
-It is ubiquitous in various image processing applications in various contexts such as denoising\cite{zhang2008}, texture editing and relighting\cite{eisemann2004}, tone management\cite{Bae2006}, stylization\cite{DeCarlo2002}, optical-flow estimations\cite{Xiao2006}, and demosaicking\cite{Ramanath2003} etc.
+Our work explores efficient parallel implementation of bilateral filtering kernel\footnotemark[1] proposed by Tomasi and Manduchi \cite{Tomasi1998} on multicore and manycore architectures.
+Bilateral filtering kernel is used to smooth an image without blurring its edges.
+It is ubiquitous in various image processing applications in various contexts such as denoising\cite{zhang2008}, texture editing and relighting\cite{eisemann2004}, tone management\cite{Bae2006}, stylization\cite{DeCarlo2002}, optical-flow estimations\cite{Xiao2006}, and demosaicking\cite{Ramanath2003}, etc.
 
@@
 In this paper, we propose a set of optimizations and a non-naive algorithm for bilateral filtering kernel.
-Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to reduce processing time to half.
-We propose architecture specific optimizations, such as exploiting the unique capabilities of special registers available in modern multicore architectures and the rearrangement of data access patterns as per the computations to exploit special purpose instructions.
-We also propose optimizations pertinent to Nvidia's CUDA, including utilization of CUDA's implicit synchronization capability and the maximization of single-instruction-multiple-thread efficiency.
+Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to reduce processing time in half.
+Our specific contributions are
+\begin{enumerate}
+\item
+\item
+\item
+\end{enumerate}
 
@@
 This results in edge-preserved image smoothing, which is the purpose of the bilateral filter.
-It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images.
-As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e.
-one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration.
+It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image.
+Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set neighbors per Center Pixel.
 A formal expression is given below:
@@
 Moreover, as the size of the Window increases, which only happens by choice of the user as the size is user defined, the number of computations per pixel grows at a rate of $O(FilterWindowWidth^{2})$ where FilterWindowWidth is the diameter of the Window.
-By reason of the foregoing, the bulk of the time spent in processing an image using the bilateral filter is in the arithmetic-intensive computations, because the number of writes to the destination image is constant as the rate of growth of Center-Neighbor computations with respect to the Window size increase.
+Thus, the bulk of the time spent in processing an image using the bilateral filter is in the arithmetic-intensive computations, because the number of writes to the destination image is constant as the rate of growth of Center-Neighbor computations with respect to the Window size increase.
 
@@
 To take full advantage of CUDA's implicit intra-warp synchronization, divergent branching between threads in the same warp must be avoided.
-This is so because all the threads in the same warp share only one program counter.
 A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that met the branch condition to work.
-An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.
 
@@
 For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined.
-In other words, a memory transaction would be serialized if all the threads in the warp to which the transaction belonged were not aligned consecutively or did not fit within a segment of memory.
-Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory.
-For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*width + j] representation of a 2D image, for if threads in a warp were to read from or write to array elements (representing pixels) in a vertical order, there would be a separate transaction for each thread in the warp.
 
@@
 For the smaller image it is strictly less than 1\% of the original time with $exp$ instruction in place.
-To be precise it is 1.39\% maximum on Intel Nehalem for the larger image and 0.65\% for the smaller image on Core i7 processor.
+To be precise, it is 1.39\% maximum on Intel Nehalem for the larger image and 0.65\% for the smaller image on Core i7 processor.
 The lower memory transfer time for smaller image is due to the fact that the smaller image fits into the cache memory entirely and hence there are no repeated data transfers between the processor and the main memory.
@@
 
-We have implemented a number of variations of bilateral filtering kernel\footnote[2]{Our bilateral filtering code is open-source licensed under GPL and available from online repository \url{http://code.google.com/p/blfilter}.}  for multicore chips, results of which are reported in section \ref{sec:results}.
-These algorithms are labeled as follows:
+We have implemented a number of variations of bilateral filtering kernel for multicore chips, results of which are reported in section \ref{sec:results}. These algorithms are labeled as follows: 
 \begin{itemize}
@@
 For Core i7 chip compiler flag SSE4.2 was used instead of SSE2.
-We do not use hyperthreading, i.e.
-there are no more than one thread per core.
+We do not use hyperthreading, i.e., there are no more than one thread per core.
 
@@
 Moreover, when replacing shared memory transactions with dummy register transactions in our CUDA implementation of the pair-symmetric algorithm, we observed a consistent 40\% reduction in execution time.
-From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., achieving timings no less than 75\% of the naive algorithm's timings.
+From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., reducing timings by no more than than 25\%.
 
-To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code - CUDA's quasi-assembly language - of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions.
+To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code, CUDA's quasi-assembly language, of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions.
 Not only were no optimizations being made but the PTX code showed that the shared memory transactions required additional load and store instructions, further evidencing the cost of using shared memory.
 
-Another challenge in interpreting our results was porting the concept of \lq{}\lq{}speed-up" to general purpose GPU programming.
+\textbf{GPU Speedup:} Another challenge in interpreting our results was porting the concept of \lq{}\lq{}speed-up" to general purpose GPU programming.
 Since speed-up is necessarily with respect to a reference timing, we used the same reference timing that our CPU optimizations used - that is how we calculated the 235.5x speed-up figure stated in the abstract of this paper - and we also made another reference timing by running our GPU implementations using a single thread.

### 2012-03-12 trunk/LateX/paper8.tex
Added a new reference and changed a few minor things.

@@
 
+A previous study, closest to ours, is done by Xu et al.\cite{xu2009}, where they study the GPU-amenable implementation of bilateral filtering kernel.
+Our algorithms report 2x faster processing time than the timings reported in \cite{xu2009} on the same GPU - GTX 280.
+
 In this paper, we propose a set of optimizations and a non-naive algorithm for bilateral filtering kernel.
 Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to reduce processing time in half.
-Our specific contributions are
+Our specific contributions are:
 \begin{enumerate}
-\item
-\item
-\item
+\item Improving the performance of bilateral filtering kernel on multicore machines by exploiting the availability of special purpose registers.
+These improvements resulted in a speedup of up to 38x using 16 cores of AMD Barcelona each with 4-stage vector pipeline, up to 23x using 8 cores for Intel Harpertown, and up to 26x using 32 cores of Intel Nehalem multicore machines compared to a compiler-optimized code.
+\item Creating a novel pair-symmetric algorithm to reduce the number of calculations by a factor of two. 
+\item Efficiently porting the pair-symmetric algorithm on GPU by exploiting the intra-warp synchronization. 
 \end{enumerate}
@@
 \FOR{$l = -filter\_hw$ \TO $filter\_hw$}
+\STATE{/* $C_1$ and $C_2$ are user defined constants */}
 \STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i+k,j+l])^{2}}$}
@@
 \end{tabular}
-\caption{Architectural details of multicore chips employed for experiments}
+\caption{Architectural details of chips employed for experiments}
 \label{tab:archs} 
@@
 
-\subsubsection{Input Image}
-In order to analyze the behavior of our kernel with varying image sizes, we have executed our kernel with two different images.
+\textbf{Input Image:~}In order to analyze the behavior of our kernel with varying image sizes, we have executed our kernel with two different images.
 One that fits into cache easily and the other that does not fit into cache but fits into main memory.
@@
 The second image was 3000 pixels wide and 3000 pixels high and hence requires little more than 34 MB of memory to store it.
-None of the multicore cache memories were big enough to store this image entirely.
 %%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{centering}
-\caption{Performance of Naive and pair-symmetric (P.S.) bilateral filtering kernel algorithms on GTX 280 NVidia card for both smaller image and larger image.} 
+\caption{Performance of Naive and pair-symmetric (P.S.) bilateral filtering kernel algorithms on GTX 280 NVidia card.
+X and Y represent the number of threads inside a block; number of blocks is listed next to the name of the algorithm (64x64 and 32x32).}
 \label{fig:gpuresults} 
@@
 The reason behind this is the turbo core technology; when there are a maximum of three cores being used, the clock frequency scales up to 3.6 $GH_z$ but as soon as more threads are created the clock frequency steps down to 2.7 $GH_z$.
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Performance of bilateral filtering kernel on GPU}
@@
 
-To better understand the impact of the additional shared memory transactions that our pair-symmetric algorithm requires, whereas the naive algorithm does not require shared memory for output storage, we implemented a single-threaded CPU-based version of both the naive bilateral filter algorithm and pair-symmetric algorithm so that native-memory (RAM, L1, or L2 cache) could substitute for the GPU's shared memory.
+To better understand the impact of the additional shared memory transactions that our pair-symmetric algorithm requires, whereas the naive algorithm does not require shared memory for output storage, we implemented a CPU-based version of both the naive bilateral filter algorithm and pair-symmetric algorithm so that native-memory (RAM, L1, or L2 cache) could substitute for the GPU's shared memory.
 The results were remarkably illustrative of our pair-symmetric algorithm's potential.
-The naive algorithm running on the CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 1206.7 seconds, while our pair-symmetric algorithm took 611.8 seconds - a 49.23\% reduction in execution time.
+The naive algorithm running on Barcelona CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 131.98 seconds, while our pair-symmetric algorithm took 74.31 seconds - a 43.7\% reduction in execution time.
 
